\documentclass[11pt,a4paper,twoside,openright]{book}

% ============================================================
% ENCODING & FONTS
% ============================================================
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% ============================================================
% MATHEMATICS
% ============================================================
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{thmtools}
\usepackage{stmaryrd}        % \llbracket, \rrbracket
\usepackage{bm}              % bold math

% ============================================================
% GRAPHICS & DIAGRAMS
% ============================================================
\usepackage{tikz}
\usetikzlibrary{
  arrows.meta, positioning, shapes.geometric, fit, calc,
  backgrounds, decorations.pathmorphing, decorations.markings,
  trees, matrix, chains, patterns, shadows
}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}

% ============================================================
% TABLES
% ============================================================
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{tabularx}
\usepackage{multirow}
\usepackage{array}

% ============================================================
% LAYOUT
% ============================================================
\usepackage[margin=1in]{geometry}
\usepackage{enumitem}
\usepackage{multicol}
\usepackage{caption}
\usepackage{float}
\usepackage{graphicx}
\usepackage{titlesec}
\usepackage{fancyhdr}
\usepackage{emptypage}       % blank pages truly blank

% ============================================================
% HYPERLINKS & CROSS-REFERENCES
% ============================================================
\usepackage[
  colorlinks=true,
  linkcolor=defblue,
  citecolor=thmred,
  urlcolor=edgeorange,
  bookmarksdepth=3
]{hyperref}
\usepackage{cleveref}

% ============================================================
% COLORS
% ============================================================
\usepackage{xcolor}
\definecolor{defblue}{RGB}{31,78,121}
\definecolor{thmred}{RGB}{153,0,0}
\definecolor{sepgreen}{RGB}{0,100,0}
\definecolor{layergray}{RGB}{245,245,245}
\definecolor{edgeorange}{RGB}{204,102,0}
\definecolor{notegray}{RGB}{100,100,100}
\definecolor{boundbrown}{RGB}{139,90,43}
\definecolor{analogypurple}{RGB}{102,0,153}
\definecolor{goldcolor}{RGB}{184,134,11}
\definecolor{onlineblue}{RGB}{0,51,153}

% ============================================================
% COLORED BOXES
% ============================================================
\usepackage{tcolorbox}
\tcbuselibrary{theorems,skins,breakable}

\newtcolorbox{separation}{
  colback=sepgreen!5, colframe=sepgreen!50!black,
  title=Separation Result, fonttitle=\bfseries\small,
  breakable
}
\newtcolorbox{traversal}{
  colback=defblue!5, colframe=defblue!50!black,
  title=Graph Traversal, fonttitle=\bfseries\small,
  breakable
}
\newtcolorbox{obstbox}{
  colback=edgeorange!5, colframe=edgeorange!50!black,
  title=Obstruction, fonttitle=\bfseries\small,
  breakable
}
\newtcolorbox{historical}{
  colback=goldcolor!5, colframe=goldcolor!50!black,
  title=Historical Note, fonttitle=\bfseries\small,
  breakable
}
\newtcolorbox{computational}{
  colback=layergray, colframe=notegray!50!black,
  title=Computational Illustration, fonttitle=\bfseries\small,
  breakable
}
\newtcolorbox{graphnode}[1][]{
  colback=defblue!3, colframe=defblue!30!black,
  fonttitle=\bfseries\small\sffamily,
  title={Graph Node}, #1,
  breakable
}

% ============================================================
% THEOREM ENVIRONMENTS
% ============================================================
\theoremstyle{definition}
\newtheorem{defn}{Definition}[chapter]
\newtheorem{example}[defn]{Example}
\newtheorem{notation}[defn]{Notation}
\newtheorem{convention}[defn]{Convention}

\theoremstyle{plain}
\newtheorem{thm}[defn]{Theorem}
\newtheorem{prop}[defn]{Proposition}
\newtheorem{cor}[defn]{Corollary}
\newtheorem{lemma}[defn]{Lemma}
\newtheorem{claim}[defn]{Claim}

\theoremstyle{remark}
\newtheorem{remark}[defn]{Remark}
\newtheorem{openprob}[defn]{Open Problem}
\newtheorem{warning}[defn]{Warning}

% ============================================================
% PAGE STYLE
% ============================================================
\pagestyle{fancy}
\fancyhf{}
\fancyhead[LE]{\small\textit{A Textbook of Formal Learning Theory}}
\fancyhead[RO]{\small\textit{\leftmark}}
\fancyfoot[C]{\thepage}
\renewcommand{\headrulewidth}{0.4pt}

% ============================================================
% CHAPTER TITLE FORMAT
% ============================================================
\titleformat{\chapter}[display]
  {\normalfont\huge\bfseries\color{defblue}}
  {\chaptertitlename\ \thechapter}{20pt}{\Huge}
\titlespacing*{\chapter}{0pt}{50pt}{40pt}

% ============================================================
% MACROS
% ============================================================
% Dimensions
\newcommand{\VC}{\ensuremath{\mathrm{VCdim}}}
\newcommand{\Ldim}{\ensuremath{\mathrm{Ldim}}}
\newcommand{\DSdim}{\ensuremath{\mathrm{DSdim}}}
\newcommand{\Ndim}{\ensuremath{d_{\mathrm{N}}}}
\newcommand{\Pdim}{\ensuremath{\mathrm{Pdim}}}
\newcommand{\fatshat}{\ensuremath{\mathrm{fat}}}
\newcommand{\SQdim}{\ensuremath{\mathrm{SQdim}}}

% Complexities
\newcommand{\Rad}{\ensuremath{\widehat{\mathcal{R}}}}
\newcommand{\KL}{\ensuremath{\mathrm{KL}}}
\newcommand{\ERM}{\ensuremath{\mathrm{ERM}}}
\newcommand{\SOA}{\ensuremath{\mathrm{SOA}}}

% Learning criteria
\newcommand{\Ex}{\ensuremath{\mathbf{Ex}}}
\newcommand{\BC}{\ensuremath{\mathbf{BC}}}
\newcommand{\FIN}{\ensuremath{\mathbf{FIN}}}

% Probability and expectation
\newcommand{\E}{\ensuremath{\mathbb{E}}}
\renewcommand{\Pr}{\ensuremath{\mathbb{P}}}
\newcommand{\ind}{\ensuremath{\mathbf{1}}}
\newcommand{\R}{\ensuremath{\mathbb{R}}}
\newcommand{\N}{\ensuremath{\mathbb{N}}}
\newcommand{\Z}{\ensuremath{\mathbb{Z}}}

% Risk
\newcommand{\risk}{\ensuremath{R}}
\newcommand{\emprisk}{\ensuremath{\hat{R}}}

% Graph relations
\newcommand{\rel}[1]{\texttt{\small #1}}
\newcommand{\nde}[1]{\textsf{\small #1}}
\newcommand{\edge}[3]{\nde{#1}\ensuremath{\;\xrightarrow{\text{\rel{#2}}}\;}\nde{#3}}

% Growth function
\newcommand{\growth}{\ensuremath{\Pi}}

% ============================================================
% TITLE
% ============================================================
\title{%
  \textbf{\Huge A Textbook of Formal Learning Theory}\\[1em]
  \Large Six Paradigms, Their Characterization Theorems,\\
  and the Separations Between Them\\[1em]
  \large Across Six Decades of Computational and Statistical Learning Theory
}
\author{%
  Dhruv Gupta\\
  Zetesis Labs, ARTPARK @ IISc Bangalore\\
  \texttt{dhruv@zetesis.org}
}
\date{March 2026}

% ============================================================
\begin{document}
% ============================================================

\frontmatter
\maketitle

% ============================================================
% EPIGRAPH
% ============================================================
\thispagestyle{empty}
\vspace*{\fill}
\begin{center}
\begin{minipage}{0.7\textwidth}
\centering\itshape\large
``The question is not which functions can be learned,\\
but which functions can be learned \emph{given what you already know}.''
\end{minipage}
\end{center}
\vspace*{\fill}
\newpage

% ============================================================
% PREFACE
% ============================================================
\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

This book presents formal learning theory as a single, navigable mathematical structure.  It treats six paradigms---PAC, online, Gold-style, exact, universal, and multiclass learning---within a unified framework of typed relations, and devotes equal attention to the separations between them.

The need for such a book arises from a structural gap. The field's foundational results are distributed across several research traditions that developed largely in parallel:

\begin{itemize}[leftmargin=*]
\item \textbf{PAC learning} (Valiant, 1984) and its combinatorial characterization through VC dimension (Vapnik--Chervonenkis, 1971; Blumer et al., 1989).
\item \textbf{Online learning} (Littlestone, 1988) and its characterization through the Littlestone dimension.
\item \textbf{Gold-style identification in the limit} (Gold, 1967) and its mind-change hierarchy (Freivalds--Smith, Case--Smith).
\item \textbf{Universal learning} (Bousquet et al., 2021) and the trichotomy theorem.
\item \textbf{Multiclass learning} and its resolution through the DS dimension (Brukhim et al., 2022).
\item \textbf{Generalization bounds} from five distinct traditions: VC/Rademacher, PAC-Bayes, algorithmic stability, information-theoretic, and margin-based.
\end{itemize}

No single textbook covers all these traditions with their precise interconnections and, critically, their precise \emph{non}-connections. Shalev-Shwartz and Ben-David~\cite{ShalevShwartzBenDavid2014} treat PAC and online learning comprehensively but omit Gold-style identification, universal learning, and information-theoretic bounds. Jain et al.~\cite{JainOshersonRoyerSharma1999} cover inductive inference thoroughly but not PAC theory. The generalization-bound literature is scattered.

This book fills the gap by treating all traditions within a single framework organized around typed relations between concepts.  The key structural innovation is not any new theorem, but a \emph{negative layer}: the book devotes equal attention to what \emph{does not} hold---separation results with explicit witnesses, cross-paradigm analogies with their precise obstructions, and the boundaries where one framework's characterization theorem fails to generalize to another.

\subsection*{How to Read This Book}

The book is organized in five parts:

\begin{description}[leftmargin=*]
\item[Part I: Foundations] (Chapters 1--3) introduces the base types---domains, concepts, hypothesis spaces, data presentations, and the automata-theoretic vocabulary underlying identification in the limit.
\item[Part II: Paradigms] (Chapters 4--9) presents the six major learning paradigms, each with its characterization theorem proved in full: PAC learning (VC dimension), online learning (Littlestone dimension), Gold-style identification (mind-change ordinals), exact learning (query models), and universal learning (the trichotomy).
\item[Part III: Complexity Measures] (Chapters 10--13) develops the 33 complexity measures in the graph, organized by what they measure: combinatorial dimensions, sample complexity and compression, generalization bounds, and mind-change ordinals.
\item[Part IV: The Negative Layer] (Chapters 14--15) presents all separation results with full proofs and all analogy-obstruction pairs with analysis.
\item[Part V: Extensions] (Chapters 16--18) covers computational hardness, extensions beyond binary classification, and application-layer concepts.
\end{description}

Each chapter includes TikZ diagrams showing the subgraph relevant to that chapter's concepts and computational illustrations where appropriate.  Cross-references to the companion knowledge graph are given in the appendices.

\subsection*{Prerequisites}

The reader should be comfortable with:
\begin{itemize}[leftmargin=*]
\item Basic probability (expectation, concentration inequalities, Markov/Chebyshev/Hoeffding bounds).
\item Basic combinatorics (growth functions, pigeonhole principle).
\item Computability theory at the level of decidability, Turing machines, and the halting problem (for Gold-style chapters).
\item Mathematical maturity sufficient for $\varepsilon$-$\delta$ arguments.
\end{itemize}

Background in machine learning is helpful but not required; the book is self-contained from the definitions upward.

\subsection*{Companion Materials}

The book is accompanied by a machine-readable knowledge graph and a benchmark suite, available at \url{https://github.com/Zetetic-Dhruv/Formal-Learning-Theory}.  The graph encodes 142 concepts connected by 260 typed edges across 13 relation types; it is described in detail in Appendix~A (edge inventory) and Appendix~C (validation).  The repository also contains a complete bibliography and scripts for validating the graph against the textbook's content.

\subsection*{Notation}

A notation index appears in Appendix~D. We highlight the most frequently used symbols here:

\begin{center}\small
\begin{tabular}{ll@{\qquad}ll}
\toprule
$\mathcal{C}$ & concept class & $\VC(\mathcal{H})$ & VC dimension \\
$\mathcal{H}$ & hypothesis space & $\Ldim(\mathcal{H})$ & Littlestone dimension \\
$D$ & distribution on $X \times Y$ & $\DSdim(\mathcal{H})$ & DS dimension \\
$S = \{(x_i, y_i)\}_{i=1}^m$ & training sample & $\Ndim(\mathcal{H})$ & Natarajan dimension \\
$\risk(h) = \E_{(x,y) \sim D}[\ell(h(x),y)]$ & true risk & $\Pdim(\mathcal{F})$ & pseudodimension \\
$\emprisk(h) = \frac{1}{m}\sum \ell(h(x_i),y_i)$ & empirical risk & $\fatshat_\gamma(\mathcal{F})$ & fat-shattering dim \\
$\varepsilon$ & accuracy parameter & $\Rad_n(\mathcal{H})$ & Rademacher complexity \\
$\delta$ & confidence parameter & $\growth_\mathcal{H}(n)$ & growth function \\
\bottomrule
\end{tabular}
\end{center}

\vspace{1em}
\hfill \textit{Dhruv Gupta, Bangalore, March 2026}

\tableofcontents

\listoffigures
\addcontentsline{toc}{chapter}{List of Figures}

\listoftables
\addcontentsline{toc}{chapter}{List of Tables}

% ============================================================
\mainmatter
% ============================================================

\input{chapters/ch01_objects}
\input{chapters/ch02_data}
\input{chapters/ch03_automata}
\input{chapters/ch04_agents}
\input{chapters/ch05_pac}
\input{chapters/ch06_online}
\input{chapters/ch07_gold}
\input{chapters/ch08_exact}
\input{chapters/ch09_universal}
\input{chapters/ch10_dimensions}
\input{chapters/ch11_compression}
\input{chapters/ch12_generalization}
\input{chapters/ch13_ordinals}
\input{chapters/ch14_separations}
\input{chapters/ch15_analogies}
\input{chapters/ch16_computational}
\input{chapters/ch17_extensions}
\input{chapters/ch18_frontiers}

% ============================================================
\appendix
% ============================================================

\input{appendices/appA_edge_inventory}
\input{appendices/appB_traversals}
\input{appendices/appC_validation}
\input{appendices/appD_notation}

% ============================================================
\backmatter
% ============================================================

\bibliographystyle{alpha}
\bibliography{../flt_bibliography}

\end{document}
