% Chapter 3: Automata, Languages, and Computability
%   dfa, nfa, pushdown_automaton, turing_machine, cfg = Profile A (vocabulary)
%   regular_language, cfl, re_set, partial_computable = Profile A (one-line)
%   godel_numbering = Profile A-plus (brief but load-bearing for Gold)
%   restricts/extends_grammar edge pair = Profile C (Revelation: conceptual centerpiece)
%   limiting_recursion + delta_02_class = Profile G (Framework Bridge to Gold)

\chapter{Automata, Languages, and Computability}
\label{ch:automata}

This chapter assembles the automata-theoretic and computability-theoretic vocabulary required by Gold's model of identification in the limit (\Cref{ch:gold}). Most of the material is standard and is presented without proof; the reader who needs full development should consult Hopcroft and Ullman~\cite{HopcroftUllman1979} or Sipser~\cite{Sipser2013}.

Two things in this chapter are \emph{not} standard vocabulary. First, the distinction between \emph{restricts} and \emph{extends\_grammar} edges---two ways a computational model can generalize another, one conservative and one genuinely new---is illustrated here for the first time through the NFA/DFA and PDA/DFA pairs (\S\ref{sec:restricts-extends}). This distinction pervades the entire knowledge graph. Second, the notion of limiting recursion connects computability theory to learning theory by placing Gold's identification in the limit at a precise level of the arithmetical hierarchy (\S\ref{sec:bridge}).


% ================================================================
% SECTION 1: The Chomsky Hierarchy (~1 page)
% ================================================================

\section{The Chomsky Hierarchy}
\label{sec:chomsky}

The Chomsky hierarchy classifies formal languages by the computational resources needed to recognize them. Figure~\ref{fig:chomsky} displays the containment structure; the definitions follow.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
    every node/.style={font=\small},
    level/.style={draw, thick, ellipse, minimum height=1.2cm, align=center}
]
% Nested ovals, outermost first
\node[level, minimum width=10cm, minimum height=5.8cm, fill=layergray!60] (re) at (0,0) {};
\node[font=\scriptsize, anchor=north] at (re.north) {Recursively enumerable (Type 0)};

\node[level, minimum width=7.8cm, minimum height=4.2cm, fill=defblue!6] (rec) at (0,-0.3) {};
\node[font=\scriptsize, anchor=north] at (rec.north) {Recursive (decidable)};

\node[level, minimum width=5.6cm, minimum height=2.8cm, fill=edgeorange!8] (cfl) at (0,-0.5) {};
\node[font=\scriptsize, anchor=north] at (cfl.north) {Context-free (Type 2)};

\node[level, minimum width=3.2cm, minimum height=1.4cm, fill=sepgreen!10] (reg) at (0,-0.6) {};
\node[font=\scriptsize] at (reg.center) {Regular (Type 3)};

% Recognizer labels on the right
\node[font=\scriptsize\itshape, notegray, anchor=west] at (4.2,2.2) {Turing machine};
\node[font=\scriptsize\itshape, notegray, anchor=west] at (3.2,0.8) {TM (halting)};
\node[font=\scriptsize\itshape, notegray, anchor=west] at (2.2,-0.1) {Pushdown automaton};
\node[font=\scriptsize\itshape, notegray, anchor=west] at (1.2,-0.6) {DFA / NFA};
\end{tikzpicture}
\caption{The Chomsky hierarchy. Each level is strictly contained in the next. The recognizer type is indicated at right. All containments are proper.}
\label{fig:chomsky}
\end{figure}

\subsection{Recognizers}

\begin{defn}[Deterministic Finite Automaton]
\label{def:dfa}
A \emph{DFA} is a 5-tuple $M = (Q, \Sigma, \delta, q_0, F)$ where $Q$ is a finite state set, $\Sigma$ is a finite input alphabet, $\delta : Q \times \Sigma \to Q$ is a total transition function, $q_0 \in Q$ is the start state, and $F \subseteq Q$ is the set of accepting states. $M$ accepts a string $w \in \Sigma^*$ if the unique run on $w$ ends in a state in $F$.
\end{defn}

\begin{defn}[Nondeterministic Finite Automaton]
\label{def:nfa}
An \emph{NFA} is a 5-tuple $M = (Q, \Sigma, \delta, q_0, F)$ identical to a DFA except that $\delta : Q \times (\Sigma \cup \{\varepsilon\}) \to \mathcal{P}(Q)$ maps each state--symbol pair to a \emph{set} of successor states, and $\varepsilon$-transitions are permitted. $M$ accepts $w$ if \emph{some} run on $w$ ends in $F$.
\end{defn}

\begin{defn}[Pushdown Automaton]
\label{def:pda}
A \emph{pushdown automaton} (PDA) is a 7-tuple $M = (Q, \Sigma, \Gamma, \delta, q_0, Z_0, F)$ where $\Gamma$ is a finite stack alphabet, $Z_0 \in \Gamma$ is the initial stack symbol, and $\delta : Q \times (\Sigma \cup \{\varepsilon\}) \times \Gamma \to \mathcal{P}(Q \times \Gamma^*)$ maps each state--input--stack-top triple to a set of (state, stack-replacement) pairs. The PDA has access to an unbounded LIFO stack; it accepts by final state or empty stack.
\end{defn}

\begin{defn}[Turing Machine]
\label{def:tm}
A \emph{Turing machine} is a 7-tuple $M = (Q, \Sigma, \Gamma, \delta, q_0, q_{\mathrm{acc}}, q_{\mathrm{rej}})$ where $\Gamma \supseteq \Sigma$ is the tape alphabet (including a blank symbol), and $\delta : Q \times \Gamma \to Q \times \Gamma \times \{L, R\}$ is a partial transition function specifying the next state, symbol to write, and head movement. $M$ accepts $w$ if it halts in $q_{\mathrm{acc}}$.
\end{defn}

\begin{defn}[Context-Free Grammar]
\label{def:cfg}
A \emph{context-free grammar} (CFG) is a 4-tuple $G = (V, \Sigma, R, S)$ where $V$ is a finite set of variables, $\Sigma$ is the terminal alphabet, $R \subseteq V \times (V \cup \Sigma)^*$ is a finite set of production rules, and $S \in V$ is the start variable. $G$ generates the language $L(G) = \{w \in \Sigma^* : S \Rightarrow^* w\}$.
\end{defn}

\subsection{Language classes}

A language $L \subseteq \Sigma^*$ is \emph{regular} if it is accepted by some DFA. It is \emph{context-free} if it is generated by some CFG (equivalently, accepted by some PDA). It is \emph{recursively enumerable} (r.e.) if it is accepted by some Turing machine. It is \emph{recursive} (decidable) if it is accepted by a Turing machine that halts on every input.

A function $f : \N \rightharpoonup \N$ is \emph{partial computable} if there exists a Turing machine $M$ such that $M$ on input $n$ halts with output $f(n)$ when $f(n)$ is defined and diverges otherwise.

\subsection{Closure properties}

The language classes differ sharply in their closure properties. Table~\ref{tab:closure} summarizes the facts needed in later chapters.

\begin{table}[ht]
\centering
\small
\caption{Closure properties of the Chomsky hierarchy classes. $\checkmark$ = closed, $\times$ = not closed. All proofs are standard; see~\cite{HopcroftUllman1979}.}
\label{tab:closure}
\begin{tabular}{lccccc}
\toprule
\textbf{Class} & \textbf{Union} & \textbf{Intersection} & \textbf{Complement} & \textbf{Concat.} & \textbf{Kleene $*$} \\
\midrule
Regular       & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ \\
Context-free  & $\checkmark$ & $\times$     & $\times$     & $\checkmark$ & $\checkmark$ \\
Recursive     & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ \\
R.E.          & $\checkmark$ & $\checkmark$ & $\times$     & $\checkmark$ & $\checkmark$ \\
\bottomrule
\end{tabular}
\end{table}


% ================================================================
% SECTION 2: restricts vs extends_grammar (Profile C — Revelation)
% Conceptual centerpiece of the chapter
% ================================================================

\section{Two Kinds of Generalization: Restricts vs.\ Extends}
\label{sec:restricts-extends}

The Chomsky hierarchy is a containment chain. But the \emph{edges} in this chain are not all of the same type. Moving from one level to another sometimes removes a constraint (making the model less restrictive without adding new primitives) and sometimes adds a genuinely new resource (granting the model access to a mechanism it did not previously have). This distinction---between \emph{conservative} and \emph{generative} generalization---is one of the recurring structural motifs of the knowledge graph, and the automata hierarchy provides its first concrete instance.

\begin{defn}[Restricts]
\label{def:restricts}
Model $B$ \rel{restricts} model $A$ if $B$ is obtained from $A$ by \emph{removing a constraint}: every instance of $A$ is already an instance of $B$ (after trivial embedding), and the two models accept exactly the same class of languages. The generalization is conservative---it adds flexibility in representation without expanding expressive power.
\end{defn}

\begin{defn}[Extends Grammar]
\label{def:extends-grammar}
Model $B$ \rel{extends\_grammar} model $A$ if $B$ is obtained from $A$ by \emph{adding a new computational primitive}: $B$ recognizes a strictly larger class of languages than $A$. The generalization is generative---it requires a fundamentally new resource.
\end{defn}

The DFA is the anchor point for both edges. Figure~\ref{fig:restricts-extends} displays the structure.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
    box/.style={draw, thick, rounded corners, minimum width=2.4cm, minimum height=0.9cm, font=\small\bfseries, align=center},
    arr/.style={-{Stealth[length=2.5mm]}, thick},
    lbl/.style={font=\scriptsize, fill=white, inner sep=2pt, align=center}
]
\node[box, fill=sepgreen!12] (dfa) at (0,0) {DFA};
\node[box, fill=sepgreen!12] (nfa) at (5,1.5) {NFA};
\node[box, fill=edgeorange!12] (pda) at (5,-1.5) {PDA};

\draw[arr, sepgreen!70!black] (nfa) -- node[lbl, above left] {\rel{restricts}\\{\scriptsize\itshape removes determinism}\\{\scriptsize\itshape same power}} (dfa);
\draw[arr, edgeorange!80!black] (pda) -- node[lbl, below left] {\rel{extends\_grammar}\\{\scriptsize\itshape adds stack}\\{\scriptsize\itshape strictly more power}} (dfa);

% Annotations
\node[font=\scriptsize, text=sepgreen!70!black, anchor=west] at (6, 1.5) {Regular languages};
\node[font=\scriptsize, text=edgeorange!80!black, anchor=west] at (6, -1.5) {Context-free languages};
\end{tikzpicture}
\caption{Two generalizations of the DFA. The NFA removes the determinism constraint (conservative: same expressive power). The PDA adds a stack (generative: strictly more expressive power). Both arrows point toward the DFA to indicate that the DFA is the more constrained model.}
\label{fig:restricts-extends}
\end{figure}

\paragraph{The NFA--DFA edge: conservative generalization.}
An NFA differs from a DFA by allowing multiple successor states per transition and $\varepsilon$-moves. This \emph{removes} the determinism constraint. Yet the subset construction shows that every NFA with $n$ states has an equivalent DFA with at most $2^n$ states~\cite{HopcroftUllman1979}. The class of languages recognized is unchanged: $\mathrm{REG}_{\mathrm{NFA}} = \mathrm{REG}_{\mathrm{DFA}}$. The NFA is a more flexible \emph{representation} of the same expressive power.

This is a \rel{restricts} edge. The NFA does not need any new grammar or computational primitive---it simply relaxes a structural requirement. The cost is representational blowup (exponential in the worst case), not a change in what can be expressed.

\paragraph{The PDA--DFA edge: generative generalization.}
A pushdown automaton extends a finite automaton by adding an unbounded stack. This is not the removal of a constraint but the \emph{introduction of a new primitive}: the ability to store and retrieve unbounded auxiliary information in LIFO order. The consequence is a strict increase in expressive power. The language $\{a^n b^n : n \geq 0\}$ is context-free (accepted by a PDA that pushes $a$'s and pops on $b$'s) but not regular (by the pumping lemma). The containment is proper: $\mathrm{REG} \subsetneq \mathrm{CFL}$.

This is an \rel{extends\_grammar} edge. The PDA requires a fundamentally new mechanism---a stack---and this mechanism purchases genuinely new expressive power.

\paragraph{Why this distinction matters.}
The \rel{restricts}/\rel{extends\_grammar} pair recurs throughout the graph. When a new model \rel{restricts} an old one, the theory of the old model transfers intact---the same characterization theorems, the same closure properties, the same complexity measures apply. When a new model \rel{extends\_grammar}, the old theory typically \emph{breaks}: new characterization theorems are needed, closure properties may change (\Cref{tab:closure}: context-free languages lose closure under intersection and complement), and new complexity measures may be required.

This distinction will reappear in the learning-theoretic setting: when a learning criterion generalizes another by relaxing a convergence requirement (e.g., behaviorally correct learning relaxes syntactic convergence), it is a conservative generalization. When it adds a new resource (e.g., adding a teacher who provides counterexamples), it is generative. Recognizing which type of generalization is at work is essential for predicting whether existing results transfer to the new setting.


% ================================================================
% SECTION 3: Computability Foundations (~0.5 page)
% ================================================================

\section{Computability Foundations for Learning}
\label{sec:computability}

Gold's model of identification in the limit requires learners to be effective procedures that output hypotheses from an effective enumeration. This section fixes the relevant computability-theoretic vocabulary.

\begin{defn}[G\"odel Numbering]
\label{def:godel-numbering}
A \emph{G\"odel numbering} is an effective bijection between natural numbers and programs (or, equivalently, Turing machines). We write $\varphi_e$ for the partial computable function computed by the program with index $e$, and $W_e = \mathrm{dom}(\varphi_e)$ for the $e$-th r.e.\ set. The sequence $\varphi_0, \varphi_1, \varphi_2, \ldots$ is a \emph{universal effective enumeration} of all partial computable functions, and $W_0, W_1, W_2, \ldots$ is an effective enumeration of all r.e.\ sets.
\end{defn}

In Gold's framework, a learner receiving a text for a language $L$ outputs, at each time step, a natural number $e_t$. This number is interpreted as a hypothesis via the G\"odel numbering: the learner conjectures that $L = W_{e_t}$. Identification in the limit means that the sequence $e_0, e_1, e_2, \ldots$ eventually stabilizes on an index $e^*$ with $W_{e^*} = L$.

The critical point is that learners are themselves programs: a learner is a computable function $M : \Sigma^* \to \N$ mapping finite data sequences to hypothesis indices. The set of admissible learners is thus the set of total computable functions from finite sequences to $\N$. This effectivity requirement is what distinguishes Gold's model from a purely set-theoretic notion of convergence.


% ================================================================
% SECTION 4: The Learning-Theoretic Bridge (~1 page)
% ================================================================

\section{The Learning-Theoretic Bridge: Limiting Recursion and $\Delta^0_2$}
\label{sec:bridge}

The deepest connection between computability theory and learning theory is not a definition but an equivalence. It identifies Gold's criterion of identification in the limit with a precise level of the arithmetical hierarchy.

\begin{defn}[Limiting Recursion]
\label{def:limiting-recursion}
A function $f : \N \to \N$ is \emph{limiting recursive} (or \emph{$\Delta^0_2$-computable}) if there exists a total computable function $\hat{f} : \N \times \N \to \N$ such that for every $x$,
\[
f(x) = \lim_{s \to \infty} \hat{f}(x, s).
\]
That is, $f$ is the pointwise limit of a computable approximation sequence. For each input $x$, the approximation $\hat{f}(x, 0), \hat{f}(x, 1), \hat{f}(x, 2), \ldots$ eventually stabilizes on $f(x)$, but the time of stabilization is not itself computable.
\end{defn}

\begin{defn}[$\Delta^0_2$ Class]
\label{def:delta-02}
A set $A \subseteq \N$ is \emph{$\Delta^0_2$} if both $A$ and its complement $\bar{A}$ are $\Sigma^0_2$---that is, both are expressible as $\{x : \exists y\, \forall z\, R(x,y,z)\}$ for some computable predicate $R$. Equivalently, $A$ is $\Delta^0_2$ if and only if its characteristic function is limiting recursive.
\end{defn}

The connection to learning is the following classical result.

\begin{thm}[Limiting Recursion and Learnability {\cite{Gold1965,Putnam1965,ShoenfieldsLimit}}]
\label{thm:limit-bridge}
A family of r.e.\ sets $\mathcal{L} = \{L_i\}_{i \in I}$ is identifiable in the limit (in the sense of Gold) if and only if the index function $i \mapsto e_i$, mapping each language to a correct hypothesis index, is limiting recursive. The classes identifiable in the limit are exactly those whose index sets are $\Delta^0_2$.
\end{thm}

This theorem is the bridge between two fields. On the computability side, the $\Delta^0_2$ sets occupy a precise position in the arithmetical hierarchy: they are the sets computable with an oracle for the halting problem, equivalently, the sets whose membership can be decided by a procedure that is allowed to change its mind finitely many times. On the learning side, identification in the limit is exactly the requirement that the learner's conjecture sequence stabilizes---which is exactly the requirement that the hypothesis function be a limit of computable approximations.

The equivalence tells us three things that will be used in \Cref{ch:gold}:

\begin{enumerate}[leftmargin=*,itemsep=3pt]
\item \textbf{Upper bound on learnability.} No class whose index set lies above $\Delta^0_2$ in the arithmetical hierarchy can be identified in the limit. This bounds what Gold-style learning can achieve from above.

\item \textbf{Mind changes as approximation stages.} The number of times a limiting-recursive function changes its approximation on input $x$---the number of $s$ for which $\hat{f}(x,s) \neq \hat{f}(x,s+1)$---corresponds to the number of mind changes a learner makes before converging. The mind-change ordinals of \Cref{ch:ordinals} refine this correspondence into a full hierarchy.

\item \textbf{Identification in the limit is not ad hoc.} Gold's criterion might appear to be one arbitrary convergence requirement among many. The $\Delta^0_2$ characterization shows it is not arbitrary at all: it is the \emph{canonical} notion of computability-in-the-limit, the natural first level above the computable in the arithmetical hierarchy.
\end{enumerate}


\section{What This Chapter Established}
\label{sec:ch3-summary}

This chapter served two functions. As vocabulary, it fixed the definitions---DFA, NFA, PDA, Turing machine, CFG, G\"odel numbering, partial computable function---needed for the identification-in-the-limit chapters. As conceptual contribution, it introduced two structural ideas:

\begin{enumerate}[leftmargin=*,itemsep=3pt]
\item \textbf{The \rel{restricts}/\rel{extends\_grammar} distinction.} The NFA--DFA pair (conservative generalization, same power) and the PDA--DFA pair (generative generalization, strictly more power) provide the first concrete illustration of an edge-type distinction that recurs throughout the graph. Whether a generalization is conservative or generative determines whether existing theory transfers or must be rebuilt.

\item \textbf{The $\Delta^0_2$ bridge.} Identification in the limit is not merely a learning-theoretic desideratum but a precise computability-theoretic concept: it is computation in the limit, equivalent to $\Delta^0_2$-computability. This equivalence anchors Gold's model in the arithmetical hierarchy and provides the foundation for the mind-change hierarchy of \Cref{ch:ordinals}.
\end{enumerate}

