% Chapter 9: Universal Learning and the Trichotomy
% universal_learning = Profile C: what is the optimal learning RATE over all distributions?
% universal_trichotomy = Profile C: the modern capstone — three regimes,
%   classified by Littlestone trees (from online learning!) applied to the statistical setting.
% Centerpiece: the SURPRISE that the Littlestone dimension controls statistical rates.

\chapter{Universal Learning and the Trichotomy}
\label{ch:universal}

The Fundamental Theorem of \Cref{ch:pac} answers a binary question: \emph{is}
$\mathcal{H}$ learnable?  The answer is yes if and only if
$\VC(\mathcal{H}) < \infty$.  But the binary answer hides a quantitative
question.  Among all learnable classes, some are learned faster than others.
How fast?

PAC learning theory phrases this quantitatively through the sample complexity
$m(\varepsilon, \delta)$: how many samples suffice for accuracy $\varepsilon$
with confidence $1 - \delta$?  But the answer---$\Theta(d/\varepsilon)$ in
the realizable case---is a function of the \emph{accuracy parameter}.  It
describes how accuracy improves as data grows, with $d$ as a constant.  It
does not describe the \emph{rate} at which the risk itself decreases as a
function of the sample size $n$.

This chapter asks the rate question directly: given $n$ samples from an
unknown distribution $D$, how fast can the risk $\risk_D(A(S))$ be driven to
zero, \emph{uniformly over all distributions $D$ for which learning is
possible}?  The answer is the trichotomy theorem, which classifies every
hypothesis class into exactly one of three rate regimes.  The classification
involves a combinatorial object from an unexpected source.


% ================================================================
% SECTION 1: THE QUESTION
% ================================================================

\section{Beyond PAC Sample Complexity}
\label{sec:universal-question}

Fix a hypothesis class $\mathcal{H}$ with $\VC(\mathcal{H}) < \infty$, so
that $\mathcal{H}$ is PAC learnable.  The PAC framework guarantees a learner
$A$ and a sample complexity $m(\varepsilon, \delta)$ such that, for any
distribution $D$ and any $\varepsilon, \delta > 0$, drawing
$m \geq m(\varepsilon, \delta)$ samples ensures
$\Pr[\risk_D(A(S)) > \varepsilon] \leq \delta$.

Now invert the perspective.  Instead of asking ``how many samples for
accuracy $\varepsilon$?'', ask: ``given $n$ samples, what is the best
accuracy achievable?''  More precisely, define the \emph{learning rate} of a
class $\mathcal{H}$ as follows.

\begin{defn}[Universal Learning Rate]
\label{def:universal-rate}
A function $R \colon \N \to [0,1]$ is a \emph{universal learning rate} for
$\mathcal{H}$ if there exists a learner $A$ such that, for every realizable
distribution $D$ (i.e., $\risk_D(h^*) = 0$ for some $h^* \in \mathcal{H}$)
and every $n \in \N$:
\[
  \E_{S \sim D^n}\!\bigl[\risk_D(A(S))\bigr] \leq R(n).
\]
The \emph{optimal universal rate} is the fastest-decaying $R$ achievable by
any learner.
\end{defn}

Three features distinguish this from the PAC formulation:

\begin{enumerate}[leftmargin=*,itemsep=3pt]
\item \textbf{No fixed $\varepsilon$.}  The rate $R(n)$ describes how the
  expected risk shrinks with $n$, rather than asking for a threshold sample
  size at a given $\varepsilon$.

\item \textbf{Uniformity over distributions.}  The same learner $A$ and the
  same rate $R(n)$ must work for \emph{every} realizable distribution $D$.
  The learner cannot be tuned to a specific distribution.

\item \textbf{Expected risk, not high-probability.}  We use
  $\E[\risk_D(A(S))]$ rather than a $(\varepsilon,\delta)$ guarantee.  This
  is a convenience that simplifies the rate classification without changing
  the qualitative picture.
\end{enumerate}

For a class with $\VC(\mathcal{H}) = d < \infty$, the PAC upper bound gives
$R(n) = O(d/n)$: a $\Theta(1/n)$ rate.  But is $1/n$ always achievable?
Can some classes be learned faster---exponentially fast?  And are there
learnable classes where no uniform $1/n$ rate is possible, so that the rate
degrades depending on the distribution?

All three phenomena occur.  The trichotomy theorem says these are the
\emph{only} three possibilities.


% ================================================================
% SECTION 2: THE TRICHOTOMY THEOREM
% ================================================================

\section{The Trichotomy Theorem}
\label{sec:trichotomy}

The classification involves a combinatorial object that generalizes the
Littlestone tree of \Cref{ch:online}.  Recall that a mistake tree
(\Cref{def:mistake-tree}) is a complete binary tree whose internal nodes are
labeled with instances $x \in X$, such that every root-to-leaf path is
realized by some $h \in \mathcal{H}$.  We now define a variant with a
weaker consistency requirement.

\begin{defn}[VCL Tree]
\label{def:vcl-tree}
A \emph{VCL tree} (Vapnik--Chervonenkis--Littlestone tree) for
$\mathcal{H}$ over $X$ is a complete binary tree $T$ in which:
\begin{itemize}[leftmargin=*,itemsep=2pt]
\item Each internal node $v$ is labeled with an instance $x_v \in X$.
\item The left child corresponds to label $0$, the right to label $1$.
\item For every root-to-leaf path $\pi = (v_1, y_1), \ldots, (v_d, y_d)$,
  there exists a \emph{set} $\{x_{v_1}, \ldots, x_{v_d}\}$ that is
  \emph{shattered} by $\mathcal{H}$---not merely a single hypothesis
  consistent with $\pi$, but the full shattering condition: for every
  labeling $b \in \{0,1\}^d$ of these $d$ points, some $h \in \mathcal{H}$
  realizes $b$.
\end{itemize}
The tree is \emph{infinite} if it has unbounded depth.
\end{defn}

\begin{remark}[VCL trees vs.\ Littlestone trees]
\label{rmk:vcl-vs-littlestone}
Every Littlestone tree is a VCL tree (shattering is a stronger condition
than path-consistency, but the shattering requirement in the VCL definition
applies to each path individually, and a Littlestone tree already provides a
consistent hypothesis for every path---since the tree is complete, all
$2^d$ paths are realized, which implies shattering of the instances along
any single path).  The converse fails: a VCL tree requires shattering along
each path, which is a weaker global condition than the Littlestone tree's
requirement that \emph{every} root-to-leaf path has a \emph{single}
consistent hypothesis.  The distinction matters: infinite VCL trees can
exist even when $\Ldim(\mathcal{H}) < \infty$.
\end{remark}

\begin{thm}[The Trichotomy Theorem {\cite{BousquetHannekeMoranVanHandelYehudayoff2021}}]
\label{thm:trichotomy}
Let $\mathcal{H}$ be a hypothesis class with $|\mathcal{H}| \geq 3$.
Exactly one of the following three cases holds:
\begin{enumerate}[label=\textnormal{(T\arabic*)}]
\item \label{T:exp} \textbf{Exponential rate.}  $\mathcal{H}$ has no
  infinite Littlestone tree (equivalently, $\Ldim(\mathcal{H}) < \infty$).
  Then the optimal universal learning rate is \emph{exponential}:
  \[
    R(n) = \Theta\!\left(e^{-n}\right).
  \]

\item \label{T:linear} \textbf{Linear rate.}  $\mathcal{H}$ has an infinite
  Littlestone tree but no infinite VCL tree.  Then the optimal universal
  learning rate is \emph{linear}:
  \[
    R(n) = \Theta\!\left(\frac{1}{n}\right).
  \]

\item \label{T:arb} \textbf{Arbitrarily slow rates.}  $\mathcal{H}$ has an
  infinite VCL tree.  Then no uniform rate exists: for every function
  $R(n) \to 0$ (no matter how slowly), every learner $A$ fails to achieve
  rate $R(n)$ on some realizable distribution $D$.
\end{enumerate}
\end{thm}

The three cases are exhaustive and mutually exclusive.  Every learnable class
(i.e., every class with $\VC(\mathcal{H}) < \infty$ and
$|\mathcal{H}| \geq 3$) falls into exactly one regime.  The boundaries are
sharp: there is no class whose optimal rate is, say, $1/\sqrt{n}$ or
$1/n^2$.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
    decision/.style={draw, thick, diamond, aspect=2.2, inner sep=2pt,
      font=\small, fill=defblue!8, align=center, minimum width=3cm},
    outcome/.style={draw, thick, rounded corners, minimum width=2.8cm,
      minimum height=1.2cm, font=\small, align=center},
    arr/.style={-{Stealth[length=2.5mm]}, thick},
    yeslabel/.style={font=\scriptsize\bfseries, text=sepgreen},
    nolabel/.style={font=\scriptsize\bfseries, text=thmred},
]
% Decision 1
\node[decision] (d1) {Infinite\\Littlestone tree?};

% No → Exponential
\node[outcome, fill=sepgreen!12, left=2.8cm of d1, yshift=-2.5cm] (exp) {%
  \textbf{Exponential rate}\\[3pt]
  $R(n) = \Theta(e^{-n})$\\[2pt]
  {\scriptsize $\Ldim(\mathcal{H}) < \infty$}};

% Yes → Decision 2
\node[decision, right=2.8cm of d1, yshift=-2.5cm] (d2) {Infinite\\VCL tree?};

% No → Linear
\node[outcome, fill=defblue!12, left=2.5cm of d2, yshift=-2.5cm] (lin) {%
  \textbf{Linear rate}\\[3pt]
  $R(n) = \Theta(1/n)$\\[2pt]
  {\scriptsize $\Ldim = \infty$, no $\infty$-VCL}};

% Yes → Arbitrarily slow
\node[outcome, fill=thmred!12, right=2.5cm of d2, yshift=-2.5cm] (slow) {%
  \textbf{Arbitrarily slow}\\[3pt]
  $\forall\, R(n) \to 0$: fails\\[2pt]
  {\scriptsize $\infty$-VCL tree exists}};

% Arrows
\draw[arr] (d1) -- node[nolabel, left, pos=0.3] {No} (exp);
\draw[arr] (d1) -- node[yeslabel, right, pos=0.3] {Yes} (d2);
\draw[arr] (d2) -- node[nolabel, left, pos=0.3] {No} (lin);
\draw[arr] (d2) -- node[yeslabel, right, pos=0.3] {Yes} (slow);
\end{tikzpicture}
\caption{The trichotomy decision tree.  Two binary questions---does
$\mathcal{H}$ have an infinite Littlestone tree?\ does it have an infinite
VCL tree?---partition all hypothesis classes (with $|\mathcal{H}| \geq 3$)
into exactly three rate regimes.  There is no fourth regime and no
intermediate rate.}
\label{fig:trichotomy-tree}
\end{figure}

\subsection{An Online Concept in a Statistical Theorem}
\label{sec:surprise}

The trichotomy theorem is the structural capstone of this book.  To
appreciate what it says, consider the expectations each tradition brought
to the table.

The PAC tradition (\Cref{ch:pac}) characterized learnability through the VC
dimension and established the sample complexity
$\Theta(d/\varepsilon)$.  The natural expectation: the VC dimension should
also determine the learning rate.  It does not.
Two classes with the same VC dimension can have different optimal
rates---one exponential, one $1/n$.  So what draws the boundary?

The online learning tradition (\Cref{ch:online}) introduced the Littlestone
dimension to characterize mistake-bounded learning against an adversary.
A worst-case combinatorial quantity, built for a worst-case adversarial
game.  No connection to statistical convergence rates.  None expected.

And yet.  The Littlestone dimension---this adversarial, combinatorial
object---is exactly what governs the boundary between exponential and
polynomial convergence in the i.i.d.\ setting.

Neither tradition predicted this.  The trichotomy does not merely classify
rates; it reveals that the deepest structural boundary in statistical
learning theory is drawn by a quantity from a different paradigm entirely.

\subsection{Examples}
\label{sec:trichotomy-examples}

\begin{example}[The three regimes, witnessed]
\label{ex:three-regimes}
\begin{enumerate}[label=(\alph*),leftmargin=*,itemsep=4pt]
\item \textbf{Exponential: finite classes.}  Any finite class
  $\mathcal{H}$ with $|\mathcal{H}| = k$ has
  $\Ldim(\mathcal{H}) \leq \lfloor\log_2 k\rfloor < \infty$.  No infinite
  Littlestone tree exists, so $\mathcal{H}$ falls into regime~\ref{T:exp}.
  Concretely, the Halving algorithm achieves risk $\leq k \cdot 2^{-n}$
  after $n$ rounds.

\item \textbf{Exponential: halfspaces.}  The class of halfspaces in
  $\R^d$ has $\Ldim = d < \infty$ (\Cref{ex:ldim-examples}).  This class
  is learned at an exponential rate, despite having infinite cardinality.

\item \textbf{Linear: thresholds on $\R$.}  The class
  $\mathcal{H}_{\mathrm{thr}} = \{x \mapsto \ind[x \geq \theta] :
  \theta \in \R\}$ has $\VC = 1$ but $\Ldim = \infty$
  (\Cref{prop:pac-online-sep}).  It has no infinite VCL tree (the
  one-dimensional ordering prevents shattering along arbitrarily long
  adaptive paths).  The optimal rate is $\Theta(1/n)$.

\item \textbf{Arbitrarily slow: unions of intervals.}  For each $k$, let
  $\mathcal{H}_k$ be the class of unions of at most $k$ intervals on $\R$.
  The class $\mathcal{H}_\infty = \bigcup_k \mathcal{H}_k$ (unions of
  finitely many intervals, with no bound on the number) has
  $\VC < \infty$ but admits infinite VCL trees.  No uniform rate $R(n) \to 0$
  is achievable.
\end{enumerate}
\end{example}


% ================================================================
% SECTION 3: PROOF SKETCH
% ================================================================

\section{Proof Architecture}
\label{sec:proof-sketch}

The full proof of \Cref{thm:trichotomy} occupies much of
\cite{BousquetHannekeMoranVanHandelYehudayoff2021}.  We sketch the key ideas
for each regime, emphasizing why the boundaries fall where they do.


\subsection{The Exponential Case: Finite Littlestone Dimension}
\label{sec:exp-case}

Suppose $\Ldim(\mathcal{H}) = d < \infty$.  The Standard Optimal Algorithm
($\SOA$) from \Cref{ch:online} was designed for adversarial online learning,
but its structure can be adapted to the statistical setting.

The key insight is that a learner with access to i.i.d.\ samples can
simulate a version-space strategy.  Consider the following approach: given a
sample $S$ of size $n$, maintain the version space
$V = \{h \in \mathcal{H} : h \text{ consistent with } S\}$.  Return a
hypothesis from $V$ using an $\SOA$-derived selection rule.

When $\Ldim(\mathcal{H}) = d$, the version space after seeing $n$ i.i.d.\
samples has the following property: the probability (over $D$) that there
exists a hypothesis $h \in V$ with $\risk_D(h) > \varepsilon$ decreases
exponentially in $n$.  This is because each sample point, with probability
at least $\varepsilon$ under $D$, eliminates all hypotheses in $V$ that
disagree with the true label at that point.  After $n$ samples, the survival
probability of any ``bad'' hypothesis (one with risk $> \varepsilon$) is at
most $(1 - \varepsilon)^n \leq e^{-\varepsilon n}$.  The finite Littlestone
dimension controls the effective complexity of $V$ through the mistake tree
structure, ensuring that a union bound over the relevant version space
subsets remains finite.

The result:
\[
  \E[\risk_D(A(S))] \leq C \cdot 2^d \cdot e^{-n/C'}
\]
for constants $C, C'$ depending on $d$.  The rate is exponential in $n$.

The matching lower bound shows that exponential convergence is impossible
when $\Ldim(\mathcal{H}) = \infty$: an infinite Littlestone tree provides an
adversary-like construction (within the distributional framework) that
forces the learner's expected risk to decay no faster than $\Omega(1/n)$.


\subsection{The Linear Case: Infinite Littlestone, Finite VCL}
\label{sec:linear-case}

Suppose $\Ldim(\mathcal{H}) = \infty$ but $\mathcal{H}$ has no infinite VCL
tree.  The exponential strategy fails because the $\SOA$-based approach
requires finite Littlestone dimension.  But a different algorithm---the
\emph{one-inclusion predictor}---achieves the $1/n$ rate.

\begin{defn}[One-Inclusion Predictor (Informal)]
\label{def:oip}
Given a sample $S = \{(x_1, y_1), \ldots, (x_n, y_n)\}$ and a new point
$x$, form the multiset $\{x_1, \ldots, x_n, x\}$.  Consider the one-inclusion
graph $G$: the graph whose vertices are all labelings of this multiset
consistent with some $h \in \mathcal{H}$, with edges connecting labelings
that differ on exactly one point.  Predict the label $\hat{y}$ that
minimizes the maximum degree of the vertex in $G$ corresponding to the
completed labeling.
\end{defn}

Haussler, Littlestone, and Warmuth~\cite{Haussler1988} introduced the
one-inclusion graph technique.  Bousquet et
al.~\cite{BousquetHannekeMoranVanHandelYehudayoff2021} showed that when
$\mathcal{H}$ has no infinite VCL tree, the one-inclusion predictor achieves
\[
  \E[\risk_D(A(S))] \leq \frac{C}{n}
\]
for a constant $C$ depending on $\mathcal{H}$.  The absence of infinite VCL
trees is precisely the combinatorial condition needed for the one-inclusion
graph to have bounded degree growth, which in turn controls the expected
risk.

The matching lower bound is the PAC lower bound: any class with
$\VC(\mathcal{H}) \geq 1$ requires $\Omega(1/n)$ expected risk, since
distinguishing hypotheses that differ on a $1/n$-fraction of the domain
requires $\Omega(n)$ samples.


\subsection{The Slow Case: Infinite VCL Trees}
\label{sec:slow-case}

Suppose $\mathcal{H}$ has an infinite VCL tree.  The theorem claims that no
uniform rate $R(n) \to 0$ is achievable: for every learner $A$ and every
function $R(n) \to 0$, there exists a realizable distribution $D$ such that
$\E[\risk_D(A(S_n))] \geq R(n)$ for infinitely many $n$.

The proof constructs, for any candidate learner $A$ and any target rate
$R(n)$, a distribution that defeats $A$ at the desired rate.  The
construction uses the infinite VCL tree as a source of combinatorial
richness.

Descend the VCL tree adaptively, choosing at each node the branch that is
harder for $A$.  Because the tree is infinite and each path corresponds to a
shattered set, the adversary can build a distribution at any depth $d$ whose
effective VC dimension on the relevant support is $d$.  By choosing $d$
large enough (as a function of $n$), the adversary ensures that the PAC
lower bound $\Omega(d/n)$ exceeds $R(n)$.  The infinite VCL tree provides
arbitrarily large effective VC dimension along adaptive paths, preventing
any uniform rate from holding.

This is the sense in which infinite VCL trees are an obstruction: they
provide an inexhaustible reservoir of complexity that any learner eventually
confronts.


% ================================================================
% SECTION 4: THE CROSS-PARADIGM MAP
% ================================================================

\section{The Cross-Paradigm Map}
\label{sec:cross-paradigm}

The trichotomy theorem connects three chapters of this book through a single
combinatorial object.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
    regime/.style={draw, thick, ellipse, minimum width=3.6cm,
      minimum height=2.2cm, font=\small, align=center},
    example/.style={font=\scriptsize, align=center, text=notegray},
    label/.style={font=\small\bfseries, align=center},
    connection/.style={thick, dashed, notegray},
]
% Three regimes as overlapping regions
\node[regime, fill=sepgreen!10] (exp) at (-3.5, 0) {};
\node[label] at (-3.5, 0.4) {$e^{-n}$};
\node[example] at (-3.5, -0.3) {finite classes\\halfspaces\\conjunctions};

\node[regime, fill=defblue!10] (lin) at (0, 0) {};
\node[label] at (0, 0.4) {$1/n$};
\node[example] at (0, -0.3) {thresholds on $\R$\\intervals on $\R$};

\node[regime, fill=thmred!10] (slow) at (3.5, 0) {};
\node[label] at (3.5, 0.4) {arb.\ slow};
\node[example] at (3.5, -0.3) {$\bigcup_k \mathcal{H}_k$\\unbounded\\unions};

% Boundaries
\draw[very thick, sepgreen!70!black, {Stealth[length=2mm]}-{Stealth[length=2mm]}]
  (-1.6, -1.8) -- (-1.6, 1.8)
  node[above, font=\scriptsize, align=center, text=sepgreen!70!black]
  {$\Ldim < \infty$\\vs.\ $\Ldim = \infty$};

\draw[very thick, thmred!70!black, {Stealth[length=2mm]}-{Stealth[length=2mm]}]
  (1.7, -1.8) -- (1.7, 1.8)
  node[above, font=\scriptsize, align=center, text=thmred!70!black]
  {no $\infty$-VCL\\vs.\ $\infty$-VCL};

% Chapter connections
\node[font=\scriptsize\itshape, text=defblue, align=center] at (-3.5, -2.6)
  {Ch.~6: Online\\(Littlestone dim)};
\node[font=\scriptsize\itshape, text=defblue, align=center] at (0, -2.6)
  {Ch.~5: PAC\\(VC dim)};
\node[font=\scriptsize\itshape, text=defblue, align=center] at (3.5, -2.6)
  {Ch.~9: Universal\\(VCL trees)};

\draw[connection, -{Stealth[length=1.5mm]}] (-3.5, -2.2) -- (-1.8, -1.6);
\draw[connection, -{Stealth[length=1.5mm]}] (0, -2.2) -- (0, -1.2);
\draw[connection, -{Stealth[length=1.5mm]}] (3.5, -2.2) -- (1.9, -1.6);
\end{tikzpicture}
\caption{The three rate regimes and their governing combinatorial conditions.
The left boundary is drawn by the Littlestone dimension (from online
learning, \Cref{ch:online}); the right boundary by VCL trees (a hybrid of
VC shattering and Littlestone trees).  The VC dimension (\Cref{ch:pac})
determines \emph{whether} learning is possible; the Littlestone dimension
and VCL trees determine \emph{how fast}.}
\label{fig:rate-regimes}
\end{figure}

The structural lesson is a hierarchy of three combinatorial conditions,
each finer than the last:

\begin{center}\small
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{llll}
\toprule
\textbf{Condition} & \textbf{Governs} & \textbf{Chapter} & \textbf{Paradigm} \\
\midrule
$\VC(\mathcal{H}) < \infty$ & Learnability (yes/no) & \Cref{ch:pac} & PAC \\
$\Ldim(\mathcal{H}) < \infty$ & Rate: $e^{-n}$ vs.\ $1/n$ & \Cref{ch:online} & Online \\
No infinite VCL tree & Rate: $1/n$ vs.\ arb.\ slow & This chapter & Universal \\
\bottomrule
\end{tabular}
\end{center}

Each row refines the previous one.  The VC dimension draws the coarsest
line (learnable vs.\ not).  The Littlestone dimension draws a finer line
within the learnable classes (fast vs.\ slow).  The VCL tree draws the
finest line (uniformly slow vs.\ hopelessly slow).

This hierarchy was not predicted by any single tradition.  It emerged from
the trichotomy theorem's proof, which required techniques from both the PAC
and online learning literatures.  The result is a unified picture in which
the three paradigms---PAC, online, and universal---are not separate theories
with accidental parallels, but different projections of a single
combinatorial landscape.


\section{What This Chapter Established}
\label{sec:ch9-summary}

Two results, one structural lesson.

\begin{enumerate}[leftmargin=*,itemsep=3pt]
\item \textbf{Universal learning rates.}  The optimal learning rate asks a
  harder question than PAC sample complexity: not how many samples for a
  fixed accuracy, but how fast the risk decreases as a function of $n$,
  uniformly over all realizable distributions.

\item \textbf{The trichotomy.}  Every hypothesis class with
  $|\mathcal{H}| \geq 3$ falls into exactly one of three regimes:
  exponential ($e^{-n}$), linear ($1/n$), or arbitrarily slow.  The
  boundaries are determined by the Littlestone dimension and VCL trees.
  There are no intermediate rates.

\item \textbf{The cross-paradigm bridge.}  The Littlestone dimension---born
  in the adversarial online setting of \Cref{ch:online}---controls the
  boundary between exponential and polynomial convergence in the i.i.d.\
  statistical setting.  This was the central open question resolved by
  Bousquet et al.~\cite{BousquetHannekeMoranVanHandelYehudayoff2021}: not
  just what the rate regimes are, but what combinatorial objects govern
  them---and the answer came from an unexpected paradigm.
\end{enumerate}

