% Chapter 10: Combinatorial Dimensions
% 3 revelations (ds_dimension, natarajan_dimension, star_number),
% 10 definitions with brief treatment.
% Technical core of Part III. Full Sauer--Shelah proof. Multiclass detective story.

\chapter{Combinatorial Dimensions}
\label{ch:dimensions}

Part~II characterized learnability in four paradigms, and in each case the
answer was a number: the VC dimension for PAC learning, the Littlestone
dimension for online learning, the mind-change ordinal for Gold-style
identification, the teaching dimension for exact learning.  This chapter and
the three that follow develop these numbers---and the two dozen others in the
concept graph---as objects of study in their own right.

The present chapter treats \emph{combinatorial dimensions}: quantities defined
by asking how richly a hypothesis class can label finite configurations of
points.  The simplest such quantity, the VC dimension, asks how many points
can be shattered.  The others refine, extend, or replace this question for
settings beyond binary classification.  We prove the two foundational
combinatorial results in full (the Sauer--Shelah lemma and the
VC dimension characterization already established in \Cref{ch:pac}), recall the
Littlestone dimension from \Cref{ch:online}, and then turn to the chapter's
narrative centerpiece: the thirty-year search for the correct combinatorial
dimension characterizing multiclass learnability, which ended only in 2022.

\medskip

The chapter is organized as follows.

\begin{enumerate}[leftmargin=*,itemsep=2pt]
\item \textbf{VC dimension and shattering} (\Cref{sec:vc-dim}): the
  definitions, the growth function, and the full proof of the Sauer--Shelah
  lemma.
\item \textbf{The Littlestone dimension} (\Cref{sec:ldim-recall}): a brief
  recall from \Cref{ch:online} and the relationship $\VC \leq \Ldim$.
\item \textbf{Beyond binary: pseudodimension and fat-shattering}
  (\Cref{sec:beyond-binary}): scale-sensitive dimensions for real-valued
  function classes.
\item \textbf{The multiclass story: from Natarajan to DS}
  (\Cref{sec:multiclass}): the detective story of multiclass PAC
  characterization.
\item \textbf{Other dimensions} (\Cref{sec:other-dims}): a concise catalog of
  further combinatorial dimensions, with forward references.
\end{enumerate}


% ================================================================
% SECTION 1: VC DIMENSION AND SHATTERING
% ================================================================

\section{VC Dimension and Shattering}
\label{sec:vc-dim}

The definitions in this section were introduced informally in \Cref{ch:objects}
and used without proof in \Cref{ch:pac}.  We now state them precisely and prove
the combinatorial backbone that supports the Fundamental Theorem.

\begin{defn}[Restriction and Shattering]
\label{def:shattering}
Let $\mathcal{H} \subseteq \{0,1\}^X$ be a hypothesis class and let
$S = \{x_1, \ldots, x_n\} \subseteq X$ be a finite set.  The
\emph{restriction} of $\mathcal{H}$ to $S$ is
\[
  \mathcal{H}|_S \;=\; \bigl\{\bigl(h(x_1), \ldots, h(x_n)\bigr) : h \in \mathcal{H}\bigr\}
  \;\subseteq\; \{0,1\}^n.
\]
We say $\mathcal{H}$ \emph{shatters} $S$ if $\mathcal{H}|_S = \{0,1\}^n$,
i.e., every labeling of $S$ is realized by some hypothesis in $\mathcal{H}$.
\end{defn}

\begin{defn}[VC Dimension {\cite{VapnikChervonenkis1971}}]
\label{def:vc-dim}
The \emph{Vapnik--Chervonenkis dimension} of $\mathcal{H}$, denoted
$\VC(\mathcal{H})$, is the largest cardinality of a set shattered by
$\mathcal{H}$:
\[
  \VC(\mathcal{H}) \;=\; \max\bigl\{|S| : S \subseteq X,\; \mathcal{H}
  \text{ shatters } S\bigr\}.
\]
If $\mathcal{H}$ shatters arbitrarily large finite sets,
$\VC(\mathcal{H}) = \infty$.
\end{defn}

\begin{example}[Halfplanes in $\R^2$]
\label{ex:halfplanes-shatter}
Let $\mathcal{H}$ be the class of halfplanes in $\R^2$:
$\mathcal{H} = \bigl\{x \mapsto \ind[\langle w, x \rangle \geq b] : w \in
\R^2,\, b \in \R\bigr\}$.  Any three non-collinear points are shattered
(\Cref{fig:shatter-halfplane}), but no four points can be: for any four
points, at least one is inside the convex hull of the other three, and the
labeling that assigns it the opposite label from the other three cannot be
realized by a halfplane.  Hence $\VC(\mathcal{H}) = 3$.
\end{example}

\begin{figure}[ht]
\centering
\begin{tikzpicture}[scale=0.72,
    point/.style={circle, fill=defblue, inner sep=1.8pt},
    pospt/.style={circle, fill=thmred, inner sep=1.8pt},
    negpt/.style={circle, fill=defblue, inner sep=1.8pt},
    label0/.style={font=\scriptsize, text=defblue},
    label1/.style={font=\scriptsize, text=thmred},
]
  % 8 labelings of 3 points arranged in 2 rows of 4
  \foreach \col/\la/\lb/\lc in {
    0/0/0/0, 1/0/0/1, 2/0/1/0, 3/0/1/1,
    4/1/0/0, 5/1/0/1, 6/1/1/0, 7/1/1/1%
  } {
    \pgfmathtruncatemacro{\cx}{mod(\col,4)*3}
    \pgfmathtruncatemacro{\cy}{-int(\col/4)*3}
    \begin{scope}[shift={(\cx,\cy)}]
      % Draw the three points
      \ifnum\la=1
        \node[pospt] (a\col) at (0,0) {};
        \node[label1, below=1pt] at (a\col) {$1$};
      \else
        \node[negpt] (a\col) at (0,0) {};
        \node[label0, below=1pt] at (a\col) {$0$};
      \fi
      \ifnum\lb=1
        \node[pospt] (b\col) at (2,0) {};
        \node[label1, below=1pt] at (b\col) {$1$};
      \else
        \node[negpt] (b\col) at (2,0) {};
        \node[label0, below=1pt] at (b\col) {$0$};
      \fi
      \ifnum\lc=1
        \node[pospt] (c\col) at (1,1.5) {};
        \node[label1, above=1pt] at (c\col) {$1$};
      \else
        \node[negpt] (c\col) at (1,1.5) {};
        \node[label0, above=1pt] at (c\col) {$0$};
      \fi
    \end{scope}
  }
\end{tikzpicture}
\caption{All $2^3 = 8$ labelings of three non-collinear points in $\R^2$.
Each labeling is realized by some halfplane, so the points are shattered.
Red (filled dark) indicates label~$1$; blue indicates label~$0$.}
\label{fig:shatter-halfplane}
\end{figure}


\subsection{The Growth Function}

The growth function measures how many distinct labelings $\mathcal{H}$ can
produce on sets of a given size.

\begin{defn}[Growth Function]
\label{def:growth-function}
The \emph{growth function} (or \emph{shatter coefficient}) of $\mathcal{H}$ is
\[
  \growth_\mathcal{H}(n) \;=\; \max_{S \subseteq X,\; |S|=n}
  \bigl|\mathcal{H}|_S\bigr|.
\]
Equivalently, $\growth_\mathcal{H}(n)$ is the maximum number of distinct
behaviors $\mathcal{H}$ can exhibit on any $n$ points.
\end{defn}

The growth function satisfies $\growth_\mathcal{H}(n) \leq 2^n$ always, with
equality when $\mathcal{H}$ shatters some set of size~$n$.  The remarkable
content of the Sauer--Shelah lemma is that once shattering fails---at
$n = \VC(\mathcal{H}) + 1$---the growth function drops from exponential to
polynomial \emph{immediately and permanently}.


\subsection{The Sauer--Shelah Lemma}

\begin{lemma}[Sauer--Shelah {\cite{Sauer1972,Shelah1972}}]
\label{lem:sauer-shelah}
Let $\mathcal{H} \subseteq \{0,1\}^X$ with $\VC(\mathcal{H}) = d$.  Then for
all $n \geq 1$,
\[
  \growth_\mathcal{H}(n) \;\leq\; \sum_{i=0}^{d} \binom{n}{i}.
\]
In particular, $\growth_\mathcal{H}(n) \leq \bigl(\frac{en}{d}\bigr)^d$ for
$n \geq d \geq 1$, so the growth function is polynomial in $n$ when
$d < \infty$.
\end{lemma}

The proof is by double induction and is one of the most elegant arguments in
combinatorics.  The key idea is to split the hypothesis class along one
coordinate and apply the inductive hypothesis to the two resulting classes,
whose VC dimensions are carefully controlled.

\begin{proof}
\label{lem:sauer-shelah-proof}
We prove a stronger statement: for any finite set $S$ with $|S| = n$ and
any class $\mathcal{H}' \subseteq \{0,1\}^S$ (not necessarily arising from
a restriction),
\[
  |\mathcal{H}'| \;\leq\; \bigl|\{A \subseteq S : \mathcal{H}' \text{ shatters } A\}\bigr|.
  \tag{$\star$}
\]
This implies the lemma: if $\VC(\mathcal{H}') \leq d$, then the right-hand
side counts only subsets $A$ with $|A| \leq d$, of which there are at most
$\sum_{i=0}^{d}\binom{n}{i}$.

We prove $(\star)$ by induction on $|S|$.

\smallskip\noindent
\textbf{Base case} ($|S| = 1$, say $S = \{x\}$).  If
$\mathcal{H}' = \{0\}$ or $\mathcal{H}' = \{1\}$, then $|\mathcal{H}'| = 1$
and $\mathcal{H}'$ shatters $\emptyset$, giving at least one shattered set.
If $\mathcal{H}' = \{0,1\}$, then $|\mathcal{H}'| = 2$ and $\mathcal{H}'$
shatters both $\emptyset$ and $\{x\}$, giving at least two shattered sets.
In both cases $(\star)$ holds.

\smallskip\noindent
\textbf{Inductive step.}  Let $|S| = n > 1$.  Pick any element $x \in S$
and set $S' = S \setminus \{x\}$.  Partition $\mathcal{H}'$ according to
behavior on $x$:
\begin{align*}
  \mathcal{H}_0 &= \{h|_{S'} : h \in \mathcal{H}',\; h(x) = 0\}, \\
  \mathcal{H}_1 &= \{h|_{S'} : h \in \mathcal{H}',\; h(x) = 1\}, \\
  \mathcal{H}_\cap &= \mathcal{H}_0 \cap \mathcal{H}_1.
\end{align*}
Here $\mathcal{H}_\cap$ consists of the functions on $S'$ that appear with
\emph{both} labels on $x$.

Count: each $h \in \mathcal{H}'$ contributes to exactly one of $\mathcal{H}_0$
or $\mathcal{H}_1$ (as a function on $S'$), but a function in
$\mathcal{H}_\cap$ is contributed by two elements of $\mathcal{H}'$ (one
with $h(x)=0$, one with $h(x)=1$).  Therefore,
\[
  |\mathcal{H}'| \;=\; |\mathcal{H}_0| + |\mathcal{H}_1| - |\mathcal{H}_\cap|
  + |\mathcal{H}_\cap|
  \;=\; |\mathcal{H}_0 \cup \mathcal{H}_1| + |\mathcal{H}_\cap|.
\]
More precisely, $|\mathcal{H}'| = |\mathcal{H}_0| + |\mathcal{H}_1|$, since
functions in $\mathcal{H}_0 \setminus \mathcal{H}_\cap$ appear only from
$h(x) = 0$, functions in $\mathcal{H}_1 \setminus \mathcal{H}_\cap$ appear
only from $h(x) = 1$, and functions in $\mathcal{H}_\cap$ appear twice.  So
\[
  |\mathcal{H}'| = |\mathcal{H}_0 \cup \mathcal{H}_1| + |\mathcal{H}_\cap|.
\]

Apply the inductive hypothesis (on $S'$, which has $n-1$ elements) to both
$\mathcal{H}_0 \cup \mathcal{H}_1$ and $\mathcal{H}_\cap$:
\begin{align*}
  |\mathcal{H}_0 \cup \mathcal{H}_1|
    &\leq \bigl|\{A \subseteq S' : (\mathcal{H}_0 \cup \mathcal{H}_1) \text{ shatters } A\}\bigr|, \\
  |\mathcal{H}_\cap|
    &\leq \bigl|\{A \subseteq S' : \mathcal{H}_\cap \text{ shatters } A\}\bigr|.
\end{align*}

Now we connect the shattered sets of the restricted classes back to those
of $\mathcal{H}'$:

\begin{enumerate}[leftmargin=*,itemsep=3pt]
\item If $A \subseteq S'$ is shattered by $\mathcal{H}_0 \cup \mathcal{H}_1$
  but \emph{not} by $\mathcal{H}_\cap$, then $A$ is shattered by $\mathcal{H}'$
  as well (since every labeling of $A$ is realized by some $h|_{S'} \in
  \mathcal{H}_0 \cup \mathcal{H}_1$, and at least one such $h$ exists in
  $\mathcal{H}'$).

\item If $A \subseteq S'$ is shattered by $\mathcal{H}_\cap$, then every
  labeling of $A$ is realized by functions appearing with both labels on $x$.
  Hence $\mathcal{H}'$ shatters $A \cup \{x\}$: for any labeling of
  $A \cup \{x\}$, pick the label $b$ assigned to $x$, then pick $h|_{S'} \in
  \mathcal{H}_\cap$ realizing the labeling of $A$; the corresponding
  $h \in \mathcal{H}'$ with $h(x) = b$ realizes the full labeling.
\end{enumerate}

Crucially, the sets in (1) do not contain $x$, and the sets in (2) do
contain $x$, so they are disjoint families.  Therefore,
\[
  |\mathcal{H}'|
  \;\leq\; \bigl|\{A \subseteq S : \mathcal{H}' \text{ shatters } A,\;
  x \notin A\}\bigr|
  + \bigl|\{A \subseteq S : \mathcal{H}' \text{ shatters } A,\;
  x \in A\}\bigr|
  \;=\; \bigl|\{A \subseteq S : \mathcal{H}' \text{ shatters } A\}\bigr|.
\]
This completes the induction.
\end{proof}

\begin{remark}
The lemma was proved independently by Sauer~\cite{Sauer1972} and
Shelah~\cite{Shelah1972}, with a related result by Perles and Shelah
appearing earlier.  Vapnik and Chervonenkis~\cite{VapnikChervonenkis1971}
proved a weaker bound in their 1971 paper.  The proof above follows the
elegant ``number of shattered sets'' formulation due to Alon and
Frankl.
\end{remark}

\begin{figure}[ht]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=11cm, height=6.5cm,
    xlabel={$n$},
    ylabel={Number of labelings},
    xmin=0, xmax=18, ymin=0, ymax=600,
    xtick={0,2,...,18},
    ytick={0,100,...,600},
    legend pos=north west,
    legend style={font=\small},
    grid=major,
    grid style={gray!20},
    thick,
]
% 2^n
\addplot[domain=0:9.2, samples=50, thick, thmred, dashed]
    {2^x};
\addlegendentry{$2^n$}

% Sauer--Shelah bound for d=3
\addplot[domain=0:18, samples=100, thick, defblue, solid]
    {1 + x + x*(x-1)/2 + x*(x-1)*(x-2)/6};
\addlegendentry{$\sum_{i=0}^{3}\binom{n}{i}$ \quad($d=3$)}

% Mark d+1 = 4
\draw[thick, dashed, notegray] (axis cs:3,0) -- (axis cs:3,600);
\node[font=\scriptsize, text=notegray, anchor=south west] at (axis cs:3.1,520)
    {$n = d = 3$};

% Mark where they split
\draw[thick, dashed, notegray] (axis cs:4,0) -- (axis cs:4,600);
\node[font=\scriptsize, text=notegray, anchor=south west] at (axis cs:4.1,450)
    {$n = d+1$: bound bites};

\end{axis}
\end{tikzpicture}
\caption{The growth function $\growth_\mathcal{H}(n)$ versus $2^n$.  For
$n \leq d = \VC(\mathcal{H})$, the growth function can equal $2^n$ (if
$\mathcal{H}$ shatters some set of that size).  At $n = d + 1$, the
Sauer--Shelah bound forces a permanent drop to polynomial growth.  The
polynomial bound $\sum_{i=0}^{d}\binom{n}{i}$ is plotted for $d = 3$.}
\label{fig:growth-function}
\end{figure}

The Sauer--Shelah lemma is the combinatorial engine behind the Fundamental
Theorem of Statistical Learning (\Cref{thm:fundamental}): it converts
a finite VC dimension into a polynomial bound on the growth function, which in
turn yields uniform convergence of empirical risks via a symmetrization
argument.


% ================================================================
% SECTION 2: THE LITTLESTONE DIMENSION (CROSS-REFERENCE)
% ================================================================

\section{The Littlestone Dimension}
\label{sec:ldim-recall}

The VC dimension measures shatterability of \emph{sets}: it asks for the
largest set whose every labeling is realized.  The Littlestone dimension,
introduced in \Cref{ch:online} and fully characterized there, measures
shatterability of \emph{trees}: it asks for the deepest complete binary tree
whose every root-to-leaf path is realized.  We recall the definition for
reference.

\begin{defn}[Mistake Tree --- Recall]
\label{def:mistake-tree-recall}
A \emph{mistake tree} for $\mathcal{H} \subseteq \{0,1\}^X$ is a complete
binary tree in which each internal node is labeled with an instance $x \in X$,
left edges correspond to label~$0$, right edges to label~$1$, and every
root-to-leaf path is consistent with some $h \in \mathcal{H}$
(see \Cref{def:mistake-tree}).
\end{defn}

\begin{defn}[Littlestone Dimension --- Recall]
\label{def:ldim-recall}
The \emph{Littlestone dimension} $\Ldim(\mathcal{H})$ is the maximum depth
of a mistake tree for $\mathcal{H}$ (\Cref{def:ldim}).  The
Littlestone characterization (\Cref{thm:littlestone-characterization}) establishes that
$\Ldim(\mathcal{H})$ equals the optimal worst-case mistake bound.
\end{defn}

The relationship between the two fundamental dimensions is strict:

\begin{prop}
\label{prop:vc-leq-ldim}
For any hypothesis class $\mathcal{H}$,
$\VC(\mathcal{H}) \leq \Ldim(\mathcal{H})$.
\end{prop}

\begin{proof}
If $\mathcal{H}$ shatters $\{x_1, \ldots, x_d\}$, one can construct a
mistake tree of depth $d$ by placing $x_i$ at every node of depth $i$:
since every labeling is realized, every root-to-leaf path is consistent with
some $h \in \mathcal{H}$.  Hence any shattered set yields a mistake tree of
the same depth.
\end{proof}

The converse fails: threshold classifiers on $\R$ have $\VC = 1$ but
$\Ldim = \infty$ (\Cref{ch:separations}).  Intuitively, the Littlestone
dimension is larger because it accounts for adaptive adversaries: the
adversary in the online game chooses instances sequentially, seeing the
learner's responses, while the VC dimension considers all labelings of a
fixed set simultaneously.


% ================================================================
% SECTION 3: BEYOND BINARY
% ================================================================

\section{Beyond Binary: Pseudodimension and Fat-Shattering}
\label{sec:beyond-binary}

When the hypothesis class consists of real-valued functions
$\mathcal{F} \subseteq \R^X$ rather than binary classifiers, shattering must
be redefined.  Two progressively refined notions capture the relevant
complexity.

\begin{defn}[Pseudodimension]
\label{def:pseudodim}
Let $\mathcal{F} \subseteq \R^X$.  A set $S = \{x_1, \ldots, x_n\} \subseteq X$
is \emph{pseudo-shattered} (or \emph{P-shattered}) by $\mathcal{F}$ if there
exist thresholds $t_1, \ldots, t_n \in \R$ such that, for every labeling
$b \in \{0,1\}^n$, there exists $f \in \mathcal{F}$ with
\[
  \mathrm{sgn}\bigl(f(x_i) - t_i\bigr) = (-1)^{1 - b_i}
  \quad \text{for all } i = 1, \ldots, n.
\]
The \emph{pseudodimension} $\Pdim(\mathcal{F})$ is the largest cardinality of
a P-shattered set.
\end{defn}

The thresholds $t_i$ act as ``witnesses'': they reduce a real-valued
shattering problem to a binary one.  When $\mathcal{F}$ is already
$\{0,1\}$-valued, the thresholds $t_i = 1/2$ recover the standard VC
dimension, so $\Pdim$ is a proper generalization.

\begin{defn}[Fat-Shattering Dimension {\cite{AlonBenDavidCesaBianchiHaussler1997}}]
\label{def:fat-shattering}
Let $\mathcal{F} \subseteq \R^X$ and $\gamma > 0$.  A set
$S = \{x_1, \ldots, x_n\}$ is \emph{$\gamma$-fat-shattered} by $\mathcal{F}$
if there exist thresholds $t_1, \ldots, t_n \in \R$ such that, for every
labeling $b \in \{0,1\}^n$, there exists $f \in \mathcal{F}$ with
\[
  f(x_i) \;\geq\; t_i + \gamma \;\;\text{when } b_i = 1,
  \qquad
  f(x_i) \;\leq\; t_i - \gamma \;\;\text{when } b_i = 0,
\]
for all $i$.  The \emph{fat-shattering dimension at scale $\gamma$}, denoted
$\fatshat_\gamma(\mathcal{F})$, is the largest cardinality of a
$\gamma$-fat-shattered set.
\end{defn}

The margin parameter $\gamma$ makes fat-shattering \emph{scale-sensitive}:
large margins are harder to achieve, so
$\fatshat_\gamma(\mathcal{F}) \leq \fatshat_{\gamma'}(\mathcal{F})$ whenever
$\gamma \geq \gamma'$.  As $\gamma \to 0$, the fat-shattering dimension
approaches the pseudodimension.

\begin{thm}[Characterization of Real-Valued Learnability
  {\cite{AlonBenDavidCesaBianchiHaussler1997}}]
\label{thm:fat-char}
A real-valued function class $\mathcal{F} \subseteq [0,1]^X$ is
\emph{agnostically PAC learnable} (with respect to the absolute loss) if and
only if $\fatshat_\gamma(\mathcal{F}) < \infty$ for every $\gamma > 0$.
\end{thm}

This is the real-valued analogue of the Fundamental Theorem: just as finite VC
dimension characterizes binary PAC learnability, finite fat-shattering
dimension at every scale characterizes real-valued learnability.  The
pseudodimension alone is not sufficient for characterization in the agnostic
setting, as it ignores the scale structure.

\begin{remark}[Dimension hierarchy]
For a class $\mathcal{F} \subseteq [0,1]^X$:
\[
  \fatshat_\gamma(\mathcal{F})
  \;\leq\; \Pdim(\mathcal{F})
  \;\leq\; \VC\!\bigl(\mathrm{subgraph}(\mathcal{F})\bigr)
\]
where $\mathrm{subgraph}(\mathcal{F}) = \{(x,t) \mapsto \ind[f(x) \geq t] :
f \in \mathcal{F}\}$.  When $\mathcal{F}$ is $\{0,1\}$-valued, all three
quantities coincide with $\VC(\mathcal{F})$.
\end{remark}


% ================================================================
% SECTION 4: THE MULTICLASS STORY
% ================================================================

\section{The Multiclass Story: From Natarajan to DS}
\label{sec:multiclass}

Binary classification has a clean answer: a class $\mathcal{H} \subseteq
\{0,1\}^X$ is PAC learnable if and only if $\VC(\mathcal{H}) < \infty$.
What is the corresponding answer when the label set $Y$ has $k \geq 3$
elements?  This question, first posed in the 1980s, was open for thirty
years.  Its resolution required not merely a new proof technique but a
fundamentally new combinatorial dimension.  The story is worth telling in
full, because the obvious candidate turned out to be wrong.


\subsection{The Natarajan Dimension: The Obvious Candidate}

\begin{defn}[Natarajan Dimension]
\label{def:natarajan}
Let $\mathcal{H} \subseteq Y^X$ with $|Y| \geq 2$.  The \emph{Natarajan
dimension} $\Ndim(\mathcal{H})$ is the largest $d$ such that there exist
\begin{itemize}[leftmargin=*,itemsep=1pt]
\item a set $S = \{x_1, \ldots, x_d\} \subseteq X$, and
\item two functions $f_0, f_1 \colon S \to Y$ with $f_0(x_i) \neq f_1(x_i)$
  for all $i$,
\end{itemize}
such that for every labeling $b \in \{0,1\}^d$, there exists $h \in
\mathcal{H}$ with $h(x_i) = f_{b_i}(x_i)$ for all $i$.
\end{defn}

The Natarajan dimension generalizes the VC dimension in a natural way: instead
of requiring all $2^d$ binary labelings, it requires all $2^d$ labelings
drawn from two specified ``opposing'' functions.  When $|Y| = 2$, the
Natarajan dimension equals the VC dimension exactly.

In the 1980s and 1990s, the working conjecture was:
\begin{quote}
\itshape A multiclass hypothesis class $\mathcal{H} \subseteq Y^X$ is PAC
learnable if and only if $\Ndim(\mathcal{H}) < \infty$.
\end{quote}
This conjecture was supported by partial results.  Ben-David, Cesa-Bianchi,
Haussler, and Long proved that finite Natarajan dimension is
\emph{sufficient} for multiclass PAC learnability when combined with a bound
on $|Y|$.  The ``if'' direction seemed solid.  The conjecture survived for
decades without a counterexample.

It was wrong.


\subsection{The Counterexample: When the Obvious Fails}

\begin{thm}[Failure of Natarajan Dimension
  {\cite{DanielyShalevShwartz2014}}]
\label{thm:natarajan-failure}
There exists a hypothesis class $\mathcal{H} \subseteq Y^X$ with
$\Ndim(\mathcal{H}) = 1$ that is \emph{not} PAC learnable.
\end{thm}

This is a dramatic failure: the Natarajan dimension can be as small as
possible (just one) and still the class can be unlearnable.  The problem lies
in the label set: the construction uses $|Y| = \infty$, and the Natarajan
dimension, which examines only pairs of opposing functions, fails to capture
the combinatorial explosion that arises when the label set is large.

The witness is a class constructed via a combinatorial argument.  The key
insight is that the Natarajan dimension constrains the number of ``binary
slices'' through the labeling space but does not constrain the total
complexity when the label set grows without bound.

\begin{separation}
\textbf{Natarajan vs.\ PAC learnability.}  The Natarajan dimension does not
characterize multiclass PAC learnability:
\[
  \Ndim(\mathcal{H}) < \infty
  \quad\not\!\!\!\implies\quad
  \mathcal{H} \text{ is PAC learnable}.
\]
The gap is witnessed by a class with $\Ndim = 1$, $|Y| = \infty$, that
requires unbounded sample complexity.  The correct characterization requires
the DS dimension (\Cref{def:ds-dim} below).
\end{separation}


\subsection{The DS Dimension: The Correct Answer}

\begin{defn}[DS Dimension {\cite{BrukhimCarmonDinurMoranYehudayoff2022}}]
\label{def:ds-dim}
Let $\mathcal{H} \subseteq Y^X$.  The \emph{DS dimension} (named for
Daniely and Shalev-Shwartz) $\DSdim(\mathcal{H})$ is defined via
pseudo-cubes in the one-inclusion hypergraph of $\mathcal{H}$.

Construct the \emph{one-inclusion hypergraph} $G_\mathcal{H}$ on a finite set
$S = \{x_1, \ldots, x_n\} \subseteq X$ as follows: the vertices are the
hypotheses $\mathcal{H}|_S$ (restrictions of $\mathcal{H}$ to $S$), and for
each $i \in \{1, \ldots, n\}$, add a hyperedge connecting all hypotheses
that agree on $S \setminus \{x_i\}$ (i.e., they differ only at $x_i$).

A \emph{pseudo-cube of dimension $d$} in $G_\mathcal{H}$ is a subhypergraph
isomorphic to the complete $d$-dimensional hypercube graph $\{0,1\}^d$.

The DS dimension $\DSdim(\mathcal{H})$ is the supremum over all finite
$S \subseteq X$ of the maximum dimension of a pseudo-cube in $G_\mathcal{H}$.
\end{defn}

\begin{thm}[Multiclass Characterization
  {\cite{BrukhimCarmonDinurMoranYehudayoff2022}}]
\label{thm:ds-char}
A hypothesis class $\mathcal{H} \subseteq Y^X$ is PAC learnable if and only
if $\DSdim(\mathcal{H}) < \infty$.
\end{thm}

This theorem, proved by Brukhim, Chepoi, Daniel, Moran, and Yehudayoff in
2022, resolved the multiclass characterization problem.  It is the
multiclass analogue of the Fundamental Theorem: the DS dimension plays
exactly the role that the VC dimension plays for binary classification.

\begin{historical}
\textbf{Why the Natarajan dimension fails and the DS dimension succeeds.}
The Natarajan dimension examines the labeling structure of $\mathcal{H}$ by
projecting onto pairs of opposing functions---a fundamentally binary lens
applied to a multiclass problem.  The DS dimension, by contrast, examines
the full combinatorial structure of the one-inclusion hypergraph, which
encodes all the ways hypotheses can disagree on single points.

The one-inclusion hypergraph is not new: Haussler, Littlestone, and Warmuth
introduced it in 1994 for binary classification, where it yields an optimal
PAC learner (the one-inclusion algorithm).  What Brukhim et al.\ discovered
is that the right way to measure the complexity of a multiclass hypothesis
class is to look for high-dimensional pseudo-cubes inside this hypergraph.
Pseudo-cubes are the multiclass analogue of shattered sets: they represent
configurations where the labeling structure is rich enough to prevent
learning.
\end{historical}

The dimension hierarchy for multiclass problems is:

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
    dimbox/.style={draw, thick, rounded corners=3pt, minimum width=2.6cm,
      minimum height=0.9cm, font=\small\bfseries, align=center},
    arr/.style={-{Stealth[length=2.5mm]}, thick},
    gaplabel/.style={font=\scriptsize\itshape, text=thmred, align=center},
    implabel/.style={font=\scriptsize, text=notegray, align=center},
]
  \node[dimbox, fill=defblue!10] (vc) at (0,0) {$\VC$\\{\scriptsize binary}};
  \node[dimbox, fill=edgeorange!10] (nat) at (4.5,0) {$\Ndim$\\{\scriptsize multiclass}};
  \node[dimbox, fill=sepgreen!10] (ds) at (9,0) {$\DSdim$\\{\scriptsize multiclass}};

  \draw[arr] (vc) -- node[implabel, above] {generalizes\\to $|Y| \geq 2$} (nat);
  \draw[arr] (nat) -- node[implabel, above] {$\Ndim < \infty$\\$\Leftarrow$ $\DSdim < \infty$} (ds);

  \node[gaplabel, below=0.5cm of nat] (gap)
    {$\Ndim < \infty \not\!\Rightarrow$ learnable\\
     witness: $\Ndim = 1$, not learnable};
  \draw[thick, dashed, thmred] (nat.south) -- (gap.north);

  \node[implabel, below=0.5cm of ds] (correct)
    {$\DSdim < \infty \Leftrightarrow$ learnable\\
     {\scriptsize\upshape (Brukhim et al.\ 2022)}};
  \draw[thick, sepgreen] (ds.south) -- (correct.north);

  \node[implabel, below=0.5cm of vc] (vcnote)
    {$\VC < \infty \Leftrightarrow$ learnable\\
     {\scriptsize\upshape (binary only)}};
  \draw[thick, defblue] (vc.south) -- (vcnote.north);
\end{tikzpicture}
\caption{The multiclass dimension hierarchy.  The Natarajan dimension was the
leading candidate for thirty years, but finite $\Ndim$ does not imply
learnability.  The DS dimension is the correct characterization.}
\label{fig:dimension-hierarchy}
\end{figure}


% ================================================================
% SECTION 5: OTHER DIMENSIONS
% ================================================================

\section{Other Dimensions}
\label{sec:other-dims}

Several further combinatorial dimensions appear in the concept graph.  We
collect them here in catalog form; each is developed more fully in the chapter
where it plays a characterizing role.

\begin{table}[ht]
\centering
\small
\caption{Catalog of combinatorial dimensions beyond VC, Littlestone,
pseudodimension, fat-shattering, Natarajan, and DS.}
\label{tab:other-dims}
\begin{tabularx}{\textwidth}{@{}l l X l@{}}
\toprule
\textbf{Dimension} & \textbf{Notation} & \textbf{What it measures} & \textbf{Reference} \\
\midrule
Star number
  & $\mathfrak{s}(\mathcal{H})$
  & The maximum number of sets in a sunflower whose kernel is not shattered.
    Finiteness of $\mathfrak{s}(\mathcal{H})$ is \emph{equivalent} to
    $\VC(\mathcal{H}) < \infty$---a surprising alternative
    combinatorial characterization of PAC learnability.
  & \Cref{ch:pac} \\
\addlinespace
Eluder dimension
  & $\dim_E(\mathcal{F})$
  & The length of the longest sequence of points such that each point is
    ``independent'' of the previous ones, given the function class.
    Measures exploration complexity in sequential decision problems.
  & \Cref{ch:extensions} \\
\addlinespace
SQ dimension
  & $\SQdim(\mathcal{C})$
  & The largest set of nearly uncorrelated concepts.  Determines the number
    of statistical queries needed to learn $\mathcal{C}$ in the statistical
    query model, where the learner receives expectations
    $\E[\phi(x,y)]$ rather than individual examples.
  & \Cref{ch:extensions} \\
\addlinespace
VCL tree
  & ---
  & A tree structure used in the trichotomy theorem (\Cref{ch:universal})
    to separate exponential from arbitrarily slow learners.  Combines VC
    and Littlestone structure.
  & \Cref{ch:universal} \\
\addlinespace
Ordinal VC dim
  & $\VC^\alpha$
  & A transfinite refinement of VC dimension, using ordinal-indexed
    Cantor--Bendixson ranks on the space of consistent hypotheses.
  & \Cref{ch:ordinals} \\
\addlinespace
Ordinal Littlestone dim
  & $\Ldim^\alpha$
  & A transfinite refinement of Littlestone dimension, paralleling
    ordinal VC dimension for the online setting.
  & \Cref{ch:ordinals} \\
\bottomrule
\end{tabularx}
\end{table}

\begin{remark}[Star number equivalence]
\label{rmk:star-number}
The equivalence between finite star number and finite VC dimension is
not obvious.  The star number counts a sunflower-type configuration that,
on its face, looks quite different from shattering.  That these two
combinatorial conditions coincide is one of the surprising structural
facts of the theory: there are \emph{multiple, independently motivated}
combinatorial properties of a hypothesis class that all turn out to be
equivalent to PAC learnability.
\end{remark}


% ================================================================
% CLOSING REMARKS
% ================================================================

\bigskip
\begin{center}
\rule{0.5\textwidth}{0.4pt}
\end{center}
\bigskip

The combinatorial dimensions of this chapter are the coordinates of a map.
The VC dimension locates a class in the PAC landscape; the Littlestone
dimension locates it in the online landscape; the DS dimension locates it in
the multiclass landscape; the fat-shattering dimension locates it in the
real-valued landscape.  No single dimension suffices for all purposes, because
the landscapes are genuinely different---as the separation results of
\Cref{ch:separations} will make precise.

The next chapter turns from combinatorial dimensions to a different family of
complexity measures: those based on \emph{compression} and \emph{sample
complexity}.
