% Chapter 11: Sample Complexity, Compression, and Occam's Razor
% sample_complexity = Profile B-light, unlabeled_compression = Profile D (Negative),
% occam_algorithm = Profile B-light, description_length/kolmogorov/kl = Profile A,
% covering_number = Profile A-plus.
%
% Texture: question-driven. The open conjecture (compression ≈ VC dimension)
% is the gravitational center. Everything else orbits it: sample complexity
% provides the motivation, Occam's razor provides the algorithmic bridge,
% information-theoretic notions provide the vocabulary. The chapter should
% feel like a detective story where the case remains unsolved.

\chapter{Sample Complexity, Compression, and Occam's Razor}
\label{ch:compression}

The Fundamental Theorem of \Cref{ch:pac} answers the qualitative question:
\emph{which} classes are PAC learnable?  This chapter turns to the quantitative
question: \emph{how many samples} does learning require, and what structural
properties of the hypothesis class control the answer?

The quantitative question leads, unexpectedly, to an open problem.  The tight
sample complexity bounds (\Cref{sec:sample-bounds}) show that the VC dimension
$d$ determines the sample complexity up to constant factors.  Compression
schemes (\Cref{sec:compression}) provide a different explanation of
learnability: a class is learnable if and only if the training data can be
``compressed'' to a small subset that determines the hypothesis.  The natural
conjecture---that compression of size $O(d)$ always suffices---has been open
since 1986.  It is one of the oldest and most embarrassing open problems in
learning theory, and it is the centerpiece of this chapter.

Occam's razor (\Cref{sec:occam}) provides the bridge between compression and
generalization: any algorithm that finds a ``short'' description of the data
generalizes.  The information-theoretic notions of
\Cref{sec:info-foundations}---description length, Kolmogorov complexity,
covering numbers---supply the vocabulary for making ``short'' precise.

\medskip

The chapter is organized around a single question:

\begin{center}
\itshape
Does every concept class of VC dimension $d$ admit a compression scheme of
size $O(d)$?
\end{center}

Everything we prove either approaches this question or illuminates why it
resists resolution.


% ================================================================
% SECTION 1: TIGHT SAMPLE COMPLEXITY BOUNDS
% Profile B-light. State the bounds, give proof sketch for upper bound.
% ================================================================

\section{Tight Sample Complexity Bounds}
\label{sec:sample-bounds}

The sample complexity of PAC learning was bounded in \Cref{ch:pac} through the
VC characterization.  Here we state the tight bounds precisely.

\begin{thm}[Realizable PAC sample complexity {\cite{Hanneke2016,SimonZilles2015}}]
\label{thm:sample-realizable}
Let $\mathcal{H}$ be a hypothesis class with $\VC(\mathcal{H}) = d < \infty$.
In the realizable setting, the optimal sample complexity of PAC learning
$\mathcal{H}$ satisfies
\[
  m(\varepsilon, \delta) = \Theta\!\left(\frac{d}{\varepsilon}
    + \frac{\log(1/\delta)}{\varepsilon}\right).
\]
The upper bound is achieved by any consistent learner (including $\ERM$); the
lower bound holds for every learner.
\end{thm}

\begin{remark}[History of the tight bound]
\label{rmk:sample-history}
The classical upper bound of Blumer et al.~\cite{BlumerEhrenfeuchtHausslerWarmuth1989}
gave $O\!\bigl(\frac{d}{\varepsilon}\log\frac{1}{\varepsilon} +
\frac{1}{\varepsilon}\log\frac{1}{\delta}\bigr)$, with a logarithmic overhead
in the $d/\varepsilon$ term.  Removing this overhead---proving that the correct
dependence on $d$ is linear, not $d \log(1/\varepsilon)$---required substantial
effort.  The matching lower bound was established by Ehrenfeucht et
al.~\cite{EhrenfeuchtHausslerKearnsValiant1989}.  The tight upper bound without
the logarithmic factor was completed by Simon and Zilles~\cite{SimonZilles2015}
and, in a sharper form addressing the constant, by
Hanneke~\cite{Hanneke2016}.  Hanneke's result shows that \emph{any} consistent
learner achieves the optimal bound---no algorithmic cleverness beyond
consistency is needed.
\end{remark}

\begin{proof}[Proof sketch of the upper bound]
The argument proceeds in two stages.  First, one establishes the \emph{double
sampling} bound: the probability that $\ERM$ returns a hypothesis with error
$> \varepsilon$ is at most
\[
  \Pr[\exists h \in \mathcal{H} : \emprisk_S(h) = 0 \text{ and }
    \risk_D(h) > \varepsilon]
  \;\leq\; 2\,\growth_\mathcal{H}(2m) \cdot 2^{-\varepsilon m / 2},
\]
where $\growth_\mathcal{H}(n) \leq (en/d)^d$ by the Sauer--Shelah lemma
(\Cref{lem:sauer-shelah}).  Setting the right-hand side $\leq \delta$ and
solving for $m$ gives $m = O\!\bigl(\frac{d}{\varepsilon}
\log\frac{1}{\varepsilon} + \frac{1}{\varepsilon}\log\frac{1}{\delta}\bigr)$.

The removal of the $\log(1/\varepsilon)$ factor uses a more delicate argument.
One decomposes the hypothesis class into subsets of bounded ``disagreement mass''
and applies uniform convergence separately to each piece.  The key insight
(Hanneke~\cite{Hanneke2016}) is that the number of effectively distinct
hypotheses at accuracy scale $\varepsilon$ is controlled by the \emph{star
number} of $\mathcal{H}$, which is $O(d)$ rather than $O(d \log(1/\varepsilon))$.
\end{proof}

In the agnostic setting, the sample complexity worsens:

\begin{thm}[Agnostic PAC sample complexity]
\label{thm:sample-agnostic}
In the agnostic setting, the optimal sample complexity satisfies
\[
  m(\varepsilon, \delta) = \Theta\!\left(\frac{d}{\varepsilon^2}
    + \frac{\log(1/\delta)}{\varepsilon^2}\right).
\]
\end{thm}

The gap between $d/\varepsilon$ (realizable) and $d/\varepsilon^2$ (agnostic)
is the ``$\varepsilon^2$ price'' discussed in \Cref{sec:agnostic} of
\Cref{ch:pac}.  It is not an artifact of loose analysis: the lower bound
matches.


% ================================================================
% SECTION 2: COMPRESSION SCHEMES
% Profile F (Open Frontier). CENTERPIECE of the chapter.
% ================================================================

\section{Compression Schemes}
\label{sec:compression}

Sample complexity tells us \emph{how much data} learning requires.  Compression
asks a different question: \emph{how little of the data does the learner
actually need to remember?}

\begin{defn}[Sample Compression Scheme {\cite{LittlestoneWarmuth1986,FloydWarmuth1995}}]
\label{def:compression-scheme}
A \emph{sample compression scheme} of size $k$ for a hypothesis class
$\mathcal{H}$ over domain $X$ consists of two maps:
\begin{enumerate}[leftmargin=*, itemsep=2pt]
\item A \emph{compression function} $\kappa$ that, given any sample $S$
  consistent with some $h \in \mathcal{H}$, selects a subsample
  $\kappa(S) \subseteq S$ with $|\kappa(S)| \leq k$.
\item A \emph{reconstruction function} $\rho$ that, given $\kappa(S)$, produces
  a hypothesis $\rho(\kappa(S))$ consistent with $S$.
\end{enumerate}
The reconstruction function $\rho$ sees \emph{only} the compressed subsample
$\kappa(S)$; it has no access to the remaining points.
\end{defn}

\begin{example}[Support vectors as compression]
\label{ex:svm-compression}
A support vector machine in $\R^d$ with a hard margin computes a
hyperplane determined by at most $d+1$ support vectors.  These support vectors
form a compression scheme of size $d+1$: the subsample $\kappa(S)$ consists of
the support vectors, and the reconstruction function $\rho$ returns the
maximum-margin hyperplane through them.
\end{example}

The fundamental connection between compression and learnability was established
by Littlestone and Warmuth:

\begin{thm}[Compression implies learnability {\cite{LittlestoneWarmuth1986,FloydWarmuth1995}}]
\label{thm:compression-pac}
If $\mathcal{H}$ admits a sample compression scheme of size $k$, then
$\mathcal{H}$ is PAC learnable with sample complexity
\[
  m(\varepsilon, \delta) = O\!\left(\frac{k}{\varepsilon}
    \log\frac{k}{\varepsilon} + \frac{1}{\varepsilon}\log\frac{1}{\delta}\right).
\]
\end{thm}

\begin{proof}[Proof sketch]
Fix a sample $S$ of size $m$ drawn i.i.d.\ from $D$.  The compression function
selects a subsample $\kappa(S)$ of size $\leq k$.  The number of possible
subsamples is $\binom{m}{k} \leq m^k$.  For each fixed subsample, the
reconstruction $\rho(\kappa(S))$ is a fixed hypothesis, and by a Hoeffding
bound, a single fixed hypothesis with empirical risk zero on the remaining
$m - k$ points has true risk $> \varepsilon$ with probability at most
$e^{-\varepsilon(m-k)}$.  Taking a union bound over all $m^k$ possible
subsamples:
\[
  \Pr[\risk_D(\rho(\kappa(S))) > \varepsilon] \leq m^k \cdot e^{-\varepsilon(m-k)}.
\]
Setting this $\leq \delta$ and solving for $m$ gives the stated bound.
\end{proof}

The converse direction---does learnability imply compression?---is where the
story becomes a forty-year open problem.

\subsection{The Compression Conjecture}

\begin{openprob}[Sample Compression Conjecture {\cite{LittlestoneWarmuth1986,FloydWarmuth1995}}]
\label{op:compression}
Does every concept class $\mathcal{H}$ with $\VC(\mathcal{H}) = d$ admit a
sample compression scheme of size $O(d)$?
\end{openprob}

This conjecture, stated by Littlestone and Warmuth in 1986, remains open.  It
asks for the converse of \Cref{thm:compression-pac}: if a class is learnable
(which, by the Fundamental Theorem, is equivalent to finite VC dimension $d$),
can the learner always compress its evidence to $O(d)$ points?

Four decades.  The statement is clean, the motivation is immediate, and
every natural proof strategy fails.  Topological arguments reach $2^{O(d)}$
but not $O(d)$.  Algebraic approaches founder on classes with no linear
structure.  The gap between what compression \emph{exists} and what the
conjecture \emph{demands} has not narrowed since the question was posed.
It is not known whether the obstacle is technical or fundamental.

\begin{thm}[Exponential compression exists {\cite{MoranYehudayoff2016}}]
\label{thm:moran-yehudayoff}
Every concept class $\mathcal{H}$ with $\VC(\mathcal{H}) = d$ admits a sample
compression scheme of size $2^{O(d)}$.
\end{thm}

This landmark result of Moran and Yehudayoff~\cite{MoranYehudayoff2016}
established that compression schemes \emph{exist} for every learnable class---a
fact that was itself unknown before 2016.  The exponential bound $2^{O(d)}$ is,
however, vastly larger than the conjectured $O(d)$.

\begin{proof}[Proof sketch]
The proof uses a topological argument.  By a result of
Radon, any set of $d+1$ points in a $d$-dimensional space admits a Radon
partition.  Moran and Yehudayoff generalize this to VC classes via a
``staircase'' construction: they build a sequence of hypothesis classes of
increasing VC dimension and show that each admits a compression scheme, using
the Sauer--Shelah lemma (\Cref{lem:sauer-shelah}) to control the growth of
the compression size.  The exponential blowup arises from iterating this
construction $d$ times.
\end{proof}

\begin{remark}[The gap]
\label{rmk:compression-gap}
The gap between the conjecture ($O(d)$) and the best known upper bound
($2^{O(d)}$) is enormous.  For $d = 20$, the conjecture predicts compression to
$\sim 20$ points; the Moran--Yehudayoff bound gives $\sim 10^6$.  No
polynomial-in-$d$ bound is known for general classes.
\end{remark}

\subsection{The One-Inclusion Graph}
\label{sec:one-inclusion}

The Moran--Yehudayoff proof rests on a combinatorial object that deserves
explicit treatment, both because it illuminates why maximum classes are
special and because it localizes the difficulty of the conjecture at a
precise structural point.

\begin{defn}[One-Inclusion Graph {\cite{HausslerLittlestonWarmuth1994}}]
\label{def:one-inclusion-graph}
Let $\mathcal{H}$ be a concept class over domain $X$, and let
$S = (x_1, \ldots, x_m)$ be a sequence of points from $X$.  The
\emph{one-inclusion graph} $G(\mathcal{H}, S)$ is the graph whose
vertex set is
$\{(h(x_1), \ldots, h(x_m)) : h \in \mathcal{H}\}$
(the distinct restrictions of $\mathcal{H}$ to $S$)
and whose edge set connects pairs of restrictions that differ on
exactly one coordinate.
\end{defn}

The one-inclusion graph mediates between the VC dimension and
compression.  The key property is that its density---specifically, the
minimum over all orientations of the maximum in-degree---controls
whether a compression scheme of size $d$ exists.

\begin{thm}[One-Inclusion Graph Orientation {\cite{RubinsteinKroese2007,KuzminWarmuth2007}}]
\label{thm:one-inclusion-orientation}
Let $\mathcal{H}$ be a \emph{maximum class} of VC dimension $d$
(i.e., $\growth_\mathcal{H}(m) = \sum_{i=0}^d \binom{m}{i}$ for all
$m$).  Then for any sample $S$ of size $m \geq d$, the one-inclusion
graph $G(\mathcal{H}, S)$ can be oriented so that every vertex has
in-degree at most $d$.
\end{thm}

This orientation directly yields a compression scheme of size $d$.
Given a sample $S$ consistent with concept $h^*$, the vertex
$v = h^*|_S$ in the oriented one-inclusion graph has at most $d$
incoming edges.  Each incoming edge corresponds to a sample point
$x_i$ where some neighboring concept $h'$ disagrees with $h^*$.
The compressed subsample $\kappa(S)$ consists of these $\leq d$
``witness points.''  The reconstruction function $\rho$ recovers
$h^*$ from the witness points using the acyclicity of the oriented
graph: the witness points uniquely determine $h^*$ among all
concepts in $\mathcal{H}$ consistent with $S$.

\begin{example}[One-inclusion graph for intervals]
\label{ex:oig-intervals}
Let $\mathcal{H}$ be the class of intervals $[a, b] \subseteq [0, 1]$
(labeling $x$ as $1$ iff $x \in [a,b]$), with $\VC(\mathcal{H}) = 2$.
This is a maximum class.  Given a sorted sample
$S = (x_1 < x_2 < \cdots < x_m)$, each concept restricts to a binary
string of the form $0^i 1^j 0^k$ (a contiguous block of 1s flanked by
0s).  The one-inclusion graph connects patterns that differ in a single
bit, which can only happen at the boundaries of the block: either the
leftmost $1$ becomes $0$ or the rightmost $1$ becomes $0$ (or the
reverse).  The graph can be oriented so that each vertex points toward
its two boundary points.  The compressed subsample consists of the two
boundary points of the interval---exactly the information needed to
reconstruct $[a, b]$ on the sample.

The mechanism is subtle: it is not the \emph{endpoints} of the
interval in $[0,1]$ that matter (which would require real-valued
parameters), but the two \emph{sample points} at the boundary of the
positive region.  The one-inclusion graph identifies these boundary
points combinatorially, without reference to the underlying geometry.
\end{example}

For general (non-maximum) classes, the one-inclusion graph does
\emph{not} admit an orientation with in-degree $\leq d$.  This is
the precise technical point where the conjecture resists.

\begin{remark}[The maximum-to-general gap]
\label{rmk:max-general-gap}
Moran and Yehudayoff's strategy is to reduce the general case to the
maximum case.  Any class $\mathcal{H}$ of VC dimension $d$ can be
``fattened'' to a maximum class $\mathcal{M}$ with
$\VC(\mathcal{M}) = d$ by adding concepts until the Sauer--Shelah
bound is achieved with equality.  The maximum class has compression of
size $d$, and the compression scheme can in principle be pulled back to
$\mathcal{H}$.  However, the fattening introduces auxiliary
concepts---the reconstruction function, given a compressed subsample,
must decide which of many $\mathcal{M}$-consistent concepts actually
belongs to $\mathcal{H}$.  The number of such choices is exponential
in $d$, and encoding this choice inflates the compression size to
$2^{O(d)}$.

Closing the gap to $O(d)$ requires either working directly with
non-maximum classes (for which no orientation theorem is known) or
proving that the reconstruction ambiguity can be resolved with
$O(d)$ additional bits rather than $2^{O(d)}$.
\end{remark}


The conjecture has been verified for specific class families:

\begin{enumerate}[leftmargin=*, itemsep=3pt]
\item \textbf{Maximum classes.}  A class $\mathcal{H}$ with
  $\growth_\mathcal{H}(n) = \sum_{i=0}^d \binom{n}{i}$ for all $n$ (achieving
  the Sauer--Shelah bound with equality) admits a compression scheme of size
  $d$~\cite{FloydWarmuth1995,RubinsteinKroese2007}.

\item \textbf{Intersection-closed classes.}  If $\mathcal{H}$ is closed under
  intersections, compression of size $d$ exists~\cite{Helmbold2018}.

\item \textbf{Classes of VC dimension 1.}  Trivially: a single point suffices.

\item \textbf{Halfspaces in $\R^d$.}  The support vector compression of
  \Cref{ex:svm-compression} gives size $d + 1$.

\item \textbf{Balls, rectangles, and other geometric classes.}  Known
  case-by-case.
\end{enumerate}

The intersection-closed case is worth examining in detail, because its
compression mechanism is fundamentally different from the SVM case and
reveals what structure makes compression ``easy.''

\begin{example}[Compression for intersection-closed classes]
\label{ex:intersection-closed-compression}
Let $\mathcal{H}$ be a concept class closed under intersection: if
$h_1, h_2 \in \mathcal{H}$, then $h_1 \cap h_2 \in \mathcal{H}$
(where $(h_1 \cap h_2)(x) = h_1(x) \wedge h_2(x)$).

Given a sample $S$ consistent with target $h^* \in \mathcal{H}$,
define
$\hat{h} = \bigcap\{h \in \mathcal{H} : h \text{ consistent with } S\}$.
By intersection-closure, $\hat{h} \in \mathcal{H}$, and $\hat{h}$ is
the \emph{unique minimal} concept consistent with~$S$.  The
compression scheme exploits this uniqueness.

Let $P^+ = \{x_i \in S : h^*(x_i) = 1\}$ be the positive examples.
The minimal concept $\hat{h}$ is determined by the constraint that it
must label all points in $P^+$ as positive while being minimal in
$\mathcal{H}$.  A \emph{spanning set} for $P^+$ is a subset
$T \subseteq P^+$ such that
$\bigcap\{h \in \mathcal{H} : h(x) = 1\;\forall x \in T\}
 = \bigcap\{h \in \mathcal{H} : h(x) = 1\;\forall x \in P^+\}$%
---i.e., $T$ imposes the same constraints as the full positive set.
A minimal spanning set has size $\leq d = \VC(\mathcal{H})$, because
the VC dimension bounds the number of independent constraints.  The
compression function selects a minimal spanning set; the
reconstruction function computes
$\hat{h} = \bigcap\{h \in \mathcal{H} : h(x) = 1\;\forall (x, 1) \in
\kappa(S)\}$.

The mechanism is structurally different from SVMs.  Support vector
machines use \emph{geometric} structure (margin maximization) to
select anchor points.  Intersection-closed classes use \emph{lattice}
structure (existence of meets) to guarantee a unique minimum.  Both
achieve compression of size $d$, but via incomparable mathematical
facts.  This incomparability is a symptom of the conjecture's
difficulty: there is no single mechanism that explains compression
across all class families.
\end{example}


\subsection{Why the Conjecture Resists Resolution}
\label{sec:why-hard}

The compression conjecture has resisted attack for four decades.  The
difficulty is not a lack of techniques but a \emph{gap between two
regimes}: techniques that work for structured classes but cannot
generalize, and techniques that work for all classes but cannot achieve
polynomial bounds.

\begin{obstbox}
\textbf{Three approaches and their failure modes.}

\medskip
\emph{Approach 1: Direct construction.}  For each verified class
family (halfspaces, intervals, intersection-closed, maximum classes),
one can construct a compression scheme tailored to the family's
structure.  The constructions exploit distinct properties---geometric
anchoring for SVMs, lattice minimality for intersection-closed classes,
one-inclusion graph orientation for maximum classes.  But these
properties are incomparable: no single structural assumption covers all
families, let alone arbitrary VC classes.  Every direct construction
exploits something that arbitrary classes lack.

\medskip
\emph{Approach 2: One-inclusion graph orientation.}  For maximum
classes, \Cref{thm:one-inclusion-orientation} provides an orientation
with in-degree $d$, directly yielding compression of size $d$.  For
non-maximum classes, no such orientation theorem is known.
Non-maximum one-inclusion graphs can have vertices with degree much
larger than $d$, and no known orientation strategy achieves in-degree
$O(d)$ in general.  The maximum class case is tight, but the
combinatorial extremality that makes it work---achieving the
Sauer--Shelah bound with equality---is exactly what arbitrary classes
lack.

\medskip
\emph{Approach 3: Topological embedding.}  Moran and Yehudayoff's
proof uses a topological argument related to Radon partitions in
convex geometry.  The topological machinery is powerful enough to
handle all VC classes but pays an exponential price: the Radon-type
argument iterates a doubling construction $d$ times, producing
compression of size $2^{O(d)}$.  Improving this requires either a
fundamentally different topological tool or a way to avoid the
iterative blowup---and the iterative structure appears intrinsic to
the proof method, not an artifact of a loose analysis.

\medskip
The conjecture sits at the intersection of three mathematical
territories---combinatorics (VC dimension, growth functions, maximum
classes), topology (Radon partitions, convex position), and
information theory (description length, reconstruction
complexity)---and each territory contributes techniques that
illuminate a piece of the puzzle while leaving the central question
open.  A resolution may require connecting these territories in a way
that current methods do not.
\end{obstbox}


\subsection{Labeled vs.\ Unlabeled Compression}
\label{sec:unlabeled-compression}

In \Cref{def:compression-scheme}, the compressed subsample $\kappa(S)$ retains
both the points and their labels.  One can ask whether labels are necessary.

\begin{defn}[Unlabeled Compression Scheme]
\label{def:unlabeled-compression}
An \emph{unlabeled compression scheme} of size $k$ for $\mathcal{H}$ is a pair
$(\kappa, \rho)$ where $\kappa$ selects a set of $\leq k$ \emph{points}
(without labels) from the sample, and $\rho$ reconstructs a hypothesis
consistent with $S$ from these points alone.
\end{defn}

The natural analogue of the compression conjecture for unlabeled schemes is
false.

\begin{thm}[Unlabeled compression conjecture fails {\cite{MoranYehudayoff2016,CheriayanMoranShpilka2022}}]
\label{thm:unlabeled-fails}
There exist concept classes $\mathcal{H}$ with $\VC(\mathcal{H}) = d$ that do
\emph{not} admit unlabeled compression schemes of size $f(d)$ for any function
$f$ independent of $|\mathcal{H}|$.
\end{thm}

This negative result is sharp: while labeled compression of size $2^{O(d)}$
always exists (\Cref{thm:moran-yehudayoff}), unlabeled compression cannot be
bounded by any function of $d$ alone.  The labels carry essential information
that cannot be recovered from the point locations.

\begin{remark}
The failure of unlabeled compression makes the labeled compression conjecture
(\Cref{op:compression}) more subtle than it first appears: labels are not
merely convenient but \emph{necessary} for compression to work, and the
conjecture concerns the labeled case specifically.
\end{remark}


% ================================================================
% SECTION 3: OCCAM'S RAZOR
% Profile B-light. Occam → PAC and partial converse.
% ================================================================

\section{Occam's Razor}
\label{sec:occam}

Compression schemes are a combinatorial form of Occam's razor: prefer the
hypothesis determined by the fewest data points.  The information-theoretic
form---prefer the hypothesis with the shortest description---was formalized by
Blumer, Ehrenfeucht, Haussler, and Warmuth.

\begin{defn}[Occam Algorithm {\cite{BlumerEhrenfeuchtHausslerWarmuth1987}}]
\label{def:occam}
Let $\mathcal{H}$ be a hypothesis class and let $\mathcal{R}$ be a
\emph{representation class} (a set of hypotheses with associated description
lengths).  An algorithm $A$ is an \emph{Occam algorithm} for $\mathcal{H}$ with
respect to $\mathcal{R}$ if, given a sample $S$ of size $m$ consistent with some
$h \in \mathcal{H}$ of description length $n$, the algorithm outputs a
hypothesis $h' \in \mathcal{R}$ that:
\begin{enumerate}[leftmargin=*, nosep]
\item is consistent with $S$, and
\item has description length $|h'| \leq n^\alpha m^\beta$ for some constants
  $\alpha \geq 1$ and $0 \leq \beta < 1$.
\end{enumerate}
The sub-linear dependence on $m$ (exponent $\beta < 1$) is the key requirement:
the output hypothesis must be ``simpler'' than the data.
\end{defn}

\begin{thm}[Occam's Razor Theorem {\cite{BlumerEhrenfeuchtHausslerWarmuth1987}}]
\label{thm:occam}
If $A$ is an Occam algorithm for $\mathcal{H}$ with parameters $\alpha, \beta$,
then $A$ PAC-learns $\mathcal{H}$ with sample complexity
\[
  m(\varepsilon, \delta) = O\!\left(\frac{1}{\varepsilon}\left(
    n^\alpha \log\frac{1}{\varepsilon}
    + \log\frac{1}{\delta}\right)\right)^{1/(1-\beta)}.
\]
\end{thm}

\begin{proof}[Proof sketch]
The argument is a counting and union bound.  The number of hypotheses in
$\mathcal{R}$ with description length $\leq L$ is at most $2^L$.  An Occam
algorithm outputs a hypothesis of length $\leq n^\alpha m^\beta$.  Fixing a
sample of size $m$, the number of possible outputs is at most
$2^{n^\alpha m^\beta}$.  For each fixed hypothesis with empirical risk zero, the
probability that its true risk exceeds $\varepsilon$ is at most
$(1-\varepsilon)^m \leq e^{-\varepsilon m}$.  A union bound gives
\[
  \Pr[\risk_D(A(S)) > \varepsilon] \leq 2^{n^\alpha m^\beta} \cdot e^{-\varepsilon m}.
\]
Setting the right-hand side $\leq \delta$ and solving for $m$ (using $\beta < 1$
to ensure that the exponential decay in $m$ dominates the polynomial growth of
$2^{m^\beta}$) gives the stated bound.
\end{proof}

\begin{remark}[Occam and compression]
\label{rmk:occam-compression}
Every compression scheme of size $k$ induces an Occam algorithm: the compressed
subsample has description length $O(k \log m)$, which is sub-linear in $m$.
Conversely, an Occam algorithm is not necessarily a compression scheme, because
the output hypothesis need not be determined by a subsample of the input.
Occam's razor is thus a relaxation of compression.
\end{remark}

The converse of the Occam theorem---does every PAC-learnable class admit an
Occam algorithm?---was partially answered by Board and Pitt:

\begin{thm}[Partial converse {\cite{BoardPitt1992}}]
\label{thm:board-pitt}
If the class of all polynomial-size circuits is not PAC learnable by
polynomial-size circuits (a widely believed complexity-theoretic assumption),
then there exist PAC-learnable classes that do not admit polynomial-time Occam
algorithms.
\end{thm}

Thus the Occam property is \emph{strictly stronger} than PAC learnability
(under standard complexity assumptions): there exist learnable classes where
no efficient algorithm can find short descriptions of the data.  The gap is
computational, not information-theoretic.


% ================================================================
% SECTION 4: INFORMATION-THEORETIC FOUNDATIONS
% Profile A and A-plus. Brief definitions.
% ================================================================

\section{Information-Theoretic Foundations}
\label{sec:info-foundations}

The notions of ``description length'' and ``covering'' that underlie Occam's
razor and compression have precise information-theoretic formulations.

\subsection{Description Length}

\begin{defn}[Description Length]
\label{def:description-length}
The \emph{description length} of a hypothesis $h$ in a representation class
$\mathcal{R}$ is the length $|h|$ of the shortest binary string encoding $h$ in
$\mathcal{R}$.  A \emph{prefix-free} representation ensures that
$\sum_{h \in \mathcal{R}} 2^{-|h|} \leq 1$ (the Kraft inequality).
\end{defn}

Description length connects to PAC learning through the observation that a
class of hypotheses with bounded description lengths has bounded effective size:
$|\{h \in \mathcal{R} : |h| \leq L\}| \leq 2^L$.  This makes union bounds
over the class tractable.

\subsection{Kolmogorov Complexity}

\begin{defn}[Kolmogorov Complexity]
\label{def:kolmogorov}
The \emph{Kolmogorov complexity} $K(x)$ of a string $x$ is the length of the
shortest program (on a fixed universal Turing machine) that outputs $x$ and
halts.  The \emph{conditional Kolmogorov complexity} $K(x \mid y)$ is the length
of the shortest program that outputs $x$ given $y$ as input.
\end{defn}

Kolmogorov complexity is uncomputable but provides the theoretical gold standard
for description length: $K(x)$ is the ultimate compression of $x$.

\subsection{KL Complexity}

\begin{defn}[KL Complexity]
\label{def:kl-complexity}
The \emph{Kullback--Leibler divergence} (KL divergence) between distributions
$P$ and $Q$ on a measurable space is
\[
  \KL(P \| Q) = \E_P\!\left[\log\frac{dP}{dQ}\right],
\]
when $P \ll Q$, and $+\infty$ otherwise.  In the PAC-Bayes framework
(\Cref{ch:generalization}), $\KL(\rho \| \pi)$ measures the ``complexity'' of a
posterior $\rho$ relative to a prior $\pi$: it is the number of additional
nats needed to encode a hypothesis drawn from $\rho$ using a code designed for
$\pi$.
\end{defn}

\subsection{Covering Numbers}

\begin{defn}[Covering Number]
\label{def:covering-number}
Let $(\mathcal{F}, d)$ be a (pseudo)metric space.  The \emph{$\varepsilon$-covering
number} $\mathcal{N}(\varepsilon, \mathcal{F}, d)$ is the minimum number of
balls of radius $\varepsilon$ (under $d$) needed to cover $\mathcal{F}$.
\end{defn}

Covering numbers measure the ``effective size'' of a function class at
resolution $\varepsilon$.  They connect to sample complexity through the
classical chaining argument:

\begin{thm}[Dudley's entropy integral {\cite{Dudley1967}}]
\label{thm:dudley}
Let $\mathcal{F}$ be a class of functions $f \colon X \to [0,1]$, and let
$d_n(f,g) = \bigl(\frac{1}{n}\sum_{i=1}^n (f(x_i) - g(x_i))^2\bigr)^{1/2}$
be the empirical $L^2$ metric on a sample $(x_1, \ldots, x_n)$.  Then
\[
  \E\!\left[\sup_{f \in \mathcal{F}} \left|\frac{1}{n}\sum_{i=1}^n f(x_i)
    - \E[f]\right|\right]
  \leq \frac{C}{\sqrt{n}} \int_0^1
    \sqrt{\log \mathcal{N}(\varepsilon, \mathcal{F}, d_n)}\; d\varepsilon,
\]
where $C$ is a universal constant.
\end{thm}

The integral $\int_0^1 \sqrt{\log \mathcal{N}(\varepsilon, \mathcal{F}, d_n)}\;
d\varepsilon$ is the \emph{Dudley entropy integral}.  It bounds the expected
supremum of the empirical process indexed by $\mathcal{F}$---the quantity that
controls uniform convergence and hence generalization.  When
$\mathcal{N}(\varepsilon, \mathcal{F}, d_n) \leq (C/\varepsilon)^d$ (as for
VC classes), the integral evaluates to $O(\sqrt{d})$, recovering the standard
sample complexity bounds.


% ================================================================
% SECTION 5: CHAPTER SUMMARY AND OPEN QUESTIONS
% ================================================================

\section{The State of the Art}
\label{sec:compression-summary}

The landscape of this chapter is dominated by the compression conjecture.
\Cref{fig:compression-landscape} summarizes the logical dependencies.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
    node distance=1.5cm and 2.5cm,
    box/.style={draw, thick, rounded corners, minimum width=3cm,
      minimum height=0.8cm, font=\small, align=center},
    open/.style={box, dashed, fill=edgeorange!10, text=edgeorange!80!black},
    known/.style={box, fill=defblue!10},
    negative/.style={box, fill=thmred!10},
    arr/.style={-{Stealth[length=2.5mm]}, thick},
    lbl/.style={font=\scriptsize, midway, fill=white, inner sep=1pt}
]
% Nodes
\node[known] (vc) {$\VC(\mathcal{H}) = d < \infty$};
\node[known, below left=1.5cm and 0.5cm of vc] (pac) {PAC learnable};
\node[known, below right=1.5cm and 0.5cm of vc] (uc) {Uniform convergence};
\node[known, below=3cm of vc] (compress-exp) {Compression\\size $2^{O(d)}$};
\node[open, below=1.5cm of compress-exp] (compress-d) {Compression\\size $O(d)$\;?};
\node[known, right=2.5cm of compress-exp] (occam) {Occam algorithm};
\node[negative, left=2.5cm of compress-d] (unlabeled) {Unlabeled comp.\\size $f(d)$: \textbf{FALSE}};

% Edges
\draw[arr] (vc) -- node[lbl, left] {\Cref{ch:pac}} (pac);
\draw[arr] (vc) -- node[lbl, right] {\Cref{ch:pac}} (uc);
\draw[arr] (vc) -- node[lbl, right] {\cite{MoranYehudayoff2016}} (compress-exp);
\draw[arr, dashed, edgeorange] (compress-exp) -- node[lbl] {Open!} (compress-d);
\draw[arr] (compress-exp) -- node[lbl, above] {implies} (occam);
\draw[arr] (occam) -- node[lbl, right] {\Cref{thm:occam}} (pac);
\draw[arr, thmred, dashed] (compress-d) -- node[lbl, above] {fails} (unlabeled);
\end{tikzpicture}
\caption{Logical landscape of compression and learnability.  Solid arrows
  denote known implications; dashed arrows denote open or negative results.
  The central open question is whether compression of size $O(d)$ is
  achievable.}
\label{fig:compression-landscape}
\end{figure}

The following table summarizes the best known compression bounds:

\begin{table}[ht]
\centering\small
\caption{Known compression bounds by class family.}
\label{tab:compression-bounds}
\begin{tabular}{lcc}
\toprule
\textbf{Class family} & \textbf{VC dim.} & \textbf{Compression size} \\
\midrule
Maximum classes & $d$ & $d$ \\
Intersection-closed & $d$ & $d$ \\
Halfspaces in $\R^d$ & $d+1$ & $d+1$ \\
VC dim.\ 1 & 1 & 1 \\
Arbitrary VC class & $d$ & $2^{O(d)}$ \\
Unlabeled (arbitrary) & $d$ & No $f(d)$ bound \\
\bottomrule
\end{tabular}
\end{table}

\begin{openprob}[Polynomial compression]
\label{op:poly-compression}
Does every concept class of VC dimension $d$ admit a compression scheme of
size $\mathrm{poly}(d)$?  Even this weaker form of the compression conjecture
remains open.
\end{openprob}

The chapter's three main themes---tight sample complexity, compression, and
Occam's razor---are unified by a single idea: \emph{learnability is equivalent
to the existence of a short summary of the evidence}.  The VC dimension tells
us \emph{that} a short summary exists; the compression conjecture asks
\emph{how short}.  The answer, after forty years, remains: we do not know.


% ================================================================
% EXERCISES
% ================================================================

\subsection*{Exercises}

\begin{enumerate}[leftmargin=*,itemsep=6pt]

\item \textbf{Compression lower bound.}
  Let $\mathcal{H}$ be a concept class with $\VC(\mathcal{H}) = d$, and
  let $(\kappa, \rho)$ be a sample compression scheme of size $k$ for
  $\mathcal{H}$.  Prove that $k \geq d$.

  \emph{Hint:} Let $C = \{x_1, \ldots, x_d\}$ be a shattered set.
  For each of the $2^d$ labelings $b \in \{0,1\}^d$, the compression
  function selects a subsequence $\kappa(S_b) \subseteq S_b$ of size
  $\leq k$.  Show that if $k < d$, a pigeonhole argument forces two
  distinct labelings $b \neq b'$ to compress to the same subsequence
  $\kappa(S_b) = \kappa(S_{b'})$ with the same labels, contradicting
  the requirement that reconstruction produces hypotheses consistent
  with both $S_b$ and $S_{b'}$.

\item \textbf{Compression for unions of intervals.}
  Let $\mathcal{H}_k$ be the class of unions of at most $k$ disjoint
  intervals on $[0,1]$: $h \in \mathcal{H}_k$ iff $h^{-1}(1)$ is a
  union of at most $k$ intervals.
  \begin{enumerate}[label=(\alph*)]
  \item Prove that $\VC(\mathcal{H}_k) = 2k$.
  \item Construct an explicit compression scheme of size $2k$ for
    $\mathcal{H}_k$.  (\emph{Hint:} Generalize
    \Cref{ex:oig-intervals}; the compressed subsample consists of
    the $2k$ boundary points of the positive regions.)
  \item Use the result of Exercise~1 to conclude that no compression
    scheme of size $< 2k$ exists.  This verifies the compression
    conjecture for $\mathcal{H}_k$ with equality.
  \end{enumerate}

\item \textbf{Compression and teaching dimension.}
  The \emph{teaching dimension} $\mathrm{TD}(\mathcal{H})$ is the
  minimum $k$ such that every concept $h \in \mathcal{H}$ has a
  \emph{teaching set} of size $k$: a set $T \subseteq X$ with
  $|T| = k$ such that $h$ is the unique concept in $\mathcal{H}$
  consistent with $h|_T$.
  \begin{enumerate}[label=(\alph*)]
  \item Show that if $\mathcal{H}$ has teaching dimension $k$, then
    $\mathcal{H}$ admits a compression scheme of size $k$.  (The
    teaching set is the compressed subsample.)
  \item Show that a compression scheme of size $k$ does \emph{not}
    imply teaching dimension $\leq k$: construct a class $\mathcal{H}$
    with a compression scheme of size $1$ but teaching dimension
    $> 1$.  (\emph{Hint:} The compression function may depend on the
    full sample, not just on the target concept; the teaching set must
    work for a single concept without knowing the other data.)
  \item Prove that $\mathrm{TD}(\mathcal{H}) \geq \VC(\mathcal{H})$
    for all classes $\mathcal{H}$.  Combined with the compression
    conjecture, this would imply
    $\VC \leq \text{compression size} \leq \mathrm{TD}$: compression
    sits between VC dimension and teaching dimension.  But whether
    the compression conjecture implies
    $\text{compression size} \leq O(\mathrm{TD})$ or
    $\text{compression size} \leq O(\VC)$ remains open---these are
    distinct (and incomparable) questions.
  \end{enumerate}

\end{enumerate}
