% Chapter 12: Generalization Bounds
%   uniform_convergence: Profile B (load-bearing theorem --- symmetrization
%     + concentration proof IS the content)
%   Five bound traditions: Profile G (framework bridge --- parallax)
%   rademacher_complexity: Profile A-to-B bridge (definition matters, but
%     comparison to VC is the content)
%   pac_bayes_bound: Profile E (obstruction --- looks Bayesian but prior
%     has a fundamentally different role)
%
% Centerpiece: the symmetrization argument (Sec 2) and the five-framework
% parallax (Sec 8).

\chapter{Generalization Bounds}
\label{ch:generalization}

\Cref{ch:pac} established \emph{that} finite VC dimension characterizes PAC
learnability.  \Cref{ch:compression} refined the quantitative dependence
through compression.  This chapter asks a different question: given a
specific learning algorithm~$A$ and a specific training sample~$S$, what
can we say about the generalization error of the hypothesis~$A(S)$?

The answer is not unique.  Five distinct research traditions have developed
bounds on the same quantity---the gap between true and empirical
risk---each using different information about the learner:

\begin{enumerate}[leftmargin=*, itemsep=3pt]
\item \textbf{VC/uniform convergence} measures the combinatorial richness
  of the hypothesis class through the growth function.
\item \textbf{Rademacher complexity} refines this to a data-dependent
  measure of how well the class correlates with random noise.
\item \textbf{PAC-Bayes bounds} measure the ``distance'' of a learned
  posterior from a data-independent prior.
\item \textbf{Algorithmic stability} measures how much the output changes
  when a single training example is perturbed.
\item \textbf{Margin theory} measures the ``confidence'' of predictions
  and connects to scale-sensitive dimensions.
\end{enumerate}

\noindent
We also discuss information-theoretic bounds, which measure the mutual
information between the training sample and the learned hypothesis.

Each framework produces a bound on the same quantity but requires
different structural assumptions and succeeds in different regimes.  The
centerpiece of this chapter is twofold: (i)~the \emph{symmetrization
argument} that powers uniform convergence (\Cref{sec:uniform-convergence}),
proved in full as the engine behind classical generalization theory; and
(ii)~the \emph{comparison} among the five frameworks
(\Cref{sec:comparison}), which reveals what each sees that the others miss.

\medskip

Throughout this chapter, $\mathcal{H}$ denotes a hypothesis class,
$D$~a distribution on $X \times Y$,
$S = \{(x_i, y_i)\}_{i=1}^m$ an i.i.d.\ sample from~$D$,
$A$~a learning algorithm, and
$\ell \colon \mathcal{H} \times (X \times Y) \to [0,1]$ a bounded loss
function.


% ================================================================
% SECTION 1: GENERALIZATION ERROR
% Profile A.  Definition.
% ================================================================

\section{Generalization Error}
\label{sec:gen-error}

\begin{defn}[Generalization Error]
\label{def:gen-error}
The \emph{generalization error} (or \emph{generalization gap}) of a
hypothesis $h$ with respect to distribution~$D$ and sample~$S$ is
\[
  \mathrm{gen}(h, S) \;=\; \risk_D(h) - \emprisk_S(h)
    \;=\; \E_{(x,y) \sim D}[\ell(h, (x,y))]
    - \frac{1}{m}\sum_{i=1}^m \ell(h, (x_i, y_i)).
\]
A \emph{generalization bound} is any high-probability or in-expectation
upper bound on $|\mathrm{gen}(h, S)|$ or on $\mathrm{gen}(A(S), S)$.
\end{defn}

The generalization error is a random variable (through~$S$), and when
$h = A(S)$ depends on~$S$, controlling it requires accounting for the
dependence between the hypothesis and the data that produced it.  This is
the core technical challenge that each framework addresses differently.


% ================================================================
% SECTION 2: UNIFORM CONVERGENCE
% Profile B (load-bearing theorem).  FULL proof.
% ================================================================

\section{Uniform Convergence: The Engine}
\label{sec:uniform-convergence}

The classical approach to generalization does not look at the algorithm at
all.  Instead, it asks: does the empirical risk converge to the true risk
\emph{uniformly over all hypotheses}?  If so, any hypothesis with small
empirical risk automatically has small true risk---including the one the
algorithm selects.

\begin{defn}[Uniform Convergence]
\label{def:uniform-convergence}
A hypothesis class~$\mathcal{H}$ has the \emph{uniform convergence
property} with respect to loss~$\ell$ if, for every $\varepsilon > 0$,
\[
  \sup_{h \in \mathcal{H}} |\risk_D(h) - \emprisk_S(h)|
  \xrightarrow{\;m \to \infty\;} 0
  \quad \text{in probability, uniformly over all distributions } D.
\]
\end{defn}

The content of uniform convergence is the proof technique: the
\emph{symmetrization argument} introduced by Vapnik and Chervonenkis in
1971.  The argument proceeds in three stages, each with a distinct role.

\begin{thm}[VC Uniform Convergence Bound]
\label{thm:vc-uniform}
Let $\mathcal{H} \subseteq \{0,1\}^X$ with $\VC(\mathcal{H}) = d < \infty$.
Then for any distribution~$D$ on $X \times \{0,1\}$, any $\varepsilon > 0$,
and any $\delta > 0$, with probability at least $1 - \delta$ over
$S \sim D^m$:
\[
  \sup_{h \in \mathcal{H}} |\risk_D(h) - \emprisk_S(h)|
  \;\leq\; \sqrt{\frac{8\,d \log(2em/d) + 8\log(4/\delta)}{m}}.
\]
\end{thm}

\begin{proof}
The proof has three stages: symmetrization, reduction to a finite
problem, and concentration.

\textbf{Stage 1: Symmetrization (the ghost sample).}
Let $S = (z_1, \ldots, z_m)$ be the training sample and let
$S' = (z_1', \ldots, z_m')$ be an independent ``ghost'' sample drawn
from the same distribution.  Define
$\Phi(S) = \sup_{h \in \mathcal{H}} (\risk_D(h) - \emprisk_S(h))$.

For any fixed~$h$, the true risk equals
$\risk_D(h) = \E_{S'}[\emprisk_{S'}(h)]$, so
\[
  \Phi(S) = \sup_{h \in \mathcal{H}} \E_{S'}\!\left[
    \emprisk_{S'}(h) - \emprisk_S(h)\right].
\]
Moving the supremum inside the expectation (Jensen's inequality applied
to the convex functional $\sup$) gives:
\begin{equation}
\label{eq:symm-step}
  \E_S[\Phi(S)]
  \;\leq\; \E_{S,S'}\!\left[\sup_{h \in \mathcal{H}}
    \bigl(\emprisk_{S'}(h) - \emprisk_S(h)\bigr)\right].
\end{equation}

\emph{Why this helps.}  The left side involves the population
quantity~$\risk_D(h)$, which we cannot compute.  The right side involves
only \emph{two finite samples}---no population quantities remain.  This is
the symmetrization trick: the ghost sample replaces the unknown
distribution.

\textbf{Stage 1b: Rademacher sign-flip.}
The difference $\emprisk_{S'}(h) - \emprisk_S(h) =
\frac{1}{m}\sum_{i=1}^m (\ell(h,z_i') - \ell(h,z_i))$ is a sum of
symmetric random variables (since $z_i$ and $z_i'$ are identically
distributed, swapping them does not change the joint distribution).
Introducing i.i.d.\ Rademacher variables $\sigma_1,\ldots,\sigma_m$
(uniform on $\{-1,+1\}$):
\begin{align}
  \E_{S,S'}\!\left[\sup_{h} \frac{1}{m}\sum_{i=1}^m
    \bigl(\ell(h,z_i') - \ell(h,z_i)\bigr)\right]
  &= \E_{S,S',\sigma}\!\left[\sup_{h} \frac{1}{m}\sum_{i=1}^m
    \sigma_i\bigl(\ell(h,z_i') - \ell(h,z_i)\bigr)\right]
    \nonumber \\
  &\leq 2\,\E_{S,\sigma}\!\left[\sup_{h} \frac{1}{m}\sum_{i=1}^m
    \sigma_i \ell(h,z_i)\right],
    \label{eq:rademacher-step}
\end{align}
where the last step uses the triangle inequality for suprema and the
identical distribution of $S$ and~$S'$.

\textbf{Stage 2: Reduction to a finite class.}
The supremum in \eqref{eq:rademacher-step} ranges over all
$h \in \mathcal{H}$, but on the fixed sample~$S$, only finitely many
distinct behavior patterns occur.  Specifically, the loss vectors
$(\ell(h,z_1), \ldots, \ell(h,z_m))$ take at most $\growth_\mathcal{H}(m)$
distinct values, where $\growth_\mathcal{H}$ is the growth function.

For binary classification with 0--1 loss, each behavior pattern is a
binary vector in $\{0,1\}^m$, and the Sauer--Shelah lemma
(\Cref{ch:dimensions}) gives
\begin{equation}
\label{eq:sauer-shelah-app}
  \growth_\mathcal{H}(m) \;\leq\; \left(\frac{em}{d}\right)^d.
\end{equation}

Applying Massart's finite lemma to the at most $\growth_\mathcal{H}(m)$
distinct vectors, each with Euclidean norm at most $\sqrt{m}$:
\begin{equation}
\label{eq:massart}
  \E_\sigma\!\left[\sup_{h} \frac{1}{m}\sum_{i=1}^m \sigma_i \ell(h,z_i)
  \right]
  \;\leq\; \sqrt{\frac{2\log \growth_\mathcal{H}(m)}{m}}
  \;\leq\; \sqrt{\frac{2d\log(em/d)}{m}}.
\end{equation}

\textbf{Stage 3: Concentration.}
The bound so far controls $\E[\Phi(S)]$.  To obtain a high-probability
bound, observe that $\Phi(S) = \sup_h (\risk_D(h) - \emprisk_S(h))$
satisfies the bounded-differences condition: replacing any single~$z_i$
changes $\Phi(S)$ by at most $1/m$ (since each loss value lies in
$[0,1]$).  By McDiarmid's inequality,
\[
  \Pr\!\left[\Phi(S) > \E[\Phi(S)] + t\right]
  \leq \exp(-2mt^2).
\]
Setting $t = \sqrt{\log(2/\delta)/(2m)}$ and combining with
\eqref{eq:rademacher-step}--\eqref{eq:massart}, the one-sided bound
follows.  Applying the same argument to
$\sup_h (\emprisk_S(h) - \risk_D(h))$ and taking a union bound yields
the two-sided result.
\end{proof}

\begin{historical}
The symmetrization argument was introduced by Vapnik and Chervonenkis
(1971) in the original paper that defined what is now called the VC
dimension.  The three-stage structure---ghost sample, sign-flip,
finite reduction---was not initially presented in this clean form.  The
modern presentation, which passes through Rademacher complexity as an
intermediate quantity, crystallized in the work of Koltchinskii (2001)
and Bartlett--Mendelson (2002).  The Rademacher step (Stage~1b) was
implicit in the original argument but was not isolated as a self-standing
concept until much later.
\end{historical}

\begin{traversal}
The uniform convergence theorem connects three graph nodes:
$\edge{growth\_function}{characterizes}{uniform\_convergence}$,
$\edge{sauer\_shelah\_lemma}{used\_in\_proof}{vc\_generalization\_bound}$,
and $\edge{vc\_dimension}{upper\_bounds}{rademacher\_complexity}$.
The proof above traces this path: the growth function enters at Stage~2,
Sauer--Shelah provides the polynomial bound, and the Rademacher
intermediate appears at Stage~1b.
\end{traversal}


\begin{example}[Halfspaces in $\R^d$]
\label{ex:halfspace-uc}
Let $\mathcal{H} = \{x \mapsto \ind[\langle w, x \rangle \geq b]
: w \in \R^d, b \in \R\}$ be the class of halfspaces.
Then $\VC(\mathcal{H}) = d+1$ (\Cref{ch:dimensions}), and the uniform
convergence bound gives: with probability $\geq 1 - \delta$,
\[
  \sup_{h \in \mathcal{H}} |\risk_D(h) - \emprisk_S(h)|
  \leq \sqrt{\frac{8(d+1)\log(2em/(d+1)) + 8\log(4/\delta)}{m}}.
\]
For $d = 10$ and $m = 10{,}000$, this gives a bound of approximately
$0.12$ --- nontrivial and meaningful.  For a two-layer neural network
with $p = 10^6$ parameters (where naive VC bounds give
$\VC \approx O(p \log p)$), the same formula gives a bound exceeding~$1$:
the uniform convergence bound is \emph{vacuous}.
\end{example}


% ================================================================
% SECTION 3: RADEMACHER COMPLEXITY
% Profile A-to-B bridge.  Definition + comparison to VC.
% ================================================================

\section{Rademacher Complexity}
\label{sec:rademacher}

The Rademacher intermediate that appeared in Stage~1b of the
symmetrization proof is itself a fundamental complexity measure.  It
refines VC dimension in two ways: it is \emph{data-dependent} (computed on
the actual sample, not worst-case) and it applies to \emph{real-valued}
function classes (not just binary hypotheses).

\begin{defn}[Empirical Rademacher Complexity]
\label{def:rademacher}
Let $\mathcal{F}$ be a class of functions $f \colon Z \to \R$ and let
$S = (z_1, \ldots, z_m)$ be a fixed sample.  The \emph{empirical
Rademacher complexity} of $\mathcal{F}$ on~$S$ is
\[
  \Rad_S(\mathcal{F}) = \E_{\sigma}\!\left[\sup_{f \in \mathcal{F}}
    \frac{1}{m}\sum_{i=1}^m \sigma_i f(z_i)\right],
\]
where $\sigma_1, \ldots, \sigma_m$ are independent Rademacher random
variables (uniform on $\{-1, +1\}$).  The \emph{Rademacher complexity} of
$\mathcal{F}$ at sample size~$m$ is
$\mathcal{R}_m(\mathcal{F}) = \E_S[\Rad_S(\mathcal{F})]$.
\end{defn}

The Rademacher complexity measures how well $\mathcal{F}$ can
\emph{correlate with pure noise}.  A class that achieves high correlation
with random $\pm 1$ labels is rich enough to fit arbitrary patterns,
including the noise in the training data.

\begin{thm}[Rademacher Generalization Bound
{\cite{BartlettMendelson2002}}]
\label{thm:rademacher-bound}
Let $\mathcal{F}$ be a class of functions $f \colon Z \to [0,1]$, let
$D$ be a distribution on~$Z$, and let $S = (z_1,\ldots,z_m) \sim D^m$.
Then for any $\delta > 0$, with probability at least $1-\delta$ over~$S$,
for all $f \in \mathcal{F}$ simultaneously:
\[
  \E_D[f] \leq \frac{1}{m}\sum_{i=1}^m f(z_i)
    + 2\,\Rad_S(\mathcal{F})
    + 3\sqrt{\frac{\log(2/\delta)}{2m}}.
\]
\end{thm}

\begin{proof}
The proof follows the same three-stage architecture as the uniform
convergence theorem: symmetrization, Rademacher sign-flip, and
concentration.

Define $\Phi(S) = \sup_{f \in \mathcal{F}} (\E_D[f] -
\frac{1}{m}\sum_i f(z_i))$.

\textbf{Step 1: Concentration of $\Phi$.}
Replacing any single $z_i$ in~$S$ changes $\Phi(S)$ by at most $1/m$
(since $f$ is bounded in $[0,1]$).  By McDiarmid's inequality:
\begin{equation}
\label{eq:rad-conc}
  \Pr[\Phi(S) > \E[\Phi(S)] + t] \leq \exp(-2mt^2).
\end{equation}
Setting $t = \sqrt{\log(2/\delta)/(2m)}$ gives
$\Phi(S) \leq \E[\Phi(S)] + \sqrt{\log(2/\delta)/(2m)}$
with probability $\geq 1 - \delta/2$.

\textbf{Step 2: Symmetrization.}
Let $S' = (z_1', \ldots, z_m')$ be an independent ghost sample.  Then
\begin{align*}
  \E_S[\Phi(S)]
  &= \E_S\!\left[\sup_{f \in \mathcal{F}} \left(\E_{S'}
    \!\left[\frac{1}{m}\sum_{i=1}^m f(z_i')\right]
    - \frac{1}{m}\sum_{i=1}^m f(z_i)\right)\right] \\
  &\leq \E_{S, S'}\!\left[\sup_{f \in \mathcal{F}}
    \frac{1}{m}\sum_{i=1}^m \bigl(f(z_i') - f(z_i)\bigr)\right].
\end{align*}
Since $z_i$ and $z_i'$ are i.i.d., the differences are symmetric
random variables, so introducing Rademacher signs does not change the
distribution:
\begin{align*}
  \E_{S, S'}\!\left[\sup_{f}
    \frac{1}{m}\sum_i \bigl(f(z_i') - f(z_i)\bigr)\right]
  &= \E_{S, S', \sigma}\!\left[\sup_{f}
    \frac{1}{m}\sum_i \sigma_i\bigl(f(z_i') - f(z_i)\bigr)\right] \\
  &\leq 2\,\E_{S, \sigma}\!\left[\sup_{f}
    \frac{1}{m}\sum_i \sigma_i f(z_i)\right]
  = 2\,\mathcal{R}_m(\mathcal{F}).
\end{align*}

\textbf{Step 3: Data-dependent form.}
Apply McDiarmid again: $\Rad_S(\mathcal{F})$ has bounded differences
$1/m$ (changing one sample point changes the Rademacher average by at
most $1/m$), so
$\mathcal{R}_m(\mathcal{F}) \leq \Rad_S(\mathcal{F})
  + \sqrt{\log(2/\delta)/(2m)}$
with probability $\geq 1 - \delta/2$.

A union bound over the two $\delta/2$ events yields the stated bound.
\end{proof}


\subsection{Rademacher Complexity versus VC Dimension}
\label{sec:rad-vs-vc}

The following proposition makes the relationship precise.

\begin{prop}[VC dimension bounds Rademacher complexity]
\label{prop:rad-vc}
For any binary hypothesis class $\mathcal{H} \subseteq \{0,1\}^X$ with
$\VC(\mathcal{H}) = d$:
\[
  \mathcal{R}_m(\mathcal{H})
  \;\leq\; \sqrt{\frac{2d \log(em/d)}{m}}.
\]
\end{prop}

\begin{proof}
On any fixed sample of size~$m$, the restriction $\mathcal{H}|_S$ has at
most $\growth_\mathcal{H}(m) \leq (em/d)^d$ elements (Sauer--Shelah).
Each element is a binary vector of Euclidean norm at most $\sqrt{m}$.
By Massart's finite lemma, the Rademacher complexity of a finite set~$V$
of vectors in $\R^m$ satisfies
$\E_\sigma[\sup_{v \in V} \frac{1}{m}\langle \sigma, v\rangle]
\leq \frac{\max_v \|v\|_2}{m}\sqrt{2\log|V|}$.
Substituting $|V| \leq (em/d)^d$ and $\|v\|_2 \leq \sqrt{m}$ gives
the result.
\end{proof}

\begin{remark}[When Rademacher is tighter than VC]
\label{rmk:rad-tighter}
The VC bound is worst-case: it uses the maximum of
$|\mathcal{H}|_S|$ over all samples of size~$m$.  The empirical
Rademacher complexity $\Rad_S(\mathcal{H})$ uses the \emph{actual}
restriction $\mathcal{H}|_S$, which may be much smaller.  Consider the
class of all threshold functions on $\R$:
$\mathcal{H} = \{x \mapsto \ind[x \geq \theta] : \theta \in \R\}$.
This class has $\VC(\mathcal{H}) = 1$ and
$\growth_\mathcal{H}(m) = m + 1$, so the VC bound gives
$\mathcal{R}_m \leq \sqrt{2\log(em)/m}$.  But on a sample where all
points lie in $\{0,1\}$, the restriction has at most $3$ elements, and the
empirical Rademacher complexity is $O(1/\sqrt{m})$ without the
logarithmic factor.  The gain is modest here but becomes significant for
classes with distribution-dependent structure.
\end{remark}

\begin{computational}
\textbf{Rademacher complexity of linear classifiers.}
Let $\mathcal{F} = \{x \mapsto \langle w, x \rangle : \|w\|_2 \leq W\}$
and suppose $\|x_i\|_2 \leq B$ for all $i$.  Then
\[
  \Rad_S(\mathcal{F})
  = \E_\sigma\!\left[\sup_{\|w\| \leq W}
    \frac{1}{m}\left\langle w, \sum_i \sigma_i x_i\right\rangle\right]
  = \frac{W}{m}\,\E_\sigma\!\left[\left\|\sum_i \sigma_i x_i\right\|_2
    \right]
  \leq \frac{WB}{\sqrt{m}},
\]
where the last step uses
$\E[\|\sum_i \sigma_i x_i\|_2^2] = \sum_i \|x_i\|_2^2 \leq mB^2$ and
Jensen's inequality.  Note: no dependence on ambient dimension.  A
linear classifier in $\R^{10^6}$ with bounded norm has the same
Rademacher complexity as one in $\R^{10}$---only the norm and the data
radius matter.
\end{computational}


% ================================================================
% SECTION 4: THE GROWTH FUNCTION AND SAUER--SHELAH
% Brief recall (proved in Ch 10).
% ================================================================

\section{The Growth Function}
\label{sec:growth-function}

The growth function $\growth_\mathcal{H}(m) = \max_{|S|=m}
|\mathcal{H}|_S|$ mediates between VC dimension and generalization bounds.
It is the growth function---not the VC dimension directly---that enters
the uniform convergence argument (Stage~2 of the proof of
\Cref{thm:vc-uniform}).  The VC dimension is a single number; the growth
function is a whole sequence.

The Sauer--Shelah lemma (\Cref{ch:dimensions}) provides the critical
link: if $\VC(\mathcal{H}) = d$, then
$\growth_\mathcal{H}(m) \leq \sum_{i=0}^d \binom{m}{i}
\leq (em/d)^d$ for $m \geq d$.
The dichotomy is sharp: either $\growth_\mathcal{H}(m) = 2^m$ for all~$m$
(when $\VC = \infty$) or $\growth_\mathcal{H}(m) = O(m^d)$
(when $\VC = d < \infty$).  There is no intermediate growth rate.


% ================================================================
% SECTION 5: PAC-BAYES BOUNDS
% Profile E (obstruction).  The prior is NOT a Bayesian prior.
% ================================================================

\section{PAC-Bayes Bounds}
\label{sec:pac-bayes}

The PAC-Bayes framework replaces the worst-case uniform convergence of
Rademacher bounds with a \emph{distribution over hypotheses}.  Instead of
bounding the risk of every $h \in \mathcal{H}$ simultaneously, it bounds
the expected risk of a hypothesis drawn from a learned posterior~$\rho$,
measured relative to a data-independent prior~$\pi$.

\begin{defn}[PAC-Bayes Setting]
\label{def:pac-bayes-setting}
Fix a hypothesis space~$\mathcal{H}$, a loss
$\ell \colon \mathcal{H} \times (X \times Y) \to [0,1]$,
a \emph{prior} distribution~$\pi$ over~$\mathcal{H}$ chosen
\emph{before} seeing data, and a training sample $S \sim D^m$.
A \emph{posterior}~$\rho$ is any distribution over~$\mathcal{H}$ that
may depend on~$S$.  Define
\[
  \risk_D(\rho) = \E_{h \sim \rho}[\risk_D(h)], \qquad
  \emprisk_S(\rho) = \E_{h \sim \rho}[\emprisk_S(h)].
\]
\end{defn}

\begin{thm}[McAllester's PAC-Bayes Bound
{\cite{McAllester1999}}]
\label{thm:pac-bayes}
For any prior~$\pi$ over~$\mathcal{H}$ (chosen independently of~$S$),
any $\delta > 0$, with probability at least $1 - \delta$ over
$S \sim D^m$: for all posteriors~$\rho$ over~$\mathcal{H}$
simultaneously,
\[
  \risk_D(\rho) \leq \emprisk_S(\rho)
    + \sqrt{\frac{\KL(\rho \| \pi) + \log(2\sqrt{m}/\delta)}{2m}}.
\]
\end{thm}

\begin{proof}
The proof has four steps: change-of-measure, moment generating function
bound, prior expectation, and optimization.

\textbf{Step 1: Change-of-measure inequality.}
For any measurable function $\phi \colon \mathcal{H} \to \R$ and
distributions $\rho$,~$\pi$, the Donsker--Varadhan variational formula
gives
\[
  \E_{h \sim \rho}[\phi(h)] \leq \KL(\rho \| \pi)
    + \log \E_{h \sim \pi}[e^{\phi(h)}].
\]
This holds for every~$\rho$; it converts a bound on the moment generating
function under~$\pi$ into a bound under any~$\rho$, at the cost of
$\KL(\rho \| \pi)$.

\textbf{Step 2: Moment generating function bound.}
Set $\phi(h) = \lambda(\risk_D(h) - \emprisk_S(h))$ for a parameter
$\lambda > 0$ to be optimized.  For a fixed~$h$ (independent of~$S$),
$\emprisk_S(h) = \frac{1}{m}\sum_{i=1}^m \ell(h, z_i)$ is an average of
i.i.d.\ $[0,1]$-valued random variables with mean~$\risk_D(h)$.  By
Hoeffding's lemma,
\[
  \E_S\!\left[e^{\lambda(\risk_D(h) - \emprisk_S(h))}\right]
  \leq e^{\lambda^2/(8m)}.
\]

\textbf{Step 3: Prior expectation.}
Since $\pi$ is independent of~$S$, we can exchange the expectations:
\[
  \E_S\!\left[\E_{h \sim \pi}\!\left[
    e^{\lambda(\risk_D(h) - \emprisk_S(h))}\right]\right]
  = \E_{h \sim \pi}\!\left[\E_S\!\left[
    e^{\lambda(\risk_D(h) - \emprisk_S(h))}\right]\right]
  \leq e^{\lambda^2/(8m)}.
\]
By Markov's inequality applied to the non-negative random variable
$\E_{h \sim \pi}[e^{\lambda(\risk_D(h) - \emprisk_S(h))}]$: with
probability $\geq 1 - \delta$,
\[
  \log \E_{h \sim \pi}\!\left[
    e^{\lambda(\risk_D(h) - \emprisk_S(h))}\right]
  \leq \frac{\lambda^2}{8m} + \log\frac{1}{\delta}.
\]

\textbf{Step 4: Combine and optimize.}
Applying Step~1 to the event from Step~3: with probability
$\geq 1 - \delta$, for all~$\rho$,
\[
  \lambda\bigl(\risk_D(\rho) - \emprisk_S(\rho)\bigr)
  \leq \KL(\rho \| \pi) + \frac{\lambda^2}{8m} + \log\frac{1}{\delta}.
\]
Dividing by~$\lambda$ and optimizing
$\lambda = \sqrt{8m(\KL(\rho\|\pi) + \log(1/\delta))}$ yields
\[
  \risk_D(\rho) - \emprisk_S(\rho)
  \leq \sqrt{\frac{2(\KL(\rho \| \pi) + \log(1/\delta))}{m}}.
\]
The slightly refined constant in \Cref{thm:pac-bayes} follows from
a tighter application using Catoni's approach~\cite{Catoni2007}.
\end{proof}


\begin{obstbox}
\textbf{The ``Bayesian'' in PAC-Bayes is an obstruction, not a
feature.}

The PAC-Bayes bound looks Bayesian: it involves a ``prior''~$\pi$
and a ``posterior''~$\rho$, and the complexity term $\KL(\rho\|\pi)$
penalizes departure from the prior.  But the prior plays a
fundamentally different role from a Bayesian prior.

In Bayesian inference, the prior $\pi$ represents \emph{beliefs}
about which hypothesis is true.  The posterior is obtained by Bayes'
rule: $\rho(h) \propto \pi(h) \cdot \prod_i p(z_i \mid h)$.  The
prior must be honest---it should reflect actual uncertainty.

In PAC-Bayes, the prior is a \emph{reference measure} chosen for
technical convenience.  It need not represent beliefs.  The ``posterior''
is \emph{any} distribution over~$\mathcal{H}$, not necessarily the
Bayesian posterior.  The bound holds for all~$\rho$ simultaneously,
so one can optimize~$\rho$ to minimize the bound.  The optimal
$\rho$ is the \emph{Gibbs posterior}
$\rho_\lambda \propto e^{-\lambda \emprisk_S(h)} \pi(h)$, which
resembles a Bayesian posterior but with a free temperature
parameter~$\lambda$ that a true Bayesian would set to~$m$.

\emph{Type mismatch:} Bayesian prior lives in the space of beliefs;
PAC-Bayes prior lives in the space of reference measures.  They have
the same mathematical type (distributions over~$\mathcal{H}$) but
different \emph{epistemic types}.  This is why PAC-Bayes bounds hold
for \emph{any} prior, including bad ones---the bound simply becomes
large.  A Bayesian analysis with a bad prior can give arbitrarily
wrong posteriors; a PAC-Bayes analysis with a bad prior gives a
correct but loose bound.
\end{obstbox}


% ================================================================
% SECTION 6: ALGORITHMIC STABILITY
% ================================================================

\section{Algorithmic Stability}
\label{sec:stability}

Stability bounds abandon the hypothesis class entirely and focus on the
\emph{algorithm}: if replacing a single training example changes the
output hypothesis only slightly, the algorithm generalizes.

\begin{defn}[Uniform Stability {\cite{BousquetElisseeff2002}}]
\label{def:uniform-stability}
A (possibly randomized) algorithm~$A$ is \emph{$\beta$-uniformly stable}
with respect to loss~$\ell$ if, for all samples
$S = (z_1, \ldots, z_m)$ and all samples~$S^{(i)}$ obtained by
replacing~$z_i$ with an independent copy~$z_i'$:
\[
  \sup_{z \in X \times Y}
    \bigl|\ell(A(S), z) - \ell(A(S^{(i)}), z)\bigr|
  \leq \beta.
\]
\end{defn}

\begin{thm}[Stability Generalization Bound
{\cite{BousquetElisseeff2002}}]
\label{thm:stability-bound}
If $A$ is $\beta$-uniformly stable and $\ell$ takes values in $[0,1]$,
then for any $\delta > 0$, with probability at least $1 - \delta$:
\[
  |\risk_D(A(S)) - \emprisk_S(A(S))|
  \leq 2\beta + (4m\beta + 1)\sqrt{\frac{\log(1/\delta)}{2m}}.
\]
In particular, if $\beta = O(1/m)$, then the generalization gap is
$O(1/\sqrt{m})$.
\end{thm}

\begin{proof}
Define $\Delta(S) = \risk_D(A(S)) - \emprisk_S(A(S))$.  The proof
proceeds in two stages.

\emph{Stage 1: Bound $\E[\Delta(S)]$.}
Write $\risk_D(A(S)) = \E_{z'}[\ell(A(S), z')]$ and
$\emprisk_S(A(S)) = \frac{1}{m}\sum_i \ell(A(S), z_i)$.  By symmetry of
the i.i.d.\ draw,
$\E[\ell(A(S), z_i)] = \E[\ell(A(S^{(i)}), z_i')]$
up to an error of~$\beta$ (by uniform stability applied to~$S$
vs.~$S^{(i)}$).  Summing over~$i$ and using
$\E[\ell(A(S^{(i)}), z_i')] = \E[\risk_D(A(S^{(i)}))] =
\E[\risk_D(A(S))]$ (by identical distribution), one obtains
$|\E[\Delta(S)]| \leq \beta$.

\emph{Stage 2: Concentration.}
Replacing~$z_i$ changes $\emprisk_S(A(S))$ by at most $1/m$ (direct
computation) and changes $\risk_D(A(S))$ by at most~$\beta$ (via
stability).  So $\Delta(S)$ has bounded differences
$c_i \leq 2\beta + 1/m$.  McDiarmid's inequality gives
\[
  \Pr[|\Delta(S) - \E[\Delta(S)]| > t]
  \leq 2\exp\!\left(\frac{-2t^2}{\sum_i c_i^2}\right)
  \leq 2\exp\!\left(\frac{-2t^2}{m(2\beta+1/m)^2}\right).
\]
Combining with $|\E[\Delta(S)]| \leq \beta$ gives the stated bound.
\end{proof}

\begin{example}[Regularized ERM is stable]
\label{ex:regularized-erm-stable}
Consider regularized empirical risk minimization:
$A(S) = \arg\min_{h \in \mathcal{H}} \emprisk_S(h) +
\lambda \|h\|^2$.  For $\lambda$-strongly convex regularizers and
$L$-Lipschitz losses, $\beta \leq L^2/(2\lambda m)$, giving
$\beta = O(1/m)$ and hence $O(1/\sqrt{m})$
generalization~\cite{BousquetElisseeff2002}.
\end{example}

\begin{remark}[Stability beyond binary classification]
\label{rmk:stability-beyond}
The deepest result in stability theory is due to Shalev-Shwartz et
al.\ (2010): beyond binary classification with 0--1 loss,
\emph{learnability is equivalent to the existence of a stable empirical
risk minimizer}.  Uniform convergence fails to characterize learnability
in this more general setting, but stability succeeds.  This is a
fundamental structural result: stability replaces uniform convergence as
the correct notion when one leaves the binary classification world.
\end{remark}


% ================================================================
% SECTION 7: INFORMATION-THEORETIC BOUNDS
% ================================================================

\section{Information-Theoretic Bounds}
\label{sec:info-bound}

The most recent entry in the generalization-bound landscape connects
generalization to the mutual information between the training sample and
the learned hypothesis.

\begin{thm}[Information-Theoretic Generalization Bound
{\cite{XuRaginsky2017}}]
\label{thm:xu-raginsky}
Let $A$ be a (possibly randomized) learning algorithm, $\ell$ a loss
taking values in $[0,1]$, and $W = A(S)$ the output hypothesis.  Let
$I(S; W)$ denote the mutual information between~$S$ and~$W$.  Then
\[
  \bigl|\E[\risk_D(W) - \emprisk_S(W)]\bigr|
  \leq \sqrt{\frac{I(S; W)}{2m}}.
\]
\end{thm}

\begin{proof}[Proof sketch]
The argument uses the transportation lemma, a consequence of Pinsker's
inequality and the Donsker--Varadhan representation of KL divergence.

For each index~$i$, define
$g_i(w) = \E_Z[\ell(w, Z)] - \ell(w, Z_i)$, where the first expectation
is over a fresh test point $Z \sim D$.  Then
$\E[\mathrm{gen}(W,S)] = \frac{1}{m}\sum_{i=1}^m \E[g_i(W)]$.

For each~$i$, the function $g_i$ satisfies $|g_i| \leq 1$, so $g_i(W)$
is $1$-sub-Gaussian under the marginal distribution of~$W$ given all
other samples.  The transportation lemma gives
$|\E[g_i(W)]| \leq \sqrt{2 I(W; Z_i)}$.
Averaging over~$i$ and using the chain rule
$\sum_i I(W; Z_i) \leq I(W; S)$ (by the data-processing inequality
applied to the Markov chain
$Z_i \to S \to W$) yields the result.
\end{proof}

\begin{remark}[Conditional mutual information]
\label{rmk:cmi}
The Xu--Raginsky bound can be vacuous for deterministic algorithms,
since $I(S; W) = H(W)$ when $W = A(S)$ is a deterministic function
of~$S$.  Steinke and Zakynthinou (2020) introduced the \emph{conditional
mutual information} (CMI) framework, which replaces $I(S; W)$ with
$I(A(\tilde{S}_U); U \mid \tilde{S})$, where $\tilde{S}$ is a
``supersample'' of $2m$ points and $U \in \{0,1\}^m$ selects which half
to train on.  The CMI is always bounded by $m \log 2$ and gives
nontrivial bounds even for deterministic algorithms.  This framework
unifies PAC-Bayes, VC, and stability bounds within a single
information-theoretic language.
\end{remark}


% ================================================================
% SECTION 8: MARGIN THEORY
% Profile F (Open Frontier).
% ================================================================

\section{Margin Theory}
\label{sec:margin}

Margin-based bounds were the original motivation for support vector
machines and remain the primary tool for analyzing neural network
generalization.

\begin{defn}[Margin]
\label{def:margin}
For a classifier $f \colon X \to \R$ and a labeled example $(x, y)$
with $y \in \{-1, +1\}$, the \emph{margin} is
$\gamma(f, x, y) = y \cdot f(x)$.  The margin is positive when the
prediction is correct, and its magnitude measures the ``confidence'' of
the prediction.
\end{defn}

\begin{thm}[Margin-Based Generalization Bound]
\label{thm:margin-bound}
Let $\mathcal{F}$ be a class of functions $f \colon X \to \R$ with
$\|f\|_\infty \leq B$, and let $\gamma > 0$.  Then for any $\delta > 0$,
with probability at least $1 - \delta$, for all $f \in \mathcal{F}$:
\[
  \Pr_{(x,y) \sim D}[y \cdot f(x) \leq 0]
  \leq \frac{1}{m}\sum_{i=1}^m \ind[y_i f(x_i) \leq \gamma]
  + \frac{4}{\gamma}\,\Rad_S(\mathcal{F})
  + \sqrt{\frac{\log(2/\delta)}{2m}}.
\]
\end{thm}

\begin{proof}[Proof sketch]
Define the ramp loss $\ell_\gamma(t) = \min(1, \max(0, 1 - t/\gamma))$,
which is $1/\gamma$-Lipschitz and satisfies
$\ind[t \leq 0] \leq \ell_\gamma(t) \leq \ind[t \leq \gamma]$.  Then
\begin{align*}
  \Pr_D[y f(x) \leq 0]
  &\leq \E_D[\ell_\gamma(y f(x))] \\
  &\leq \frac{1}{m}\sum_i \ell_\gamma(y_i f(x_i))
    + 2\Rad_S(\ell_\gamma \circ \mathcal{F})
    + \sqrt{\frac{\log(2/\delta)}{2m}} \\
  &\leq \frac{1}{m}\sum_i \ind[y_i f(x_i) \leq \gamma]
    + \frac{2}{\gamma}\cdot 2\Rad_S(\mathcal{F})
    + \sqrt{\frac{\log(2/\delta)}{2m}},
\end{align*}
using the Rademacher contraction lemma
($\Rad_S(\phi \circ \mathcal{F}) \leq L \cdot \Rad_S(\mathcal{F})$
for $L$-Lipschitz~$\phi$) in the last step.
\end{proof}

\begin{example}[Halfspaces with margin]
\label{ex:halfspace-margin}
Let $\mathcal{F} = \{x \mapsto \langle w, x \rangle : \|w\|_2 \leq W\}$
with data satisfying $\|x_i\| \leq B$.  From the computational
illustration in \Cref{sec:rademacher},
$\Rad_S(\mathcal{F}) \leq WB/\sqrt{m}$.  The margin bound gives
\[
  \Pr_D[y \langle w, x \rangle \leq 0]
  \leq \hat{L}_\gamma(w) + \frac{4WB}{\gamma\sqrt{m}}
    + \sqrt{\frac{\log(2/\delta)}{2m}},
\]
where $\hat{L}_\gamma(w)$ is the fraction of training points with margin
$\leq \gamma$.  Observe: the bound depends on $WB/\gamma$ (the ratio of
norm to margin), not on the ambient dimension.  This is why SVMs with
large-margin solutions generalize well even in very high dimensions.
\end{example}

\begin{example}[Neural networks: the open frontier]
\label{ex:neural-margin}
For a depth-$L$ network $f_W(x) = W_L \sigma(W_{L-1} \sigma(\cdots
W_1 x \cdots))$ with ReLU activation~$\sigma$, Bartlett, Foster, and
Telgarsky (2017) showed that the Rademacher complexity satisfies
\[
  \Rad_S(\mathcal{F})
  \leq \frac{B_x \cdot (\sqrt{2\log(2d_{\max})})^L
    \cdot \prod_{i=1}^L \|W_i\|_\sigma}{\sqrt{m}},
\]
where $\|W_i\|_\sigma$ is the spectral norm of the $i$-th layer and
$d_{\max}$ is the maximum width.  Combined with the margin bound, this
gives a generalization bound scaling as the product of spectral norms
divided by the margin.

The problem: for practical networks, this product is large.  A ResNet-18
on CIFAR-10 has $\prod_i \|W_i\|_\sigma \approx 10^3$, making the bound
vacuous (larger than~1).  Yet the network generalizes perfectly well.  The
resolution of this tension---finding bounds that are both non-vacuous and
predictive---remains one of the central open problems in learning theory.
\end{example}

\begin{openprob}[Tight deep network bounds]
\label{op:deep-margins}
For deep neural networks, all known generalization bounds---margin-based,
PAC-Bayes, stability, and information-theoretic---are \emph{loose} in
the following sense: the bounds scale with quantities that grow with
network size, yet empirically, larger networks generalize \emph{better}.

The specific open problem: find a generalization bound for deep networks
trained by stochastic gradient descent that (a)~is non-vacuous (smaller
than~1) on standard benchmarks, (b)~correctly predicts that increasing
width improves generalization, and (c)~does not require post-hoc
compression or quantization.  As of 2025, no known framework achieves all
three simultaneously.
\end{openprob}


% ================================================================
% SECTION 9: META-LEARNING BOUNDS
% Profile A-plus.  Brief.
% ================================================================

\section{Meta-Learning Bounds}
\label{sec:meta-pac}

\begin{defn}[Meta-PAC Bound {\cite{Baxter2000}}]
\label{def:meta-pac}
In the meta-learning setting, the learner observes $T$ tasks, each
providing $m$ samples from a task-specific distribution~$D_t$.  The tasks
are drawn from a task distribution~$\mathcal{T}$.  Baxter's meta-PAC
bound states that if the bias class~$\mathcal{B}$ has finite complexity,
then with $T$ tasks and $m$ samples per task:
\[
  \risk_\mathcal{T}(\hat{b}) \leq \emprisk_{T,m}(\hat{b})
  + O\!\left(\sqrt{\frac{\log \mathcal{N}(\varepsilon,\mathcal{B},d)}{T}}
    + \sqrt{\frac{d_{\mathrm{task}}}{m}}\right),
\]
where $d_{\mathrm{task}}$ is the VC dimension of the task-level
hypothesis class induced by the bias.  The first term controls
meta-generalization; the second controls within-task generalization.
\end{defn}


% ================================================================
% SECTION 10: THE COMPARISON
% CENTERPIECE: Profile G.  Five frameworks in parallax.
% ================================================================

\section{Five Frameworks Compared}
\label{sec:comparison}

The five frameworks bound the same quantity but use different information
and succeed in different regimes.  \Cref{tab:five-frameworks} compares
them along five axes.

\begin{table}[ht]
\centering\small
\caption{Five generalization-bound frameworks compared.  Each row
describes what the framework measures, what it bounds, where it is
tightest, and where it fails.}
\label{tab:five-frameworks}
\renewcommand{\arraystretch}{1.4}
\begin{tabularx}{\textwidth}{>{\bfseries\small}l X X X X}
\toprule
\textbf{Framework} & \textbf{Key quantity} & \textbf{What it bounds}
  & \textbf{Tight for} & \textbf{Fails on} \\
\midrule
Rademacher
  & $\Rad_S(\mathcal{F})$: noise correlation
  & Uniform over all $f \in \mathcal{F}$
  & Finite VC classes; linear models
  & Overparameterized models (vacuous) \\
\addlinespace
PAC-Bayes
  & $\KL(\rho \| \pi)$: posterior--prior divergence
  & Expected risk under posterior~$\rho$
  & Bayesian and ensemble methods
  & Requires meaningful prior \\
\addlinespace
Stability
  & $\beta$: one-sample sensitivity
  & $|\risk - \emprisk|$ for algorithm~$A$
  & Regularized ERM; SGD
  & Non-regularized ERM (unstable) \\
\addlinespace
Information
  & $I(S; W)$: mutual information
  & $|\E[\risk - \emprisk]|$
  & Noisy algorithms (DP-SGD)
  & Deterministic algorithms (vacuous) \\
\addlinespace
Margin
  & $\gamma$: prediction confidence
  & $\Pr[\text{error}]$ via margin loss
  & SVMs; kernel methods
  & Deep networks (norms too large) \\
\bottomrule
\end{tabularx}
\end{table}


\subsection{The Parallax Diagram}

\Cref{fig:five-frameworks} displays the implication and obstruction
relationships among the five frameworks.  The diagram should be read as a
\emph{parallax}: five views of the same phenomenon (generalization),
each projecting the three-way interaction between data, class, and
algorithm onto a different axis.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
    fbox/.style={draw, thick, rounded corners=4pt, minimum width=2.8cm,
      minimum height=0.9cm, font=\small\bfseries, align=center},
    arr/.style={-{Stealth[length=2.5mm]}, thick},
    darr/.style={arr, defblue},
    oarr/.style={{Stealth[length=2.5mm]}-{Stealth[length=2.5mm]}, thick,
      analogypurple, dashed},
    xarr/.style={arr, thmred, dashed},
    lbl/.style={font=\scriptsize, midway, fill=white, inner sep=1pt}
]
% Nodes
\node[fbox, fill=defblue!10] (vc) at (-3.5,0) {VC / Uniform\\Convergence};
\node[fbox, fill=defblue!15] (rad) at (0,0) {Rademacher};
\node[fbox, fill=sepgreen!10] (pb) at (4.5,0) {PAC-Bayes};
\node[fbox, fill=goldcolor!10] (stab) at (-1.5,-2.8) {Stability};
\node[fbox, fill=edgeorange!10] (info) at (3,-2.8) {Information-\\Theoretic};
\node[fbox, fill=boundbrown!10] (margin) at (0,2.5) {Margin};

% Implications (solid blue arrows)
\draw[darr] (margin) -- node[lbl, left] {specializes} (rad);
\draw[darr] (rad) -- node[lbl, above] {refines} (vc);
\draw[darr] (stab) -- node[lbl, below] {implies} (info);

% Dual relationship (dashed purple double arrow)
\draw[oarr] (pb) -- node[lbl, right, xshift=2pt]
  {\textcolor{analogypurple}{dual views}} (info);

% Obstructions (dashed red)
\draw[xarr] (info) -- node[lbl, left, xshift=-2pt, rotate=30]
  {\textcolor{thmred}{orthogonal}} (rad);
\draw[xarr] (stab) -- node[lbl, below, yshift=-2pt]
  {\textcolor{thmred}{type mismatch}} (pb);

% Dotted connection: VC to Rademacher via Sauer-Shelah
\draw[darr, densely dotted] (vc) to[bend right=20]
  node[lbl, below] {\scriptsize Sauer--Shelah} (rad);
\end{tikzpicture}
\caption{Implication and obstruction relationships among the five
generalization frameworks.  Solid arrows: one framework implies or
refines the other.  Dashed purple: dual or analogous frameworks.  Dashed
red: obstruction (the frameworks measure orthogonal aspects or have a
type mismatch).  Dotted: the Sauer--Shelah lemma mediates the
VC-to-Rademacher connection.}
\label{fig:five-frameworks}
\end{figure}


\subsection{Cross-Framework Relationships}

The five frameworks are not independent.  Several implication, analogy,
and obstruction relationships connect them:

\begin{enumerate}[leftmargin=*, itemsep=3pt]
\item \textbf{Rademacher $\Rightarrow$ VC.}  The Rademacher complexity of
  a binary class satisfies $\mathcal{R}_m(\mathcal{H}) \leq
  \sqrt{2\VC(\mathcal{H}) \log(em/\VC(\mathcal{H}))/m}$
  (\Cref{prop:rad-vc}), so Rademacher bounds subsume and refine VC
  bounds.  The refinement is strict: Rademacher complexity adapts to the
  sample, whereas VC dimension does not.

\item \textbf{Margin $\Rightarrow$ Rademacher.}  The margin bound
  (\Cref{thm:margin-bound}) is derived \emph{from} Rademacher complexity
  applied to the ramp loss, so margin bounds are a specialization.

\item \textbf{Stability $\Rightarrow$ Information.}  A $\beta$-uniformly
  stable algorithm satisfies $I(S; W) \leq O(m \beta^2)$ (via the
  data-processing inequality applied to the Markov chain
  $Z_i \to S \to W$).  Thus stability bounds imply information-theoretic
  bounds.

\item \textbf{PAC-Bayes $\leftrightarrow$ Information.}  The PAC-Bayes
  bound with a data-independent prior can be interpreted as an
  information-theoretic bound: the KL divergence
  $\KL(\rho \| \pi)$ upper-bounds a specific mutual information
  term.  Conversely, the information-theoretic framework recovers
  PAC-Bayes via a change-of-measure argument.  The two are essentially
  dual views~\cite{HellstromDurisi2020}.

\item \textbf{Obstruction: Information $\not\Rightarrow$ Rademacher.}
  Information-theoretic bounds depend on the \emph{algorithm}; Rademacher
  bounds depend on the \emph{class}.  A deterministic ERM algorithm has
  maximal $I(S; W)$ but may operate on a class with small Rademacher
  complexity.  The two frameworks capture orthogonal aspects of
  generalization.

\item \textbf{Obstruction: Stability $\not\Rightarrow$ PAC-Bayes.}
  Stability measures perturbation sensitivity of the \emph{algorithm}.
  PAC-Bayes measures divergence of the \emph{posterior} from the
  \emph{prior}.  An algorithm defines a posterior (connecting them), but
  the primary objects are different types: stability lives in the space of
  algorithms, PAC-Bayes in the space of distributions.
\end{enumerate}


\subsection{What Each Framework Sees}

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
  every node/.style={font=\small},
  axis/.style={thick, -{Stealth[length=3mm]}},
  dot/.style={circle, fill, inner sep=2pt},
  lbl/.style={font=\footnotesize, text width=2.5cm, align=center}
]
% Triangle of information
\coordinate (D) at (0,0);
\coordinate (H) at (6,0);
\coordinate (A) at (3,4.5);

\draw[thick, fill=layergray] (D) -- (H) -- (A) -- cycle;

\node[dot, defblue] at (D) {};
\node[dot, thmred] at (H) {};
\node[dot, sepgreen] at (A) {};

\node[below left] at (D) {Data $D$};
\node[below right] at (H) {Class $\mathcal{H}$};
\node[above] at (A) {Algorithm $A$};

% Framework labels along edges/vertices
\node[lbl, defblue, below] at (3,0)
  {Rademacher\\Margin\\(class axis)};
\node[lbl, sepgreen, rotate=55, anchor=south] at (1.2,2.6)
  {Stability\\(algo + data)};
\node[lbl, thmred, rotate=-55, anchor=south] at (4.8,2.6)
  {PAC-Bayes\\(algo + class)};
\node[lbl, edgeorange] at (3,1.8)
  {Information\\$I(S;W)$\\(channel)};
\end{tikzpicture}
\caption{Each framework projects the data--class--algorithm triangle
onto a different axis.  Rademacher and margin bounds project onto the
class axis; stability onto the algorithm--data edge; PAC-Bayes onto
the algorithm--class edge; information-theoretic bounds onto the
``channel'' in the interior.}
\label{fig:triangle}
\end{figure}


\subsection{Which Framework When?}

No single framework dominates.  The choice depends on what is known:

\begin{itemize}[leftmargin=*, itemsep=3pt]
\item \textbf{If you know the class but not the algorithm}: use
  Rademacher or margin bounds.
\item \textbf{If you know the algorithm but not the class}: use
  stability or information-theoretic bounds.
\item \textbf{If you can choose a prior}: use PAC-Bayes.
\item \textbf{If you want algorithm-specific bounds for SGD}: stability
  and information-theoretic bounds apply; Rademacher bounds do not.
\item \textbf{If you want non-vacuous bounds for deep networks}:
  PAC-Bayes with learned priors is currently the most successful
  approach, but the bounds remain far from tight
  (\Cref{op:deep-margins}).
\end{itemize}


\begin{example}[Three bounds on the same problem]
\label{ex:three-bounds}
Consider linear classification with $\|w\| \leq 1$ on data with
$\|x\| \leq 1$ in $\R^d$, with $m$ training points and 0--1 loss.

\textbf{VC bound.}  $\VC = d+1$, giving generalization gap
$O(\sqrt{d \log m / m})$.  This scales linearly with dimension.

\textbf{Rademacher bound.}  $\Rad_S \leq 1/\sqrt{m}$
(from the computation in \Cref{sec:rademacher}), giving generalization
gap $O(1/\sqrt{m})$.  No dimension dependence.

\textbf{Margin bound.}  If the margin is $\gamma$, the bound is
$\hat{L}_\gamma + O(1/(\gamma\sqrt{m}))$.  If the data is
well-separated ($\gamma = 0.1$, $\hat{L}_\gamma = 0$), this gives
$O(10/\sqrt{m})$---worse constant but the zero margin-loss term
compensates.

The three bounds are complementary.  In dimension $d = 1000$ with
$m = 5000$, the VC bound gives approximately $0.6$ (nearly vacuous), the
Rademacher bound gives approximately $0.04$ (useful), and the margin
bound with $\gamma = 0.1$ gives approximately $0.14$ plus the margin
loss.
\end{example}


% ================================================================
% SECTION 11: CONCRETE COMPARISON ON NEURAL NETWORKS
% ================================================================

\section{A Case Study: Neural Networks}
\label{sec:nn-comparison}

The five frameworks yield strikingly different results on neural
networks, making them the ideal testbed for understanding what each
framework captures.

\begin{computational}
\textbf{Two-layer ReLU network on $\R^{100}$.}
Consider $f_W(x) = v^\top \sigma(Wx)$ with $W \in \R^{k \times 100}$,
$v \in \R^k$, width $k = 1000$, and ReLU activation $\sigma$.

\smallskip
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Framework} & \textbf{Bound expression} & \textbf{With $m=10^4$} \\
\midrule
VC & $O(\sqrt{k \cdot 100 \cdot \log(m) / m})$ & $\approx 3.4$ (vacuous) \\
Rademacher & $O(\|v\|_1 \|W\|_\sigma B_x / \sqrt{m})$
  & $\approx 0.3$ (if norms bounded) \\
PAC-Bayes & $O(\sqrt{\KL(\rho\|\pi)/m})$
  & $\approx 0.05$ (with good prior) \\
Stability & $O(L^2/(\lambda m))$ for reg.\ ERM
  & $\approx 0.1$ (if regularized) \\
Margin & $O(\|v\|_1 \|W\|_\sigma / (\gamma\sqrt{m}))$
  & depends on $\gamma$ \\
\bottomrule
\end{tabular}

\smallskip
The VC bound is vacuous because it counts parameters ($\sim 10^5$).
Rademacher and margin bounds depend on norms, which can be controlled by
regularization.  PAC-Bayes gives the tightest bound when a meaningful
prior is available (e.g., the initialization distribution of SGD).
Stability bounds require regularization or early stopping to
ensure~$\beta = O(1/m)$.  The information-theoretic bound depends on the
noise level of SGD: more noise means less $I(S;W)$ and a tighter bound,
but also worse training performance.
\end{computational}


% ================================================================
% SECTION 12: HISTORICAL CONTEXT
% ================================================================

\begin{historical}
The study of generalization bounds spans five decades and five research
communities.

The VC theory (Vapnik--Chervonenkis, 1971) established the first uniform
convergence bounds.  The connection to PAC learning was made explicit by
Blumer et al.\ (1989).  Rademacher complexity was developed independently
by Koltchinskii (2001) and Bartlett--Mendelson (2002) as a data-dependent
refinement.

PAC-Bayes bounds originated with Shawe-Taylor and Williamson (1997) and
were formalized by McAllester (1998, 1999).  Catoni (2007) developed the
optimal ``thermodynamic'' form.  The framework gained renewed interest
around 2017 when Dziugaite and Roy showed that PAC-Bayes bounds can be
made non-vacuous for deep networks.

Algorithmic stability was introduced by Devroye and Wagner (1979) in a
simpler form, and the modern uniform stability framework was established
by Bousquet and Elisseeff (2002).  The connection to learnability beyond
binary classification was established by Shalev-Shwartz et al.\ (2010).

Information-theoretic bounds are the most recent tradition, beginning
with Russo and Zou (2016) and Xu and Raginsky (2017).  The conditional
mutual information framework of Steinke and Zakynthinou (2020) provided
the unifying perspective.

Margin theory has the longest applied history (Vapnik, 1995; Bartlett,
1998) and was reinvigorated by Bartlett, Foster, and Telgarsky (2017)
for deep networks.

The unification of these traditions---showing they are all projections
of a single underlying phenomenon---remains an active research
direction.
\end{historical}


% ================================================================
% NOTES
% ================================================================

\bigskip

\begin{remark}[The chapter in context]
This chapter completes Part~III's treatment of quantitative measures of
learning complexity.  \Cref{ch:dimensions} developed the combinatorial
dimensions (VC, Littlestone, DS).  \Cref{ch:compression} connected
learnability to sample compression.  The present chapter adds the
\emph{analytic} layer: bounds that depend not just on the class but on
the algorithm, the data, and the geometry of the learned hypothesis.
The five frameworks exemplify a recurring theme of this book: the same
mathematical phenomenon---in this case, the gap between training and
test performance---looks different depending on which structural
information one chooses to measure.  No single framework captures the
full picture, and the obstructions between frameworks (\Cref{fig:five-frameworks})
are as informative as the implications.
\end{remark}
