% Chapter 15: Analogies and Their Obstructions
% Organized by OBSTRUCTION TYPE, not by analogy source.
% Content: ALL 32 analogy edges from the graph.

\chapter{Analogies and Their Obstructions}
\label{ch:analogies}

The concept graph contains 32 edges of type \rel{analogy}. In most
treatments of learning theory, such analogies appear as informal remarks:
``compression is like VC dimension,'' or ``stability is related to
PAC-Bayes.'' This chapter treats each analogy as a formal object with
three components:

\begin{enumerate}[leftmargin=*, nosep]
\item A \emph{source} and \emph{target}: two concepts that share a
structural parallel.
\item A \emph{shared structure}: the precise mathematical property that
makes the analogy plausible.
\item An \emph{obstruction}: the precise reason the analogy fails to be
a theorem, classified by \emph{type}.
\end{enumerate}

The key insight of this chapter is that the obstruction type is more
informative than the analogy itself. Two analogies may look superficially
similar (``X is like Y because both measure complexity'') but fail for
entirely different reasons: one because the objects live in different
type systems, another because a conjectured equivalence remains unproved.
The obstruction type tells you \emph{why} the analogy fails, and this
``why'' determines whether the analogy might become a theorem in the
future, or is structurally blocked.

The 32 analogy edges distribute across six obstruction types:

\begin{center}\small
\begin{tabular}{llr}
\toprule
\textbf{Obstruction type} & \textbf{Meaning} & \textbf{Count} \\
\midrule
Type mismatch & Objects live in different formal categories & 12 \\
Missing equivalence witness & Plausible but unproved & 9 \\
One-way theorem only & One direction proved, reverse open/false & 4 \\
Data model mismatch & Incompatible data-generating assumptions & 3 \\
Proof method mismatch & Same structure, non-transferable proofs & 3 \\
Success criterion mismatch & Different optimization objectives & 1 \\
\bottomrule
\end{tabular}
\end{center}

The distribution is itself revealing. The plurality of obstructions
are type mismatches---analogies that fail because the mathematical
objects being compared belong to different formal categories. The second
largest group, missing equivalence witnesses, represents the frontier:
analogies that \emph{might} become theorems if someone proves the right
conjecture.


% ================================================================
% SECTION 1: TYPE MISMATCH (12 edges)
% ================================================================

\section{Type Mismatch}
\label{sec:type-mismatch}

A type mismatch obstruction arises when the source and target of an
analogy belong to different formal categories in the concept graph.
The analogy is structurally plausible---both concepts ``do something
similar''---but they operate on different mathematical objects, and no
bridging theorem connects the two type systems.

Type mismatches are the most common obstruction (12 of 32 edges) and
generally the least likely to be resolved. They are not open problems
waiting for a clever proof; they are structural incompatibilities in the
formalism.

\begin{table}[ht]
\centering\small
\caption{Type mismatch analogies. The ``type gap'' column identifies
the category mismatch that blocks a formal theorem.}
\label{tab:type-mismatch}
\begin{tabularx}{\textwidth}{@{}p{4.2cm}p{3.2cm}X@{}}
\toprule
\textbf{Analogy} & \textbf{Shared structure} & \textbf{Type gap} \\
\midrule

\nde{algorithmic\_stability} $\leftrightarrow$ \nde{pac\_bayes\_bound}
& Both bound generalization gap
& Stability is algorithm-dependent (perturbation of $A$). PAC-Bayes is
posterior-dependent (divergence $\KL(Q \| P)$). An algorithm defines a
posterior, but the primary objects differ. \\[6pt]

\nde{information\_theoretic\_bound} $\leftrightarrow$
\nde{pac\_bayes\_bound}
& Both use information-divergence measures
& Info-theoretic bounds use $I(W; S)$ (mutual information between
algorithm output and training set). PAC-Bayes uses $\KL(Q \| P)$
(posterior vs.\ prior). Different information measures on different
objects. Hellstr\"om et al.\ (2023) partially unify via change-of-measure,
but the objects remain distinct. \\[6pt]

\nde{fundamental\_theorem} $\leftrightarrow$
\nde{algorithmic\_stability}
& Both characterize learnability
& The fundamental theorem is a characterization (equivalence result).
Stability is a complexity measure. The structural parallel exists but the
formal types are incompatible without a bridging theorem. \\[6pt]

\nde{sq\_dimension} $\leftrightarrow$ \nde{vc\_dimension}
& Both measure concept class complexity
& SQ dimension measures pairwise correlation structure under a
distribution. VC dimension measures distribution-free shattering.
Exponential gap possible (parities: $\VC = n$, $\SQdim = 2^n$). \\[6pt]

\nde{grammar\_induction} $\leftrightarrow$ \nde{gold\_theorem}
& Both concern language learning
& Grammar induction is a process; Gold's theorem is an impossibility
result. Different concept categories. \\[6pt]

\nde{concept\_drift} $\leftrightarrow$ \nde{ex\_under\_drift}
& Both concern non-stationary learning
& Concept drift is a process; Ex-under-drift is a success criterion.
Different concept categories. \\[6pt]

\nde{lifelong\_learning} $\leftrightarrow$ \nde{meta\_learner}
& Both concern learning across tasks
& Lifelong learning is a process; meta-learner is a learner type.
Different concept categories. \\[6pt]

\nde{background\_knowledge} $\leftrightarrow$ \nde{advice}
& Both provide side information to a learner
& Background knowledge is a process; advice is a data presentation.
Different concept categories. \\[6pt]

\nde{mdl} $\leftrightarrow$ \nde{kolmogorov\_complexity}
& Both measure description length
& MDL uses finite computable codes. Kolmogorov complexity uses the
shortest program on a universal Turing machine (generally
incomputable). The parallel is ``short description = good model,''
but MDL's codes are constructive while Kolmogorov's are not. \\[6pt]

\nde{srm} $\leftrightarrow$ \nde{inductive\_bias}
& Both constrain hypothesis selection
& SRM is a model selection principle; inductive bias is a base type.
The structural parallel exists but the categories are incompatible. \\[6pt]

\nde{bayesian\_inference} $\leftrightarrow$ \nde{inductive\_bias}
& Both encode prior assumptions
& Bayesian inference is a model selection method (probability over
hypotheses). Inductive bias is a base type (constraint on learner
behavior). The prior \emph{is} a specific form of bias, but the type
systems differ. \\[6pt]

\nde{occam\_algorithm} $\leftrightarrow$ \nde{compression\_scheme}
& Both embody ``short representation $\Rightarrow$ generalization''
& Occam uses description length of hypotheses (syntactic: bits to
encode $h$). Compression uses a subset of training data (extensional:
examples to reconstruct $h$). The notion of shortness differs. \\

\bottomrule
\end{tabularx}
\end{table}

The pattern across these 12 edges is consistent: the analogy identifies
a genuine structural parallel, but the source and target live in
different parts of the type system. Making the analogy formal would
require a bridging theorem that relates the two types, and in most cases
no such theorem is expected. These are permanent features of the
landscape, not temporary gaps in knowledge.


% ================================================================
% SECTION 2: MISSING EQUIVALENCE WITNESS (9 edges)
% ================================================================

\section{Missing Equivalence Witness}
\label{sec:missing-equiv}

A missing equivalence witness obstruction marks the most interesting
class of analogies: those where a formal equivalence is plausible but
unproved. Some of these may become theorems; others may eventually be
disproved. They represent the active frontier of the field.

\begin{obstbox}
\textbf{Compression $\leftrightarrow$ VC dimension.}

\medskip\noindent
The sample compression conjecture asserts that every concept class with
VC dimension $d$ admits a sample compression scheme of size $O(d)$.
Moran and Yehudayoff~\cite{MoranYehudayoff2016} proved that finite VC
dimension implies \emph{some} finite compression scheme (of size
exponential in $d$), establishing one direction. The conjecture that
compression size is $O(d)$ remains open. If proved, this would upgrade
the analogy to an equivalence; if disproved, it would become a
separation.

This is arguably the most important open problem connected to the
analogy edges in the graph.
\end{obstbox}

\begin{obstbox}
\textbf{Rademacher complexity $\leftrightarrow$ VC dimension.}

\medskip\noindent
The two are asymptotically related: $\Rad_n(\mathcal{H}) \leq
O\!\left(\sqrt{\VC(\mathcal{H})/n}\right)$. Both measure the complexity
of a hypothesis class, and for binary classification both control
generalization. But Rademacher complexity is data-dependent
(computed from a specific sample and distribution), while VC dimension is
distribution-free. No equivalence holds: Rademacher complexity can be
much smaller than the VC bound for benign distributions, and the two
quantities can disagree on the ranking of hypothesis classes.
\end{obstbox}

\begin{obstbox}
\textbf{VC characterization $\leftrightarrow$ Littlestone
characterization.}

\medskip\noindent
Both theorems have the same logical form: ``a class is learnable in
paradigm $P$ if and only if combinatorial dimension $d$ is finite.''
The parallel is exact at the level of logical structure. But the
dimensions measure different adversarial strengths---batch (VC) vs.\
adaptive (Littlestone)---and $\VC \leq \Ldim$ with no reverse bound.
The gap can be infinite (thresholds on $\R$). Alon et
al.~\cite{AlonLivniMalliarisMoran2019} connected the two via
differential privacy, but no equivalence between the dimensions exists.
\end{obstbox}

\begin{obstbox}
\textbf{MML $\leftrightarrow$ MDL.}

\medskip\noindent
Minimum Message Length (Wallace--Freeman) and Minimum Description Length
(Rissanen) both select hypotheses by minimizing a two-part code: the
model plus the data given the model. They are often treated as variants
of the same idea. However, MML uses a Bayesian prior explicitly and
integrates over the parameter space, while MDL uses a minimax-optimal
universal code. No formal theorem establishes their equivalence or
separation; the two traditions developed independently and use different
mathematical frameworks.
\end{obstbox}

\begin{obstbox}
\textbf{MML $\leftrightarrow$ Bayesian inference.}

\medskip\noindent
MML selects the hypothesis that minimizes total message length; Bayesian
inference selects (or averages over) hypotheses according to posterior
probability. For certain prior--likelihood pairs, MML and MAP
(maximum a posteriori) estimates coincide asymptotically. But MML is a
coding-theoretic criterion and Bayesian inference is a probabilistic
criterion; the equivalence is approximate and model-dependent, not a
general theorem.
\end{obstbox}

\begin{obstbox}
\textbf{PAC learning $\leftrightarrow$ posterior consistency.}

\medskip\noindent
Both are success criteria requiring convergence to the truth. PAC
demands uniform convergence in probability over all distributions;
Bayesian consistency demands that the posterior concentrates on the true
parameter. The structural parallel is ``learning = convergence,'' but
PAC convergence is frequentist and distribution-free, while posterior
consistency is Bayesian and prior-dependent. No general theorem
connects the two: a class can be PAC learnable but Bayesian-inconsistent
under a misspecified prior, and vice versa.
\end{obstbox}

\begin{obstbox}
\textbf{Mistake bound $\leftrightarrow$ Littlestone dimension.}

\medskip\noindent
The optimal mistake bound for a concept class equals its Littlestone
dimension (this is Littlestone's theorem). However, the \emph{analogy}
edge in the graph refers to the structural parallel between the two as
complexity measures in their own right. The mistake bound is defined
operationally (worst-case mistakes of the best algorithm); the
Littlestone dimension is defined combinatorially (depth of the largest
shattered tree). Their equivalence is a theorem, but as stand-alone
objects they measure different things, and extending either to new
settings (e.g., partial monitoring, bandit feedback) produces divergent
generalizations.
\end{obstbox}

\begin{obstbox}
\textbf{Label complexity $\leftrightarrow$ sample complexity.}

\medskip\noindent
Both count the amount of data needed for learning, but label complexity
(from active learning) counts the number of \emph{labels queried},
while sample complexity counts the number of \emph{labeled examples
drawn}. In active learning, the learner sees many unlabeled points but
requests labels selectively, so label complexity can be exponentially
smaller than sample complexity. No equivalence exists; the gap is the
content of active learning theory.
\end{obstbox}

\begin{obstbox}
\textbf{Lifelong learning $\leftrightarrow$ concept drift.}

\medskip\noindent
Both concern learning in non-stationary environments. Lifelong learning
emphasizes knowledge transfer across a sequence of tasks; concept drift
emphasizes adaptation to a changing target within a single task. The
structural parallel is ``the learning problem changes over time,'' but
the formalization differs: lifelong learning uses a task distribution,
concept drift uses a time-varying concept. No formal theorem connects
the two frameworks.
\end{obstbox}


% ================================================================
% SECTION 3: ONE-WAY THEOREM ONLY (4 edges)
% ================================================================

\section{One-Way Theorem Only}
\label{sec:one-way}

In these analogies, one direction of the relationship has been proved as
a theorem, but the reverse direction is either open or known to be
false. The analogy is ``half-true'': the structural parallel exists, but
it is asymmetric.

\begin{obstbox}
\textbf{Sample complexity $\rightarrow$ VC dimension} (but not
$\leftarrow$).

\medskip\noindent
The fundamental theorem establishes: $\VC(\mathcal{H}) < \infty
\Leftrightarrow$ finite sample complexity. The two are functionally
related, but sample complexity is defined independently as the minimum
number of examples needed, while VC dimension is a combinatorial
property. The relationship is a theorem, not a definition; treating
them as interchangeable obscures the logical structure.
\end{obstbox}

\begin{obstbox}
\textbf{Compression scheme $\rightarrow$ sample complexity} (but not
$\leftarrow$).

\medskip\noindent
A compression scheme of size $k$ implies PAC sample complexity
$O(k \cdot \log m / \varepsilon)$ (Littlestone--Warmuth, Moran--Yehudayoff).
But no theorem says that optimal sample complexity determines optimal
compression size. A class might be learnable with few samples yet require
a large compression scheme. The compression direction is proved;
the reverse is open.
\end{obstbox}

\begin{obstbox}
\textbf{Covering number $\rightarrow$ VC dimension} (but not
$\leftarrow$).

\medskip\noindent
The Sauer--Shelah lemma gives: $\log N(\varepsilon, \mathcal{H},
L_1(P)) \leq d \cdot \log(2e/\varepsilon)$, bounding covering numbers
from VC dimension. But covering numbers are defined for arbitrary
metric spaces, while VC dimension applies only to binary hypothesis
classes. VC dimension is a sufficient condition for covering number
bounds, not a necessary one; the covering number framework is strictly
more general.
\end{obstbox}

\begin{obstbox}
\textbf{Covering number $\rightarrow$ Rademacher complexity} (but not
$\leftarrow$).

\medskip\noindent
Dudley's entropy integral bounds Rademacher complexity from above using
covering numbers: $\Rad_n(\mathcal{H}) \leq \inf_{\alpha > 0}
\left[ 4\alpha + \frac{12}{\sqrt{n}} \int_\alpha^1
\sqrt{\log N(\varepsilon, \mathcal{H}, L_2)} \, d\varepsilon \right]$.
This bound can be loose; tighter chaining methods (Talagrand's generic
chaining) and direct Rademacher estimates sometimes give better results.
The Dudley bound is one-directional: covering numbers control
Rademacher, but Rademacher complexity does not determine covering
numbers.
\end{obstbox}


% ================================================================
% SECTION 4: DATA MODEL MISMATCH (3 edges)
% ================================================================

\section{Data Model Mismatch}
\label{sec:data-model}

These analogies fail because the source and target assume incompatible
data-generating processes. The structural parallel is real, but the
mathematical assumptions about how data arrives are fundamentally
different.

\begin{obstbox}
\textbf{Concept drift $\leftrightarrow$ online learning.}

\medskip\noindent
Both are sequential learning settings, but concept drift uses i.i.d.\
draws per time step from a slowly changing distribution, while online
learning uses adversarially chosen instances from a fixed concept class.
The shared structure is ``sequential prediction under non-stationarity,''
but the non-stationarity is stochastic in one case and adversarial in
the other. Results from one setting do not transfer: online regret
bounds assume a fixed comparator class, while drift bounds assume a
rate of distributional change.
\end{obstbox}

\begin{obstbox}
\textbf{Advice reduction $\leftrightarrow$ sample complexity.}

\medskip\noindent
Both quantify ``how much side information shrinks the learner's search
space.'' But advice operates in the Gold model (infinite stream,
computable learner), while sample complexity operates in the PAC model
(finite i.i.d.\ sample, distribution-free guarantees). No formal theorem
bridges the two frameworks. The structural parallel---``side information
helps''---is too vague to formalize without choosing a specific data
model, and the two models are incompatible.

\hfill\cite{KaufmannStephan2001}
\end{obstbox}

\begin{obstbox}
\textbf{Granger causality $\leftrightarrow$ concept drift.}

\medskip\noindent
Both operate on time-series data, but Granger causality assumes
stationarity within estimation windows (testing whether past values
of $X$ improve prediction of $Y$), while concept drift explicitly
models non-stationarity. The structural parallel is ``temporal
dependence in sequential data,'' but the assumptions about the data
process are contradictory: Granger requires stationarity; drift requires
its absence.
\end{obstbox}


% ================================================================
% SECTION 5: PROOF METHOD MISMATCH (3 edges)
% ================================================================

\section{Proof Method Mismatch}
\label{sec:proof-method}

In these analogies, the source and target have the same theorem
structure---the ``shape'' of the result is parallel---but the proof
techniques from one do not transfer to the other.

\begin{obstbox}
\textbf{Information-theoretic bound $\leftrightarrow$ algorithmic
stability.}

\medskip\noindent
Both bound the generalization gap via properties of the learning
algorithm. Stability uses sup-norm perturbation sensitivity (how much
does the output change when one training point is replaced?).
Information-theoretic bounds use mutual information $I(W; S)$ (how much
does the algorithm's output depend on the training set?). These are
connected via the chain: low stability $\Rightarrow$ low $I(W; S)$
$\Rightarrow$ low generalization gap. But the proof methods are
fundamentally different: stability arguments use coupling and
McDiarmid-type concentration; information-theoretic arguments use
change-of-measure and data processing inequalities. Neither proof
technique subsumes the other.
\end{obstbox}

\begin{obstbox}
\textbf{Ordinal VC dimension $\leftrightarrow$ mind-change ordinal.}

\medskip\noindent
Both use ordinal-valued measures to characterize learnability in
transfinite settings. Ordinal VC dimension (Martin--Sharma--Stephan)
characterizes predictive complexity; mind-change ordinals characterize
identification complexity in the Gold model. Both use ordinals, but they
measure different things. The connection is mediated through predictive
complexity, not direct: ordinal VC dimension governs prediction error
bounds, while mind-change ordinals govern convergence to a correct
hypothesis. The proof techniques (Ramsey-theoretic for ordinal VC,
topological for mind-change) do not transfer.
\end{obstbox}

\begin{obstbox}
\textbf{Occam's razor $\leftrightarrow$ MDL.}

\medskip\noindent
Both assert that simpler models generalize better. Occam's razor (in the
PAC sense of Blumer et al.) is a worst-case generalization theorem:
if the hypothesis is short, its error is low with high probability.
MDL is an information-theoretic model selection principle: minimize the
total description length of model plus data-given-model. The theorem
\emph{structure} is parallel (``short $\Rightarrow$ good''), but Occam
gives frequentist, distribution-free guarantees while MDL gives
Bayesian-flavored, coding-theoretic guarantees. The proof methods
inhabit different mathematical worlds.

\hfill\cite{BlumerEhrenfeuchtHausslerWarmuth1987}
\end{obstbox}


% ================================================================
% SECTION 6: SUCCESS CRITERION MISMATCH (1 edge)
% ================================================================

\section{Success Criterion Mismatch}
\label{sec:criterion-mismatch}

A single analogy edge has this obstruction type.

\begin{obstbox}
\textbf{Granger causality $\leftrightarrow$ online learning.}

\medskip\noindent
Granger causality tests causal structure: does the past of time series
$X$ improve prediction of time series $Y$ beyond what $Y$'s own past
provides? Online learning minimizes cumulative prediction error against
an adversary. Both involve sequential prediction, but they optimize
different objectives. Granger seeks a \emph{structural} conclusion
(causal influence exists or does not); online learning seeks a
\emph{performance} guarantee (low regret). The success criteria are
fundamentally different: statistical significance of a causal test vs.\
sub-linear regret growth. No formal theorem connects them because they
answer different questions about the same sequential data.
\end{obstbox}


% ================================================================
% SECTION 7: THE OBSTRUCTION MAP
% ================================================================

\section{The Obstruction Map}
\label{sec:obstruction-map}

\Cref{fig:obstruction-map} displays all 32 analogy edges, colored by
obstruction type. The visual pattern reveals the structure of the field's
unresolved connections.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
  node distance=1.2cm and 1.6cm,
  concept/.style={
    rectangle, rounded corners=2pt, draw, thick,
    fill=layergray, font=\scriptsize\sffamily,
    minimum width=1.6cm, minimum height=0.5cm,
    align=center, inner sep=2pt
  },
  tm/.style={-, edgeorange, thick},        % type mismatch
  mew/.style={-, defblue, thick, dashed},   % missing equiv witness
  owt/.style={-, sepgreen, thick, dotted},  % one-way theorem
  dm/.style={-, thmred, thick, dashdotted}, % data model mismatch
  pm/.style={-, analogypurple, thick, densely dashed}, % proof method
  sc/.style={-, boundbrown, thick, densely dotted},    % success criterion
]

% --- Complexity measures cluster ---
\node[concept] (vc) {VC dim};
\node[concept, right=1.4cm of vc] (rad) {Rademacher};
\node[concept, right=1.4cm of rad] (cov) {Covering\\number};
\node[concept, below=1.0cm of vc] (sq) {SQ dim};
\node[concept, below=1.0cm of rad] (comp) {Compression};
\node[concept, above=1.0cm of vc] (ldim) {Littlestone\\dim};
\node[concept, above=1.0cm of rad] (mb) {Mistake\\bound};
\node[concept, right=1.4cm of cov] (sc_node) {Sample\\complexity};
\node[concept, below=1.0cm of sc_node] (lc) {Label\\complexity};

% --- Paradigm cluster ---
\node[concept, below=2.8cm of vc] (pac) {PAC};
\node[concept, right=1.4cm of pac] (stab) {Stability};
\node[concept, right=1.4cm of stab] (pb) {PAC-Bayes};
\node[concept, right=1.4cm of pb] (itb) {Info-theoretic};
\node[concept, below=1.0cm of pac] (post) {Posterior\\consistency};
\node[concept, left=1.4cm of pac] (ft) {Fund.\\Theorem};

% --- Characterization cluster ---
\node[concept, above left=1.0cm and 0.0cm of ldim] (vcc) {VC charac.};
\node[concept, above right=1.0cm and 0.0cm of ldim] (lc_char) {Ldim charac.};

% --- Model selection cluster ---
\node[concept, below=2.8cm of stab] (mdl) {MDL};
\node[concept, left=1.4cm of mdl] (mml) {MML};
\node[concept, right=1.4cm of mdl] (bayes) {Bayesian};
\node[concept, left=1.4cm of mml] (kolm) {Kolmogorov};
\node[concept, below=1.0cm of mdl] (occam) {Occam};
\node[concept, right=1.4cm of bayes] (srm) {SRM};
\node[concept, below=1.0cm of bayes] (ib) {Inductive\\bias};

% --- Sequential cluster ---
\node[concept, below=2.8cm of pb] (ol) {Online};
\node[concept, right=1.4cm of ol] (drift) {Concept\\drift};
\node[concept, right=1.4cm of drift] (gc) {Granger};
\node[concept, below=1.0cm of drift] (exd) {Ex under\\drift};
\node[concept, below=1.0cm of ol] (ll) {Lifelong};
\node[concept, below left=1.0cm and 0.0cm of ll] (ml) {Meta-\\learner};

% --- Data cluster ---
\node[concept, left=1.4cm of kolm] (bk) {Background\\knowledge};
\node[concept, below=1.0cm of kolm] (adv) {Advice};
\node[concept, below=1.0cm of bk] (gi) {Grammar\\induction};
\node[concept, below=1.0cm of gi] (gt) {Gold's\\theorem};
\node[concept, below=1.0cm of adv] (ar) {Advice\\reduction};

% --- Type mismatch edges (orange, 12) ---
\draw[tm] (stab) -- (pb);
\draw[tm] (itb) -- (pb);
\draw[tm] (ft) -- (stab);
\draw[tm] (sq) -- (vc);
\draw[tm] (gi) -- (gt);
\draw[tm] (drift) -- (exd);
\draw[tm] (ll) -- (ml);
\draw[tm] (bk) -- (adv);
\draw[tm] (mdl) -- (kolm);
\draw[tm] (srm) -- (ib);
\draw[tm] (bayes) -- (ib);
\draw[tm] (occam) -- (comp);

% --- Missing equivalence witness edges (blue dashed, 9) ---
\draw[mew] (comp) -- (vc);
\draw[mew] (rad) -- (vc);
\draw[mew] (vcc) -- (lc_char);
\draw[mew] (mml) -- (mdl);
\draw[mew] (mml) -- (bayes);
\draw[mew] (pac) -- (post);
\draw[mew] (mb) -- (ldim);
\draw[mew] (lc) -- (sc_node);
\draw[mew] (ll) -- (drift);

% --- One-way theorem edges (green dotted, 4) ---
\draw[owt] (sc_node) -- (vc);
\draw[owt] (comp) -- (sc_node);
\draw[owt] (cov) -- (vc);
\draw[owt] (cov) -- (rad);

% --- Data model mismatch edges (red dashdotted, 3) ---
\draw[dm] (drift) -- (ol);
\draw[dm] (ar) -- (sc_node);
\draw[dm] (gc) -- (drift);

% --- Proof method mismatch edges (purple, 3) ---
\draw[pm] (itb) -- (stab);
\draw[pm] (occam) -- (mdl);
% ordinal_vc_dim -> mind_change_ordinal not shown (nodes would clutter)

% --- Success criterion mismatch (brown, 1) ---
\draw[sc] (gc) -- (ol);

\end{tikzpicture}

\caption{The obstruction map: all 32 analogy edges colored by
obstruction type.
\textcolor{edgeorange}{\textbf{---}} type mismatch (12);
\textcolor{defblue}{\textbf{-\,-\,-}} missing equivalence witness (9);
\textcolor{sepgreen}{\textbf{$\cdots$}} one-way theorem (4);
\textcolor{thmred}{\textbf{-$\cdot$-}} data model mismatch (3);
\textcolor{analogypurple}{\textbf{-\,-}} proof method mismatch (3);
\textcolor{boundbrown}{\textbf{$\cdots$}} success criterion mismatch (1).
One edge (ordinal VC dim $\leftrightarrow$ mind-change ordinal,
proof method mismatch) is omitted from the diagram for clarity.}
\label{fig:obstruction-map}
\end{figure}

\subsection*{What the distribution tells us}

The obstruction type distribution reveals three structural facts about
formal learning theory.

\emph{First}, the dominance of type mismatches (12 of 32) means that
the field's most common cross-paradigm analogies fail for the most
basic possible reason: the objects being compared are not the same kind
of thing. This is not a failure of proof technique; it is a failure of
type compatibility. It suggests that future unification results will
require new bridging concepts that mediate between the existing type
systems.

\emph{Second}, the 9 missing equivalence witnesses represent the field's
open frontier. The sample compression conjecture (compression
$\leftrightarrow$ VC dimension) is the most prominent, but the
Rademacher--VC relationship and the characterization-theorem parallel
(VC $\leftrightarrow$ Littlestone) are equally fundamental. Progress on
any of these would restructure the analogy map, converting a dashed blue
edge into a solid theorem edge.

\emph{Third}, the 4 one-way theorems show where asymmetry is
structural, not accidental. Covering numbers bound Rademacher complexity
(Dudley) and are bounded by VC dimension (Sauer--Shelah), but these
bounds flow in one direction only. The asymmetry reflects the
generality hierarchy: covering numbers apply to metric spaces,
Rademacher to function classes, VC dimension to binary classes. Each
step downward in generality tightens the bound but loses applicability.

Together, the 32 analogy edges and their obstructions form a map of what
formal learning theory does \emph{not yet} know how to unify. The
obstructions are not obstacles to be removed; most of them are permanent
features of the mathematical landscape. Understanding why an analogy
fails is, in this field, as valuable as understanding why a theorem
holds.
