% Chapter 16: Computational vs. Information-Theoretic Learnability
% Centerpiece: the Kearns-Valiant 1994 gap — classes with finite VC dim
% that are NOT efficiently learnable (under cryptographic assumptions).
% Concepts: computational_hardness, proper_improper_separation, requires_assumption edges.
% SHORT chapter (~3-4 pages).

\chapter{Computational vs.\ Information-Theoretic Learnability}
\label{ch:computational}

The Fundamental Theorem of \Cref{ch:pac} characterizes learnability in
information-theoretic terms: $\mathcal{H}$ is PAC learnable if and only if
$\VC(\mathcal{H}) < \infty$.  The characterization is silent about
\emph{computation}.  It guarantees the \emph{existence} of a sample size
$m(\varepsilon, \delta)$ and a learner $A$ that achieves accuracy $\varepsilon$
with confidence $1 - \delta$, but it says nothing about how long $A$ takes to
run.

This silence conceals a gap.  There exist hypothesis classes with finite VC
dimension---hence information-theoretically learnable---that no polynomial-time
algorithm can PAC learn, unless widely believed cryptographic assumptions are
false.  The gap between what is \emph{learnable} and what is \emph{efficiently
learnable} is the subject of this chapter.  It is a negative result, but a
structurally important one: it shows that information-theoretic
characterizations, however elegant, do not tell the whole story.

\medskip

The chapter has three sections.

\begin{enumerate}[leftmargin=*,itemsep=2pt]
\item \textbf{The information--computation gap} (\Cref{sec:gap}): the
  Kearns--Valiant result and the cryptographic assumptions it requires.
\item \textbf{Proper vs.\ improper learning} (\Cref{sec:proper-improper}): the
  separation between proper and improper learnability, and why the
  representation matters.
\item \textbf{The \rel{requires\_assumption} edges}
  (\Cref{sec:requires-assumption}): what this edge type means structurally in
  the concept graph.
\end{enumerate}


% ================================================================
% SECTION 1: THE INFORMATION--COMPUTATION GAP
% ================================================================

\section{The Information--Computation Gap}
\label{sec:gap}

\begin{graphnode}[title={Graph Node: \nde{computational\_hardness}}]
Layer: \texttt{meta}.  Incoming edges:
$\edge{computational\_hardness}{requires\_assumption}{inductive\_bias}$.
This node represents the structural fact that efficient learnability requires
assumptions beyond finite VC dimension.
\end{graphnode}

The central result is due to Kearns and Valiant~\cite{KearnsValiant1994}.

\begin{thm}[Kearns--Valiant 1994]
\label{thm:kearns-valiant}
Under the assumption that certain cryptographic problems are hard (specifically,
that the decisional composite residuosity assumption, the hardness of factoring,
or the learning with errors assumption holds), there exist concept classes
$\mathcal{C}$ with:
\begin{enumerate}[leftmargin=*,nosep]
\item $\VC(\mathcal{C}) < \infty$, so that $\mathcal{C}$ is PAC learnable
  (information-theoretically), but
\item no polynomial-time algorithm PAC learns $\mathcal{C}$.
\end{enumerate}
\end{thm}

The original construction uses Boolean formulae: the class of polynomial-size
Boolean formulae has VC dimension polynomial in the number of variables, yet
PAC learning this class is at least as hard as breaking certain public-key
cryptosystems.  The reduction goes through the construction of
\emph{cryptographic pseudorandom function families}: if a learner could
efficiently distinguish the target concept from random, it could break the
cryptographic assumption.

\begin{remark}[The structure of the reduction]
\label{rmk:kv-reduction}
The Kearns--Valiant reduction is from a cryptographic primitive to a learning
problem, not from a worst-case problem.  The learner is given random examples
$(x, c(x))$ where $x$ is drawn from a distribution and $c$ is the target
concept.  If the learner succeeds in learning, it implicitly solves the
cryptographic problem on random instances.  This is why the hardness is
\emph{average-case}, not worst-case.
\end{remark}

\subsection{Why P $\neq$ NP Does Not Suffice}

One might hope that the information--computation gap follows from
$\mathrm{P} \neq \mathrm{NP}$.  It does not.  Applebaum, Barak, and
Xiao~\cite{ApplebaumBarakXiao2008} showed the following.

\begin{thm}[Applebaum--Barak--Xiao 2008]
\label{thm:abx}
There is no \emph{black-box} reduction from $\mathrm{NP}$-hardness to the
hardness of PAC learning.  Specifically, if PAC learning a class $\mathcal{C}$
is hard, this hardness cannot be established by a reduction that treats the
learner as an oracle, unless $\mathrm{NP} \subseteq \mathrm{coAM}$.
\end{thm}

The reason is structural.  PAC learning is an \emph{average-case} problem: the
learner receives random examples from a distribution and must generalize.
$\mathrm{NP}$-hardness is a \emph{worst-case} notion.  Reductions from
worst-case to average-case problems are notoriously difficult and, in many
settings, provably impossible via relativizing techniques.

\subsection{The Necessary Assumptions}

The assumptions used in the literature fall into three families, in
roughly increasing order of generality:

\begin{enumerate}[leftmargin=*,itemsep=3pt]
\item \textbf{Factoring and RSA.}  The original Kearns--Valiant result
  assumes that factoring $n$-bit integers requires super-polynomial time.
  This is the most classical assumption but also the most fragile: it is
  broken by quantum computers (Shor's algorithm).

\item \textbf{Decisional Composite Residuosity (DCRA).}  Used in subsequent
  refinements, DCRA assumes that it is hard to distinguish $n$th residues
  modulo $n^2$.  Like factoring, this is vulnerable to quantum attacks.

\item \textbf{Learning With Errors (LWE).}  Introduced by
  Regev~\cite{Regev2005}, the LWE assumption states that it is hard to recover
  a secret vector $\mathbf{s}$ from noisy inner products
  $\langle \mathbf{a}_i, \mathbf{s} \rangle + e_i \pmod{q}$.  LWE is
  believed to resist quantum computers and has become the standard assumption
  in modern hardness-of-learning results~\cite{Daniely2021}.
\end{enumerate}

\begin{obstbox}
The information--computation gap is an \emph{obstruction} in the concept
graph: the edge
$\edge{vc\_dimension}{characterizes}{pac\_learnability}$ is
information-theoretically valid, but the analogous computational edge does
not exist.  No combinatorial dimension characterizes efficient PAC
learnability.  This is the deepest structural gap in the theory.
\end{obstbox}


% ================================================================
% SECTION 2: PROPER VS. IMPROPER LEARNING
% ================================================================

\section{Proper vs.\ Improper Learning}
\label{sec:proper-improper}

The information--computation gap has a subtler cousin: the gap between
\emph{proper} and \emph{improper} learning.

\begin{defn}[Proper and Improper Learning]
\label{def:proper-improper}
A PAC learner for $\mathcal{H}$ is \emph{proper} if it always outputs a
hypothesis $h \in \mathcal{H}$.  It is \emph{improper} if it may output any
efficiently evaluable function $h$, not necessarily in $\mathcal{H}$.
\end{defn}

The distinction matters because the \emph{representation} of hypotheses affects
computational complexity.

\begin{thm}[Proper--Improper Separation]
\label{thm:proper-improper}
There exist concept classes $\mathcal{C}$ such that:
\begin{enumerate}[leftmargin=*,nosep]
\item proper PAC learning of $\mathcal{C}$ is $\mathrm{NP}$-hard, but
\item there exists a polynomial-time improper learner for $\mathcal{C}$.
\end{enumerate}
\end{thm}

The canonical example is the class of $3$-term DNF formulae.  Finding a
consistent $3$-term DNF is $\mathrm{NP}$-hard~\cite{PittValiant1988}, but the
class of $3$-term DNFs is contained in the class of $3$-CNF formulae, which
\emph{is} efficiently learnable.  An improper learner simply outputs a $3$-CNF
formula that is consistent with the data.  The hypothesis is not a DNF, but it
classifies correctly.

\begin{remark}[Representation vs.\ class]
\label{rmk:representation}
The proper--improper separation illustrates a principle that runs throughout
computational learning theory: what matters for computational complexity is not
the \emph{set of functions} $\mathcal{H} \subseteq \{0,1\}^X$ but the
\emph{representation} of those functions.  Two representations of the same set
of Boolean functions can have dramatically different computational properties.
This is why the VC dimension---which depends only on the set
$\mathcal{H}$---cannot capture computational learnability.
\end{remark}

The structural consequence is that efficient learnability is a property of
\emph{represented} hypothesis classes, not of hypothesis classes as sets.  The
concept graph captures this through the node \nde{proper\_improper\_separation},
which has type \texttt{separation} and connects to
\nde{computational\_hardness} via a \rel{motivates} edge.


% ================================================================
% SECTION 3: THE REQUIRES_ASSUMPTION EDGES
% ================================================================

\section{The \rel{requires\_assumption} Edges}
\label{sec:requires-assumption}

The concept graph contains several edges of type \rel{requires\_assumption}.
These edges encode a specific logical structure: the source node's validity
depends on an assumption that is not provable from the axioms of the theory
alone.

\begin{defn}[The \rel{requires\_assumption} edge type]
\label{def:requires-assumption}
An edge $\edge{A}{requires\_assumption}{B}$ means: the result or property
associated with node $A$ holds \emph{conditional on} the assumption encoded
by node $B$.  If $B$ is falsified, the status of $A$ becomes indeterminate.
\end{defn}

The principal instances in the graph are:

\begin{enumerate}[leftmargin=*,itemsep=3pt]
\item $\edge{computational\_hardness}{requires\_assumption}{inductive\_bias}$.
  The information--computation gap of \Cref{thm:kearns-valiant} holds only
  under cryptographic assumptions.  Without them, it remains conceivable that
  every class with finite VC dimension is efficiently learnable.  The
  \nde{inductive\_bias} node here encodes the structural assumption that the
  learner's hypothesis space is constrained---without which the hardness
  results are vacuous.

\item $\edge{posterior\_consistency}{requires\_assumption}{bayesian\_inference}$.
  Posterior consistency (\Cref{ch:agents}, \Cref{sec:bayesian-pac}) holds under
  conditions on the prior and the model class.  When these conditions
  fail---for example, when the prior assigns zero mass to neighborhoods of
  the true parameter---posterior consistency fails
  (Diaconis--Freedman~\cite{DiaconisFreedman1986}).  The edge records this
  conditional dependence.
\end{enumerate}

\begin{remark}[Structural meaning]
\label{rmk:requires-assumption-structure}
The \rel{requires\_assumption} edge type is the graph's mechanism for
representing \emph{conditional} theorems.  Unlike \rel{implies} edges, which
are unconditional, a \rel{requires\_assumption} edge signals that the source
node's content is only as secure as the target node's assumption.  This is
the graph-theoretic encoding of the phrase ``under the assumption that\ldots''
that pervades computational learning theory.
\end{remark}

\bigskip

This chapter has established the structural gap between
information-theoretic and computational learnability.  The gap is real---it
produces classes that are provably learnable but (conditionally) not
efficiently learnable.  It is also irreducible: no known combinatorial
dimension characterizes efficient PAC learnability.  The next chapter
examines a different kind of gap: the distance between binary classification
and the richer learning settings that require new mathematical primitives.
