% Chapter 17: Extensions Beyond Binary Classification
%   ds_dimension — Profile C (Revelation): The 30-year gap from Natarajan to DS
%   multiclass_learning — Profile G (Framework Bridge): Multiple paradigms tried
%   agnostic_learning — Profile B-light: The agnostic fundamental theorem
%   proper_improper_separation — Profile D (Witness Construction)
% Theme: GRAMMAR GROWTH — each extension requires new primitives.
% The extends_grammar edge type is the structural backbone.

\chapter{Extensions Beyond Binary Classification}
\label{ch:extensions}

The PAC theory of \Cref{ch:pac} and its characterization through VC dimension
apply to binary classification: the label set is $\{0,1\}$, the hypothesis
class is $\mathcal{H} \subseteq \{0,1\}^X$, and the loss is $0$-$1$.  This is
the setting where the theory is cleanest and most complete.  But most learning
problems are not binary classification.  Labels may come from a finite set of
$k > 2$ classes; the target may be a real-valued function; the data may be
noisy; the realizability assumption may fail.

This chapter asks: how does the binary theory extend to these richer settings?
The answer is not ``straightforwardly.''  Each extension requires
\emph{new mathematical primitives}---new notions of shattering, new complexity
measures, new assumptions.  The binary theory does not generalize by
substitution; it generalizes by \emph{grammar growth}.  In the concept graph,
this is captured by edges of type \rel{extends\_grammar}: the source dimension
extends the grammar of the VC dimension by introducing primitives that have no
analogue in the binary case.

\medskip

The chapter has six sections.

\begin{enumerate}[leftmargin=*,itemsep=2pt]
\item \textbf{Multiclass learning} (\Cref{sec:multiclass}): the Natarajan
  dimension, the DS dimension, and the resolution of a thirty-year open problem.
\item \textbf{Real-valued functions} (\Cref{sec:real-valued}):
  pseudodimension and fat-shattering dimension.
\item \textbf{Agnostic learning} (\Cref{sec:agnostic}): dropping realizability
  and the agnostic fundamental theorem.
\item \textbf{Noise-tolerant and partial concept learning}
  (\Cref{sec:noise-partial}): classification noise, malicious noise,
  and promise problems.
\item \textbf{Proper versus improper learning}
  (\Cref{sec:proper-improper}): the computational gap.
\item \textbf{The \rel{extends\_grammar} pattern}
  (\Cref{sec:extends-grammar}): what the extensions have in common.
\end{enumerate}


% ================================================================
%     CONCEPT MAP FIGURE
% ================================================================

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
  node distance=1.6cm and 2.8cm,
  concept/.style={
    rectangle, rounded corners=3pt, draw, thick,
    fill=layergray, font=\small\sffamily,
    minimum width=2.4cm, minimum height=0.7cm,
    align=center
  },
  ext/.style={-{Stealth[length=5pt]}, edgeorange, thick},
  chr/.style={-{Stealth[length=5pt]}, defblue, thick},
  ss/.style={-{Stealth[length=5pt]}, thmred, thick, dashed},
  lbl/.style={font=\tiny\itshape, text=notegray, align=center}
]

% --- Binary core ---
\node[concept] (vc) {VC\\Dimension};
\node[concept, right=3.2cm of vc] (pac) {PAC\\Learning};

% --- Multiclass branch ---
\node[concept, above right=1.8cm and 0.8cm of vc] (nat) {Natarajan\\Dimension};
\node[concept, above=1.4cm of nat] (ds) {DS\\Dimension};

% --- Real-valued branch ---
\node[concept, below right=1.8cm and 0.8cm of vc] (pdim) {Pseudo-\\dimension};
\node[concept, below=1.4cm of pdim] (fat) {Fat-Shattering\\Dimension};

% --- Agnostic ---
\node[concept, right=3.2cm of pac] (agn) {Agnostic\\PAC};

% --- Proper/Improper ---
\node[concept, below=2.0cm of pac] (pi) {Proper vs\\Improper};

% --- Noise ---
\node[concept, above=2.0cm of pac] (noise) {Noise-Tolerant\\Learning};

% --- Edges ---
\draw[ext] (nat) -- node[lbl, left] {\rel{extends\_grammar}} (vc);
\draw[ext] (pdim) -- node[lbl, left] {\rel{extends\_grammar}} (vc);
\draw[ss] (ds) -- node[lbl, left] {\rel{strictly\_stronger}} (nat);
\draw[chr] (vc) -- node[lbl, above] {\rel{characterizes}} (pac);
\draw[chr] (ds) to[bend left=20] node[lbl, above right] {\rel{characterizes}} (pac);
\draw[chr] (fat) to[bend right=20] node[lbl, below right] {\rel{characterizes}} (pac);
\draw[chr] (vc) to[bend right=15] node[lbl, below] {\rel{characterizes}} (agn);

\draw[-{Stealth[length=5pt]}, sepgreen, thick]
  (agn) -- node[lbl, above] {\rel{restricts}} (pac);
\draw[-{Stealth[length=5pt]}, boundbrown, thick]
  (pi) -- node[lbl, right] {\rel{lower\_bounds}} (pac);

\end{tikzpicture}

\caption{Concept map for this chapter.  Orange:~\rel{extends\_grammar}.
Dashed red:~\rel{strictly\_stronger}.  Blue:~\rel{characterizes}.
Green:~\rel{restricts}.  Brown:~\rel{lower\_bounds}.  Each arrow
represents an edge in the concept graph.}
\label{fig:ch17-concept-map}
\end{figure}


% ================================================================
% SECTION 1: MULTICLASS LEARNING
% ================================================================

\section{Multiclass Learning: From Natarajan to DS}
\label{sec:multiclass}

Let $Y = [k] = \{1, 2, \ldots, k\}$ for some $k \geq 2$, and let
$\mathcal{H} \subseteq Y^X$ be a multiclass hypothesis class.  The question
is: what characterizes PAC learnability of $\mathcal{H}$?

For $k = 2$, the answer is the VC dimension.  For $k > 2$, the answer was only
fully resolved in 2022, after a thirty-year journey through two competing
dimensions, each introducing a new shattering primitive.  That journey---and its
resolution---is the narrative backbone of this section.


\subsection{The Natarajan Dimension}

Natarajan~\cite{Natarajan1989} introduced the first multiclass complexity
measure by generalizing shattering to require two functions instead of one.

\begin{defn}[Natarajan Dimension {\cite{Natarajan1989}}]
\label{def:natarajan-dim}
The \emph{Natarajan dimension} $\Ndim(\mathcal{H})$ of a multiclass hypothesis
class $\mathcal{H} \subseteq Y^X$ is the largest size $d$ of a set
$\{x_1, \ldots, x_d\} \subseteq X$ for which there exist two functions
$f_0, f_1 \colon \{x_1, \ldots, x_d\} \to Y$ satisfying:
\begin{enumerate}[leftmargin=*,nosep]
\item $f_0(x_i) \neq f_1(x_i)$ for all $i \in [d]$, and
\item for every binary string $b \in \{0,1\}^d$, there exists $h \in
  \mathcal{H}$ with $h(x_i) = f_{b_i}(x_i)$ for all $i$.
\end{enumerate}
\end{defn}

The new primitive is the \emph{pair of functions} $(f_0, f_1)$.  In binary
classification, $f_0$ and $f_1$ are forced to be the constant functions $0$
and $1$, and the definition reduces to VC shattering.  In the multiclass case,
the choice of which two labels to use at each point is part of the
combinatorial structure.

\begin{example}[Natarajan shattering with $k=3$]
\label{ex:natarajan-k3}
Let $X = \{a, b\}$, $Y = \{1,2,3\}$.  Define $f_0(a)=1$, $f_0(b)=2$
and $f_1(a)=3$, $f_1(b)=1$.  To N-shatter $\{a,b\}$, we need four
hypotheses realizing every binary string:
\[
  h_{00}(a)=1,\, h_{00}(b)=2;\quad
  h_{01}(a)=1,\, h_{01}(b)=1;\quad
  h_{10}(a)=3,\, h_{10}(b)=2;\quad
  h_{11}(a)=3,\, h_{11}(b)=1.
\]
Any $\mathcal{H} \supseteq \{h_{00}, h_{01}, h_{10}, h_{11}\}$ has $\Ndim \geq 2$.
\end{example}

\begin{historical}
\textbf{The Natarajan gap (1989--2022).}
Natarajan showed that $\Ndim(\mathcal{H}) < \infty$ is \emph{sufficient}
for PAC learnability and provided sample complexity bounds.  The best
known bounds, due to Ben-David et al.~\cite{BenDavidCesaBianchiHausslerLong1995},
were:
\[
  C_1 \cdot \frac{\Ndim}{\varepsilon}
  \;\leq\; m_{\mathrm{PAC}}(\varepsilon,\delta)
  \;\leq\; C_2 \cdot \frac{\Ndim \cdot \ln k \cdot \ln(1/\varepsilon)}{\varepsilon}.
\]
Two gaps persisted.  First, the $\Theta(\log k)$ factor between upper
and lower bounds.  Second, and more fundamentally: is finiteness of $\Ndim$
also \emph{necessary}?  For finite label sets $|Y| = k < \infty$, the answer
is yes (the $\log k$ factor is at most a polynomial overhead).  But when
$|Y| = \infty$, a class can have $\Ndim = 1$ and yet fail to be PAC learnable.
This was the open problem that persisted for over thirty years.
\end{historical}


\begin{graphnode}[title={Graph Node: \nde{natarajan\_dimension}}]
Layer~5: \texttt{complexity\_measure}.  Key edge:
$\edge{natarajan\_dimension}{extends\_grammar}{vc\_dimension}$.
The grammar extension is the two-function shattering primitive $(f_0, f_1)$.
Open frontier: tight multiclass sample complexity without the $\log k$ factor.
\end{graphnode}


\subsection{The DS Dimension: A Thirty-Year Resolution}
\label{subsec:ds-dim}

The correct characterization of multiclass PAC learnability was established
by Brukhim et al.~\cite{BrukhimCarmonDinurMoranYehudayoff2022} using the
\emph{DS dimension}, named after Daniely and
Shalev-Shwartz~\cite{DanielyShalevShwartz2014} who introduced the
underlying concept of pseudo-cubes in one-inclusion hypergraphs.

\begin{defn}[One-Inclusion Hypergraph]
\label{def:oig}
Let $\mathcal{H} \subseteq Y^X$ and let $S = \{x_1, \ldots, x_n\} \subseteq X$
be a finite set.  The \emph{one-inclusion hypergraph} $\mathrm{OIG}(\mathcal{H}, S)$
has vertex set $\mathcal{H}|_S = \{(h(x_1), \ldots, h(x_n)) : h \in \mathcal{H}\}$
and, for each coordinate $i \in [n]$, a hyperedge connecting all vertices that
agree on all coordinates except possibly coordinate $i$.
\end{defn}

\begin{defn}[DS Dimension {\cite{DanielyShalevShwartz2014,BrukhimCarmonDinurMoranYehudayoff2022}}]
\label{def:ds-dim}
A set $\{x_1, \ldots, x_d\} \subseteq X$ is \emph{DS-shattered} by
$\mathcal{H} \subseteq Y^X$ if there exists a function
$f \colon \{x_1, \ldots, x_d\} \to Y$ such that for every $i \in [d]$,
there exists $h_i \in \mathcal{H}$ with:
\begin{enumerate}[leftmargin=*,nosep]
\item $h_i(x_i) \neq f(x_i)$, and
\item $h_i(x_j) = f(x_j)$ for all $j \neq i$.
\end{enumerate}
The \emph{DS dimension} $\DSdim(\mathcal{H})$ is the size of the largest
DS-shattered set.
\end{defn}

The structural content of DS-shattering is different from Natarajan shattering
in a revealing way.  Natarajan shattering asks for a \emph{global} pair of
labeling functions with \emph{all} $2^d$ interpolations realized.
DS-shattering asks for a single ``default'' labeling $f$ and, for each
point, a \emph{local} witness that deviates at exactly that point.  The
requirement is weaker per point (only one deviation, not arbitrary binary
combinations) but more structural: it encodes a property of the
one-inclusion hypergraph of $\mathcal{H}$.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
  scale=0.85,
  point/.style={circle, fill=defblue, inner sep=1.8pt},
  hyp/.style={font=\small},
  brace/.style={decorate, decoration={brace, amplitude=5pt}}
]
% Default labeling
\node[hyp] at (-1.5, 2.0) {Default $f$:};
\foreach \i/\lbl in {1/a, 2/b, 3/c} {
  \node[point] (x\i) at (1.5*\i, 2.0) {};
  \node[above=3pt] at (x\i) {$x_{\i}$};
  \node[below=5pt, font=\small, text=defblue] at (x\i) {$\lbl$};
}

% Witness hypotheses
\node[hyp] at (-1.5, 0.6) {$h_1$:};
\foreach \i/\lbl/\col in {1/{\neq a}/thmred, 2/b/defblue, 3/c/defblue} {
  \node[circle, fill=\col!20, draw=\col, inner sep=1.8pt] (h1\i) at (1.5*\i, 0.6) {};
  \node[below=5pt, font=\small, text=\col] at (h1\i) {$\lbl$};
}

\node[hyp] at (-1.5, -0.4) {$h_2$:};
\foreach \i/\lbl/\col in {1/a/defblue, 2/{\neq b}/thmred, 3/c/defblue} {
  \node[circle, fill=\col!20, draw=\col, inner sep=1.8pt] (h2\i) at (1.5*\i, -0.4) {};
  \node[below=5pt, font=\small, text=\col] at (h2\i) {$\lbl$};
}

\node[hyp] at (-1.5, -1.4) {$h_3$:};
\foreach \i/\lbl/\col in {1/a/defblue, 2/b/defblue, 3/{\neq c}/thmred} {
  \node[circle, fill=\col!20, draw=\col, inner sep=1.8pt] (h3\i) at (1.5*\i, -1.4) {};
  \node[below=5pt, font=\small, text=\col] at (h3\i) {$\lbl$};
}

% Annotations
\draw[brace] (5.3, 0.9) -- (5.3, -1.7)
  node[midway, right=6pt, font=\small, align=left]
  {Each $h_i$ deviates\\from $f$ at $x_i$ only};

\end{tikzpicture}
\caption{DS-shattering of $\{x_1, x_2, x_3\}$.  The default labeling $f$
assigns labels $a, b, c$.  Each witness $h_i$ disagrees with $f$ at
exactly one coordinate (shown in red).}
\label{fig:ds-shattering}
\end{figure}


\begin{thm}[Multiclass Characterization {\cite{BrukhimCarmonDinurMoranYehudayoff2022}}]
\label{thm:multiclass-char}
A multiclass hypothesis class $\mathcal{H} \subseteq Y^X$ is PAC learnable if
and only if $\DSdim(\mathcal{H}) < \infty$.
\end{thm}

\begin{proof}[Proof sketch]
We outline both directions.

\medskip\noindent
\textbf{Sufficiency} ($\DSdim(\mathcal{H}) < \infty \Rightarrow$ learnable).
The proof proceeds through the one-inclusion hypergraph.  For a finite
restriction $\mathcal{H}|_S$ to a sample $S$ of size $m$, the one-inclusion
hypergraph $\mathrm{OIG}(\mathcal{H}, S)$ has vertices $\mathcal{H}|_S$ and
hyperedges indexed by coordinates.  The key insight of Daniely and
Shalev-Shwartz is that a \emph{proper orientation} of this hypergraph---an
assignment of each hyperedge to one of its vertices---yields a learning
algorithm.  Specifically:

\begin{enumerate}[leftmargin=*,nosep]
\item Given a sample $S = \{(x_1, y_1), \ldots, (x_m, y_m)\}$, restrict
  $\mathcal{H}$ to $\{x_1, \ldots, x_m\}$ and build $\mathrm{OIG}(\mathcal{H}, S)$.
\item Find a proper orientation with \emph{minimum outdegree}.
\item Return the hypothesis whose vertex has minimum outdegree in the
  orientation.
\end{enumerate}

If $\DSdim(\mathcal{H}) = d < \infty$, then the one-inclusion hypergraph
of any restriction has no $d$-dimensional pseudo-cube, and a combinatorial
argument (via the Helly property of CAT(0) cube complexes) shows that
a proper orientation with outdegree at most $O(d)$ exists.  The learner's
error is bounded by $O(d/m)$, yielding PAC learnability with sample
complexity $O(d/\varepsilon)$.

\medskip\noindent
\textbf{Necessity} ($\DSdim(\mathcal{H}) = \infty \Rightarrow$ not learnable).
If $\DSdim(\mathcal{H}) = \infty$, then for every $d$, there exists a
finite set that is DS-shattered.  The proof constructs, for each $d$, a
distribution $D_d$ over $X \times Y$ such that:
\begin{itemize}[leftmargin=*,nosep]
\item The Bayes-optimal classifier has zero error.
\item Any learner given $m < d$ samples has expected error
  $\Omega(d/m)$ on $D_d$.
\end{itemize}
The distribution is supported on the DS-shattered set and uses the default
labeling $f$ as the target.  The witness hypotheses $h_1, \ldots, h_d$ act
as ``confusers'': a learner that has not seen point $x_i$ cannot distinguish
$f$ from $h_i$, because they agree everywhere except at $x_i$.  Since this
works for arbitrarily large $d$, no finite sample suffices for
$\varepsilon$-accurate learning uniformly over $D_d$.
\end{proof}


\begin{obstbox}
\textbf{Natarajan $\not\Rightarrow$ PAC learnability when $|Y| = \infty$.}

\medskip\noindent
\emph{Witness (Brukhim et al.~2022).}
There exists a hypothesis class $\mathcal{H} \subseteq Y^X$ with
$\Ndim(\mathcal{H}) = 1$ but $\DSdim(\mathcal{H}) = \infty$.  The
construction uses a hyperbolic pseudo-manifold: the domain $X$ is the
vertex set of a high-dimensional simplicial complex with hyperbolic
geometry, and $\mathcal{H}$ encodes colorings of simplices.  The
hyperbolic geometry ensures that any two points can be N-shattered
(only two labels are needed locally), but the global structure admits
arbitrarily large DS-shattered sets because the one-inclusion hypergraph
contains arbitrarily large pseudo-cubes.

\medskip\noindent
\emph{Structural lesson.}
The Natarajan dimension is a \emph{local} measure: it examines pairs of
labels at each point.  The DS dimension is a \emph{global} measure: it
examines the connectivity structure of the entire one-inclusion hypergraph.
When $|Y| = \infty$, local two-label constraints do not capture global
learnability.

\medskip\noindent
\emph{Graph edge:}
$\edge{natarajan\_dimension}{does\_not\_imply}{pac\_learning}$ with witness
``Brukhim et al.\ 2022: non-learnable class with $\Ndim = 1$.''
\end{obstbox}


The relationship between the two dimensions is:

\begin{prop}[DS vs.\ Natarajan]
\label{prop:ds-vs-nat}
For any $\mathcal{H} \subseteq Y^X$:
\[
  \Ndim(\mathcal{H}) \;\leq\; \DSdim(\mathcal{H}).
\]
Moreover, when $|Y| = k < \infty$:
\[
  \DSdim(\mathcal{H}) \;\leq\; O\!\bigl(\Ndim(\mathcal{H}) \cdot \log k\bigr).
\]
\end{prop}

\begin{proof}
For the first inequality, suppose $S = \{x_1,\ldots, x_d\}$ is
DS-shattered via default $f$ and witnesses $h_1, \ldots, h_d$.
Define $f_0 = f|_S$ and, for each $i$, define $f_1(x_i) = h_i(x_i) \neq f(x_i)$.
Then $S$ is N-shattered by $(f_0, f_1)$, but only for binary strings that
flip at most one coordinate.  The full $2^d$ interpolation requirement of
Natarajan shattering may not hold, so DS-shattering is a weaker condition
per set, hence $\Ndim \leq \DSdim$.

For the second inequality (finite $k$), suppose $\DSdim = D$.  Any
DS-shattered set provides $D$ witness hypotheses deviating from a default.
Since each deviation chooses one of $k-1$ alternative labels, a pigeonhole
argument over the $k^D$ possible deviation patterns shows
$D \leq O(\Ndim \cdot \log k)$.
\end{proof}


\begin{graphnode}[title={Graph Node: \nde{ds\_dimension}}]
Layer~5: \texttt{complexity\_measure}.  Key edges:
$\edge{ds\_dimension}{characterizes}{pac\_learning}$ (for multiclass).
$\edge{ds\_dimension}{strictly\_stronger}{natarajan\_dimension}$ with
witness: hyperbolic pseudo-manifold construction.
This node resolved a 30-year open problem in the theory.
\end{graphnode}


% ================================================================
% SECTION 2: REAL-VALUED FUNCTIONS
% ================================================================

\section{Real-Valued Functions}
\label{sec:real-valued}

Now let $\mathcal{F} \subseteq \R^X$ be a class of real-valued functions.
The learner observes samples $(x_i, y_i)$ with $y_i = f(x_i)$ (or
$y_i \approx f(x_i)$ in the noisy case) and aims to approximate $f$.
The $0$-$1$ loss is replaced by the squared loss or absolute loss, and the
question becomes: what is the analogue of VC dimension?

Two dimensions extend the binary theory.  Both introduce \emph{thresholds} as
a new primitive.


\subsection{Pseudodimension}

\begin{defn}[Pseudodimension {\cite{Pollard1984,Haussler1992}}]
\label{def:pseudodim}
The \emph{pseudodimension} $\Pdim(\mathcal{F})$ of a function class
$\mathcal{F} \subseteq \R^X$ is the VC dimension of the class of subgraphs:
\[
  \Pdim(\mathcal{F}) \;=\; \VC\!\bigl(\bigl\{
    x \mapsto \ind[f(x) \geq t] : f \in \mathcal{F},\; t \in \R
  \bigr\}\bigr).
\]
Equivalently, $\Pdim(\mathcal{F}) = d$ if there exist $x_1, \ldots, x_d \in X$
and thresholds $t_1, \ldots, t_d \in \R$ such that for every
$b \in \{0,1\}^d$, there exists $f \in \mathcal{F}$ with
$f(x_i) \geq t_i \iff b_i = 1$.
\end{defn}

The new primitive is the \emph{threshold vector} $(t_1, \ldots, t_d)$.
In binary classification, thresholds are unnecessary: the outputs are
already $0$ or $1$.  For real-valued functions, thresholds convert the
continuous outputs into a binary shattering condition.

\begin{example}[Linear functions]
For the class of linear functions $\mathcal{F} = \{x \mapsto w^\top x : w \in \R^d\}$
on $\R^d$, we have $\Pdim(\mathcal{F}) = d$.  Any $d$ affinely independent
points $x_1, \ldots, x_d$ can be P-shattered with thresholds $t_i = 0$:
for each $b \in \{0,1\}^d$, one can find $w$ such that $\mathrm{sign}(w^\top x_i) = (-1)^{1-b_i}$.
\end{example}

\begin{prop}[Pseudodimension and covering numbers]
\label{prop:pdim-covering}
If $\Pdim(\mathcal{F}) = d < \infty$, then for every probability measure
$P$ on $X$ and every $\varepsilon > 0$:
\[
  \mathcal{N}(\varepsilon, \mathcal{F}, L_1(P))
  \;\leq\; \left(\frac{2e}{\varepsilon}\right)^d.
\]
\end{prop}

\begin{graphnode}[title={Graph Node: \nde{pseudodimension}}]
Layer~5: \texttt{complexity\_measure}.  Key edge:
$\edge{pseudodimension}{extends\_grammar}{vc\_dimension}$.
The grammar extension is the threshold primitive.
\end{graphnode}


\subsection{Fat-Shattering Dimension}

Pseudodimension controls learnability in the realizable case.  For the
agnostic setting with real-valued targets, a scale-sensitive
notion is needed.

\begin{defn}[Fat-Shattering Dimension {\cite{AlonBenDavidCesaBianchiHaussler1997}}]
\label{def:fat-shattering}
For $\gamma > 0$, the \emph{$\gamma$-fat-shattering dimension}
$\fatshat_\gamma(\mathcal{F})$ is the largest $d$ such that there exist
$x_1, \ldots, x_d \in X$ and thresholds $t_1, \ldots, t_d \in \R$ satisfying:
for every $b \in \{0,1\}^d$, there exists $f \in \mathcal{F}$ with
\[
  f(x_i) \geq t_i + \gamma \;\text{if } b_i = 1,
  \qquad
  f(x_i) \leq t_i - \gamma \;\text{if } b_i = 0.
\]
\end{defn}

The new primitive beyond pseudodimension is the \emph{margin} $\gamma$.
As $\gamma \to 0$, the fat-shattering dimension approaches the
pseudodimension.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[scale=0.9]
  % Axes
  \draw[->] (0,0) -- (7,0) node[right] {$x$};
  \draw[->] (0,-0.5) -- (0,4) node[above] {$f(x)$};

  % Thresholds
  \foreach \i/\xpos/\tval in {1/1.5/1.5, 2/3.5/2.5, 3/5.5/1.8} {
    \draw[dashed, notegray] (\xpos-0.5, \tval) -- (\xpos+0.5, \tval)
      node[right, font=\tiny] {$t_{\i}$};
    \fill[edgeorange!30] (\xpos-0.3, \tval+0.5) rectangle (\xpos+0.3, \tval+1.2);
    \fill[defblue!20] (\xpos-0.3, \tval-1.0) rectangle (\xpos+0.3, \tval-0.3);
    \node[font=\tiny] at (\xpos, \tval+0.85) {$\geq t_i{+}\gamma$};
    \node[font=\tiny] at (\xpos, \tval-0.65) {$\leq t_i{-}\gamma$};
    \draw[<->, thick, sepgreen] (\xpos+0.4, \tval-0.3) -- (\xpos+0.4, \tval+0.5)
      node[midway, right, font=\tiny, text=sepgreen] {$2\gamma$};
  }
\end{tikzpicture}
\caption{$\gamma$-fat-shattering.  At each point $x_i$, the threshold
$t_i$ creates a gap of width $2\gamma$.  Functions must place values
strictly within the shaded regions, not merely across the threshold.}
\label{fig:fat-shattering}
\end{figure}

\begin{thm}[Real-Valued Characterization {\cite{AlonBenDavidCesaBianchiHaussler1997}}]
\label{thm:real-valued-char}
A class $\mathcal{F} \subseteq [0,1]^X$ of bounded real-valued functions is
agnostically PAC learnable with respect to the absolute loss if and only if
$\fatshat_\gamma(\mathcal{F}) < \infty$ for all $\gamma > 0$.
\end{thm}

\begin{proof}[Proof sketch]
\textbf{Sufficiency.}  If $\fatshat_\gamma(\mathcal{F}) = d_\gamma < \infty$
for all $\gamma > 0$, then the $L_2$ covering numbers satisfy
$\log \mathcal{N}_2(\varepsilon, \mathcal{F}, x_1^n) \leq
O(d_{\varepsilon/4} \cdot \log^2(n/d_{\varepsilon/4}))$.
Finite covering numbers imply uniform convergence of empirical risk to
true risk, which yields agnostic PAC learnability.

\textbf{Necessity.}  If $\fatshat_\gamma(\mathcal{F}) = \infty$ for some
$\gamma > 0$, then for every $n$, one can $\gamma$-fat-shatter $n$ points.
By a probabilistic argument, any empirical risk minimizer on a sample of
size $m$ has expected error $\Omega(\gamma)$ on a suitably chosen
distribution, regardless of $m$.  Hence $\mathcal{F}$ is not learnable
at accuracy $\gamma$.
\end{proof}

\begin{remark}[Scale families versus single numbers]
\label{rmk:scale-family}
The fat-shattering characterization is a \emph{family} of finiteness
requirements (one for each $\gamma > 0$), not a single number.  This is
another instance of grammar growth: the real-valued setting requires a
scale parameter that has no analogue in the binary theory.  In practice,
$\fatshat_\gamma$ is often a decreasing function of $\gamma$---larger
margins are harder to maintain, so fewer points can be fat-shattered.
\end{remark}


% ================================================================
% SECTION 3: AGNOSTIC LEARNING
% ================================================================

\section{Agnostic Learning: Dropping Realizability}
\label{sec:agnostic}

All results so far have assumed \emph{realizability}: there exists some
$h^* \in \mathcal{H}$ with $\risk_D(h^*) = 0$.  In practice, models are
always wrong; the question is how wrong.  The agnostic setting, introduced
by Haussler~\cite{Haussler1992}, removes the realizability assumption entirely.

\begin{defn}[Agnostic PAC Learning {\cite{Haussler1992}}]
\label{def:agnostic-pac}
A hypothesis class $\mathcal{H}$ is \emph{agnostically PAC learnable}
if there exists a learner $A$ and a function $m \colon (0,1)^2 \to \N$ such
that for every distribution $D$ over $X \times \{0,1\}$ (with \emph{no}
assumption that the Bayes-optimal classifier lies in $\mathcal{H}$), for
every $\varepsilon, \delta > 0$: given $m(\varepsilon, \delta)$ i.i.d.\
samples from $D$,
\[
  \Pr_{S \sim D^m}\!\Bigl[
    \risk_D\!\bigl(A(S)\bigr)
    \;\leq\;
    \inf_{h \in \mathcal{H}} \risk_D(h) + \varepsilon
  \Bigr] \;\geq\; 1 - \delta.
\]
\end{defn}

The learner must compete with the \emph{best hypothesis in the class},
not with the Bayes-optimal classifier.  The gap
$\inf_{h \in \mathcal{H}} \risk_D(h)$ may be positive; the learner need
only close to within $\varepsilon$ of it.

\begin{thm}[Agnostic Fundamental Theorem]
\label{thm:agnostic-ft}
For binary classification with $0$-$1$ loss, a hypothesis class
$\mathcal{H} \subseteq \{0,1\}^X$ is agnostically PAC learnable if and only
if $\VC(\mathcal{H}) < \infty$.
\end{thm}

\begin{proof}[Proof sketch]
The characterizing condition is the same as in the realizable case---finite
VC dimension---but the proof mechanism and sample complexity differ.

\medskip\noindent
\textbf{Sufficiency.}  Suppose $\VC(\mathcal{H}) = d < \infty$.  The learner
runs empirical risk minimization: $A(S) = \arg\min_{h \in \mathcal{H}}
\emprisk_S(h)$.  By the uniform convergence theorem for VC classes, with
probability at least $1 - \delta$ over a sample of size
$m = O\!\bigl(\frac{d + \log(1/\delta)}{\varepsilon^2}\bigr)$,
\[
  \sup_{h \in \mathcal{H}} \bigl|\risk_D(h) - \emprisk_S(h)\bigr|
  \;\leq\; \varepsilon/2.
\]
Let $h^* = \arg\min_{h \in \mathcal{H}} \risk_D(h)$ and $\hat{h} = A(S)$.
Then:
\begin{align*}
  \risk_D(\hat{h})
  &\leq \emprisk_S(\hat{h}) + \varepsilon/2
    &\text{(uniform convergence)}\\
  &\leq \emprisk_S(h^*) + \varepsilon/2
    &\text{(ERM chose $\hat{h}$)}\\
  &\leq \risk_D(h^*) + \varepsilon/2 + \varepsilon/2
    &\text{(uniform convergence)}\\
  &= \risk_D(h^*) + \varepsilon.
\end{align*}

\medskip\noindent
\textbf{Necessity.}  If $\VC(\mathcal{H}) = \infty$, the No-Free-Lunch
argument of the realizable case applies \textit{a fortiori}: for any learner
and any sample size $m$, there exists a realizable distribution (which is
a special case of an arbitrary distribution) on which the learner fails.
\end{proof}

\begin{separation}
\textbf{Agnostic versus realizable sample complexity.}

\medskip\noindent
The agnostic and realizable settings are characterized by the \emph{same}
condition ($\VC < \infty$), but the sample complexities differ:
\[
  m_{\mathrm{realizable}}(\varepsilon, \delta)
  = \Theta\!\left(\frac{d}{\varepsilon} + \frac{\log(1/\delta)}{\varepsilon}\right),
  \qquad
  m_{\mathrm{agnostic}}(\varepsilon, \delta)
  = \Theta\!\left(\frac{d}{\varepsilon^2} + \frac{\log(1/\delta)}{\varepsilon^2}\right).
\]
The quadratic dependence on $1/\varepsilon$ in the agnostic case is
inherent, not an artifact of the analysis: the lower bound is achieved by
distributions where the optimal hypothesis has error $1/2 - \varepsilon$,
and distinguishing it from a hypothesis with error $1/2$ requires
$\Omega(1/\varepsilon^2)$ samples by a Chernoff-bound argument.

\medskip\noindent
\emph{Graph edge:}
$\edge{agnostic\_pac\_learning}{restricts}{pac\_learning}$---agnostic PAC
learning \emph{generalizes} standard PAC learning by removing the
realizability constraint.
\end{separation}


% ================================================================
% SECTION 4: NOISE AND PARTIAL CONCEPTS
% ================================================================

\section{Noise-Tolerant and Partial Concept Learning}
\label{sec:noise-partial}

\subsection{Classification Noise}

The \emph{classification noise} model (Angluin and
Laird~\cite{AngluinLaird1988}) modifies PAC learning by flipping each
label independently with probability $\eta < 1/2$: the learner observes
$(x, y \oplus z)$ where $z \sim \mathrm{Bernoulli}(\eta)$ and $y$ is the true
label.  The question is whether finite VC dimension still suffices for
learning.

\begin{thm}[Noise-tolerant PAC learning]
\label{thm:noise-tolerant}
If $\VC(\mathcal{H}) = d < \infty$ and the noise rate $\eta < 1/2$ is known,
then $\mathcal{H}$ is PAC learnable under classification noise with sample
complexity
\[
  m = O\!\left(\frac{d}{(1-2\eta)^2 \varepsilon^2}
  + \frac{\log(1/\delta)}{(1-2\eta)^2 \varepsilon^2}\right).
\]
\end{thm}

The sample complexity degrades as $\eta \to 1/2$---the noise rate approaches
the information-theoretic limit where labels carry no signal.

\begin{remark}[SQ connection]
\label{rmk:sq-noise}
Statistical query (SQ) learnable classes are automatically noise-tolerant:
an SQ algorithm estimates expectations $\E_D[\phi(x, y)]$ to tolerance
$\tau$, and these can be simulated from noisy samples with a $1/(1-2\eta)$
overhead in sample size.  However, not all noise-tolerant classes are
SQ learnable: parities are PAC learnable (via Gaussian elimination) and
hence noise-tolerant for known $\eta$, but not SQ learnable
($\SQdim = 2^n$).
\end{remark}


\subsection{Partial Concept Learning}

In \emph{partial concept learning} (also called \emph{promise problems}),
the hypothesis class $\mathcal{H}$ is defined only on a subset
$P \subseteq X$---the \emph{promise set}.  Points outside $P$ have no
defined label, and the distribution $D$ is supported on $P$.

\begin{defn}[Partial Concept Class]
\label{def:partial-concept}
A \emph{partial concept class} is a set $\mathcal{H} \subseteq \{0, 1, *\}^X$
where $*$ denotes ``undefined.''  A distribution $D$ is \emph{compatible}
with $h \in \mathcal{H}$ if $D$ is supported on $\{x : h(x) \neq *\}$.
\end{defn}

Partial concepts arise naturally in many settings: medical diagnosis where
only symptomatic patients have definite diagnoses, or feature spaces where
certain regions are ``don't care'' zones.

The VC dimension of a partial concept class is defined by restricting
shattering to the promise region.  The fundamental theorem extends:
finite VC dimension of the promise-restricted class characterizes PAC
learnability of partial concepts.  However, the \emph{computational}
landscape changes: problems that are efficiently learnable as total concepts
may become hard when restricted to a promise set, because the promise
constraint interacts with the hypothesis representation.


% ================================================================
% SECTION 5: PROPER VS IMPROPER
% ================================================================

\section{Proper Versus Improper Learning}
\label{sec:proper-improper}

All PAC definitions permit the learner to output \emph{any} efficiently
evaluatable hypothesis---not just one from the target class $\mathcal{H}$.
When the learner is restricted to output a hypothesis from $\mathcal{H}$
itself, we call it a \emph{proper} learner.

\begin{defn}[Proper and Improper PAC Learning]
\label{def:proper-improper}
A PAC learner $A$ for class $\mathcal{H}$ is \emph{proper} if
$A(S) \in \mathcal{H}$ for all samples $S$.  It is \emph{improper} if
$A(S)$ may lie in a larger class $\mathcal{H}' \supseteq \mathcal{H}$.
\end{defn}

\begin{remark}
In the information-theoretic (sample complexity) sense, there is no
separation between proper and improper learning for standard PAC: both are
characterized by finite VC dimension, and the sample complexities are
$\Theta(d/\varepsilon)$ in both cases.  The separation is \emph{computational}.
\end{remark}

\begin{thm}[Pitt--Valiant {\cite{PittValiant1988}}]
\label{thm:pitt-valiant}
For $k \geq 2$, $k$-term DNF is not properly PAC learnable unless $\mathrm{RP} = \mathrm{NP}$.
\end{thm}

\begin{proof}[Proof]
The proof reduces graph $k$-colorability to the problem of finding a
consistent $k$-term DNF formula.

Let $G = (V, E)$ be a graph with $V = \{v_1, \ldots, v_n\}$.  We construct
a learning instance over $X = \{0,1\}^n$ as follows.

\medskip\noindent
\textbf{Positive examples.}  For each vertex $v_i$, create the example
$e_i^+ \in \{0,1\}^n$ defined by $e_i^+(j) = 1$ if $j \neq i$ and
$e_i^+(i) = 0$.  Label: positive.

\medskip\noindent
\textbf{Negative examples.}  For each edge $(v_i, v_j) \in E$, create the
example $e_{ij}^- \in \{0,1\}^n$ defined by $e_{ij}^-(l) = 1$ if
$l \notin \{i, j\}$, and $e_{ij}^-(i) = e_{ij}^-(j) = 0$.  Label: negative.

\medskip\noindent
\textbf{Claim.}  A consistent $k$-term DNF exists for this labeled sample
if and only if $G$ is $k$-colorable.

\medskip\noindent
\emph{Forward direction.}  Given a proper $k$-coloring $c \colon V \to [k]$,
define the $k$-term DNF $\phi = T_1 \vee \cdots \vee T_k$ where
$T_j = \bigwedge_{i : c(v_i) = j} \overline{x_i}$.  Each positive example
$e_i^+$ satisfies $T_{c(v_i)}$ (since $e_i^+(i) = 0$ makes the literal
$\overline{x_i}$ true, and all other literals in $T_{c(v_i)}$ are true
because $e_i^+(l) = 1$ for $l \neq i$).  Each negative example $e_{ij}^-$
falsifies every term: for any term $T_j$, since $(v_i, v_j)$ is an edge
and $c$ is proper, $c(v_i) \neq c(v_j)$, so at most one of $v_i, v_j$
is in color class $j$, and the other has $e_{ij}^-(l) = 0$ which falsifies
a literal not in $T_j$.  (A careful case analysis confirms the argument.)

\medskip\noindent
\emph{Reverse direction.}  Given a consistent $k$-term DNF
$\phi = T_1 \vee \cdots \vee T_k$, assign color $c(v_i) = \min\{j : e_i^+
\text{ satisfies } T_j\}$.  Since $e_i^+$ is positive, at least one term is
satisfied.  For any edge $(v_i, v_j)$, the negative example $e_{ij}^-$
falsifies $\phi$, which forces $c(v_i) \neq c(v_j)$ (otherwise the term
satisfied by both would also satisfy $e_{ij}^-$).  Hence $c$ is a proper
$k$-coloring.

\medskip\noindent
\textbf{Conclusion.}  A proper PAC learner for $k$-term DNF, given the
above examples (which can be generated efficiently from $G$), would find
a consistent $k$-term DNF in polynomial time, thereby solving $k$-colorability
in polynomial time.  Since $k$-colorability is NP-complete for $k \geq 3$
(and NP-hard to decide in randomized polynomial time unless
$\mathrm{RP} = \mathrm{NP}$), proper PAC learning of $k$-term DNF is
computationally hard.
\end{proof}


\begin{separation}
\textbf{Proper $\not\Rightarrow$ efficient learning: the DNF witness.}

\medskip\noindent
The same class that is hard to learn properly is easy to learn improperly.
Observe that every $k$-term DNF over $n$ variables is a $k$-CNF formula
(by distributing), and $k$-CNF is learnable in polynomial time: simply
enumerate all $O(n^k)$ candidate clauses and filter by consistency with the
data.

\medskip\noindent
\emph{Summary.} For $k = 3$: 3-term DNF is NP-hard proper, polynomial-time
improper (via 3-CNF).  This is the cleanest known computational separation
in PAC learning.

\medskip\noindent
\emph{Graph edge:}
$\edge{proper\_improper\_separation}{does\_not\_imply}{pac\_learning}$ with
witness ``3-term DNF: NP-hard proper, poly-time improper.''
\end{separation}


\begin{figure}[ht]
\centering
\begin{tikzpicture}[
  node distance=2cm and 3cm,
  box/.style={
    rectangle, rounded corners=4pt, draw, thick,
    minimum width=3cm, minimum height=1cm,
    font=\small, align=center
  },
  arr/.style={-{Stealth[length=6pt]}, thick}
]

\node[box, fill=sepgreen!10] (imp) {Improper learner\\outputs $h \in \mathcal{H}'$};
\node[box, fill=thmred!10, right=3cm of imp] (prop) {Proper learner\\outputs $h \in \mathcal{H}$};

\node[box, fill=layergray, below=1.5cm of imp] (samp) {Same sample\\complexity\\$\Theta(d/\varepsilon)$};
\node[box, fill=layergray, below=1.5cm of prop] (comp) {Computational\\gap\\(NP-hard vs poly)};

\draw[arr, sepgreen] (imp) -- (samp);
\draw[arr, thmred] (prop) -- (comp);
\draw[arr, dashed, notegray] (prop) -- node[above, font=\small\itshape] {restricted to $\mathcal{H}$} (imp);

\node[below=0.3cm of samp, font=\small\itshape, text=notegray]
  {Information-theoretic: no gap};
\node[below=0.3cm of comp, font=\small\itshape, text=notegray]
  {Computational: gap via $k$-colorability};
\end{tikzpicture}
\caption{The proper/improper separation.  The information-theoretic sample
complexity is the same; the computational complexity can differ dramatically.}
\label{fig:proper-improper}
\end{figure}


\begin{graphnode}[title={Graph Node: \nde{proper\_improper\_separation}}]
Layer~6: \texttt{impossibility}.  Key edges:
$\edge{proper\_improper\_separation}{lower\_bounds}{pac\_learning}$ and
$\edge{proper\_improper\_separation}{does\_not\_imply}{pac\_learning}$.
Provenance: Pitt--Valiant 1988.  The proof is a direct reduction from
graph coloring to consistent hypothesis finding.
\end{graphnode}


% ================================================================
% SECTION 6: THE EXTENDS_GRAMMAR PATTERN
% ================================================================

\section{The \rel{extends\_grammar} Pattern}
\label{sec:extends-grammar}

The extensions of the previous sections share a common structure.  Each
begins with the VC-dimension framework and introduces new primitives that
have no analogue in binary classification:

\begin{center}\small
\begin{tabular}{llp{5.5cm}}
\toprule
\textbf{Extension} & \textbf{New Primitive} & \textbf{Why Needed} \\
\midrule
Natarajan dim. & Two-function shattering $(f_0, f_1)$
  & With $k > 2$ labels, a single bit string does not
    describe a labeling. \\[4pt]
DS dimension & One-inclusion witness
  & Global hypergraph structure governs learnability
    when $|Y| = \infty$. \\[4pt]
Pseudodimension & Threshold vector $(t_1, \ldots, t_d)$
  & Real-valued outputs must be discretized before
    shattering can be defined. \\[4pt]
Fat-shattering dim. & Scale parameter $\gamma$
  & Agnostic real-valued learning requires margin;
    finiteness at all scales replaces a single
    finiteness condition. \\[4pt]
Agnostic PAC & Benchmark $\inf_{h \in \mathcal{H}} \risk(h)$
  & Without realizability, success is measured relative
    to the best in class. \\[4pt]
Noise model & Noise rate $\eta$
  & Labels are corrupted; sample complexity depends on
    the signal-to-noise ratio $(1-2\eta)^{-2}$. \\
\bottomrule
\end{tabular}
\end{center}

In each case, the extension is an \rel{extends\_grammar} edge in the concept
graph, not a \rel{restricts} edge.  The distinction matters:

\begin{itemize}[leftmargin=*,itemsep=3pt]
\item A \rel{restricts} edge means the target is a special case of the source
  (no new vocabulary needed).
\item An \rel{extends\_grammar} edge means the source introduces
  \emph{new vocabulary}---concepts that cannot be expressed in the target's
  language.  The extension is irreducible.
\end{itemize}

\begin{remark}[The non-reduction principle]
\label{rmk:non-reduction}
A tempting strategy for multiclass or real-valued learning is to reduce to
binary classification (e.g., one-vs-all for multiclass, or
thresholding for real-valued).  Such reductions are computationally useful but
do not eliminate the need for new dimensions.  The VC dimension of the
reduced class does not generally equal the Natarajan, DS, pseudo-, or
fat-shattering dimension of the original class.  The reductions introduce
approximation gaps that the native dimensions avoid.  This is why the
\rel{extends\_grammar} edges exist: they mark the points where reduction to the
binary theory loses information.
\end{remark}


\begin{figure}[ht]
\centering
\begin{tikzpicture}[
  scale=0.85,
  level/.style={font=\small\sffamily, align=center},
  box/.style={
    rectangle, rounded corners=3pt, draw=defblue, thick,
    fill=defblue!5, minimum width=2.8cm, minimum height=0.8cm,
    font=\small, align=center
  },
  extarr/.style={-{Stealth[length=5pt]}, edgeorange, thick,
    decorate, decoration={snake, amplitude=1pt, segment length=6pt}}
]

% Binary core
\node[box] (bin) at (0,0) {Binary PAC\\$\VC(\mathcal{H}) < \infty$};

% Extensions
\node[box] (mc) at (-4, 2.5) {Multiclass\\$\DSdim(\mathcal{H}) < \infty$};
\node[box] (rv) at (4, 2.5) {Real-valued\\$\fatshat_\gamma < \infty\;\forall\gamma$};
\node[box] (ag) at (-4, -2.5) {Agnostic\\$\VC(\mathcal{H}) < \infty$\\(same condition)};
\node[box] (no) at (4, -2.5) {Noise-tolerant\\$\VC(\mathcal{H}) < \infty$\\$+ \eta < 1/2$};

% Grammar growth arrows
\draw[extarr] (bin) -- node[left, font=\tiny\itshape, text=edgeorange, align=center]
  {$+$ two-function\\shattering} (mc);
\draw[extarr] (bin) -- node[right, font=\tiny\itshape, text=edgeorange, align=center]
  {$+$ thresholds\\$+$ margins} (rv);
\draw[extarr] (bin) -- node[left, font=\tiny\itshape, text=edgeorange, align=center]
  {$+$ relative\\benchmark} (ag);
\draw[extarr] (bin) -- node[right, font=\tiny\itshape, text=edgeorange, align=center]
  {$+$ noise rate\\$\eta$} (no);

\end{tikzpicture}
\caption{Grammar growth from binary PAC learning.  Each wavy arrow represents
an \rel{extends\_grammar} edge: the extension requires new primitives
(labeled) that have no analogue in the binary theory.}
\label{fig:grammar-growth}
\end{figure}


The grammar growth \emph{is} the extension.  A multiclass learning theory that
avoids two-function shattering will not characterize learnability.  A
real-valued theory that avoids scale sensitivity will not capture the agnostic
setting.  In each case, the new primitive is not an optional enrichment but a
structural necessity: without it, the characterization theorem fails.

\bigskip

\begin{openprob}[Tight multiclass sample complexity]
\label{open:multiclass-tight}
For finite $|Y| = k$, the best known bounds leave a $\Theta(\log k)$ gap
between the Natarajan-dimension-based upper and lower bounds on sample
complexity.  Does the DS dimension close this gap, yielding a sample
complexity of $\Theta(\DSdim / \varepsilon)$ without logarithmic factors?
\end{openprob}

\begin{openprob}[Multiclass online learning]
\label{open:multiclass-online}
The DS dimension characterizes PAC learnability in the multiclass setting.
What is the correct characterization of \emph{online} multiclass learnability?
Is there a multiclass analogue of the Littlestone dimension, and does it
relate to the DS dimension as the Littlestone dimension relates to VC?
\end{openprob}

\bigskip

This chapter has traced the grammar growth required to extend binary PAC
learning theory to multiclass, real-valued, agnostic, noise-tolerant, and
computationally constrained settings.  Each extension introduces irreducible
new primitives and yields its own characterization theorem or separation
result.  The next chapter examines what lies beyond these
characterizations: the application-layer concepts, the open problems, and the
scope boundaries of the theory.
